### 2017-02-15 Lab_1/acro.tex
Filled in first three questions of lab 1.

%!TEX root =labo.tex

\begin{acronym}
\acro{aes}[AES]{Advanced Encryption Standard}
\acro{ap}[AP]{access point}
\acro{arp}[ARP]{Address Resolution Protocol}
\acro{bipt}[BIPT]{Belgisch Instituut voor Postdiensten en Telecommunicatie - Belgian Institute for Postal services and Telecommunications}
\acro{bss}[BSS]{basic service set}
\acro{bssid}[BSSID]{\acl{bss} ID}
\acro{ccmp}[CCMP]{Counter Cipher Mode with Block Chaining Message Authentication Code Protocol}
\acro{cts}[CTS]{Clear to Send}
\acro{dhcp}[DHCP]{Dynamic Host Configuration Protocol}
\acro{dos}[DoS]{denial of service}
\acro{essid}[ESSID]{extended \acl{ssid}}
\acro{gid}[grID]{group ID}
\acro{ibss}[IBSS]{independent \acl{bss}}
\acro{icmp}[ICMP]{Internet Control Message Protocol}
\acro{iv}[IV]{initialization vector}
\acro{mimo}[MIMO]{multiple-input, multiple-output}
\acro{mac}[MAC]{Media Access Control}
\acro{mtu}[MTU]{Maximum Transferable Unit}
\acro{nat}[NAT]{Network Address Translation}
\acro{nd}[ND]{Neighbor Discovery}
\acro{nic}{network interface card}
\acro{ptw}[PTW]{Pychkine, Tews, Weinmann}
\acro{pxe}[PXE]{preboot execution environment}
\acro{ram}[RAM]{Random Access Memory}
\acro{rts}[RTS]{Request to Send}
\acro{rtt}[RTT]{round trip time}
\acro{ssh}[SSH]{secure shell}
\acro{ssid}[SSID]{Service Set ID}
\acro{sta}[STA]{station}
\acro{tcp}[TCP]{Transmission Control Protocol}
\acro{udp}[UDP]{User Datagram Protocol}
\acro{vap}[VAP]{virtual access point}
\acro{wan}[WAN]{wide area network}
\acro{wep}[WEP]{Wired Equivalent Privacy}
\acro{wmn}{wireless mesh node}
\acro{wnic}{wireless \acl{nic}}
\acro{wpa}[WPA]{Wi-Fi Protected Access}
\acro{wps}[WPS]{Wi-Fi Protected Setup}
\end{acronym}

### 2017-02-15 Lab_1/labo1.tex
Filled in first three questions of lab 1.

%!TEX root = labo.tex
\setcounter{chapter}{0}
\chapter{Basic Configurations}

The goal of the first lab is to make you acquainted with the hardware and software needed to perform the tasks in other modules. Furthermore we will have a look at the various sniffing possibilities in wireless networks.

\section{Device Exploration}

Have a closer look a the devices used for this course.

\begin{exercise}{Getting to know the interfaces}
	\begin{enumerate}
		\item Log in on \ac{wmn}1. \newline
		\command{ssh root@wmn1}
		\item When asked to give a name, enter
		\command{AP}
		\item Get a list of all available interfaces. \newline
		\command{\prompt{\acs{ap}} ifconfig -a}
		\item List the interfaces and their \ac{mac} addresses.\newline		\begin{esolution}
		\end{esolution}
		\item A \ac{mac} address is unique per card, but also carries a generic part, identifying the vendor of a card. List the prefixes and find out the vendors (name) of all different interfaces found in the node.\\
		\remark Google is your friend!\newline		
		\begin{esolution}
		\end{esolution}
	\end{enumerate}

\end{exercise}

\section{Access Point Setup}

\begin{figure}[h]
	\begin{center}
		\includegraphics[width=0.8\textwidth]{images/snif1.pdf} 
		\caption{The basic infrastructure setup}
		\label{fig:basic-infra}  
	\end{center}
\end{figure}

The goal is to create a basic infrastructure setup as can be found at home. The setup consists of one \ac{ap} and two connected stations as shown in figure \ref{fig:basic-infra}. In this case, two laptops are connected to the same \ac{ap} and can communicate to each other. We will not consider any communication to the outside world.

\subsection{\ac{ap} Configuration}
To set up this basic infrastructure network, start by configuring the \ac{ap}.

\begin{exercise}{Set up an \ac{ap}}
\begin{enumerate}
	\item Check the wireless parameters of wlan1:\newline
	\command{\prompt{\acs{ap}} iwconfig wlan1}\newline
	This should give you some output like
\begin{verbatim}
wlan1     IEEE 802.11abgn  ESSID:off/any  
          Mode:Managed  Access Point: Not-Associated   Tx-Power=0 dBm   
          RTS thr:off   Fragment thr:off
          Encryption key:off
          Power Management:off
\end{verbatim}

\remark Remark that we did not configure the interface yet!

Give an explanation for all available parameters which are displayed using \incommand{iwconfig} (hint: man pages).\newline
\begin{esolution}
\end{esolution}

\item Configure the wlan1 interface using your assigned channel and the \ac{essid}:\newline
\begin{lstlisting}
interface=wlan
driver=nl80211
logger_syslog=-1
logger_syslog_level=2
logger_stdout=-1
logger_stdout_level=2
debug=4
hw_mode=a
channel=x
macaddr_acl=0
auth_algs=3
eapol_key_index_workaround=0
eap_server=0
wpa=0
ssid=wmn-gid-A
\end{lstlisting}

This file can be found on the devices as \texttt{hostapd.conf}. Edit it to change the channel number, interface name and \ac{essid}, and then perform the following:\newline
\command{\prompt{\acs{ap}} hostapd -B hostapd.conf}

\incommand{\prompt{\acs{ap}} iwconfig} should now return something like: 
\begin{verbatim}
wlan1     IEEE 802.11abgn  Mode:Master  Tx-Power=20 dBm   
          RTS thr:off   Fragment thr:off
          Power Management:off
\end{verbatim}

\end{enumerate}
\end{exercise}

\subsection{Station Configuration}

\begin{exercise}{Configuring the wireless stations}

We will now configure \emph{both} stations  and verify they get associated with the \ac{ap} you just created. Make sure to repeat the commands to configure the second station.

\begin{enumerate}
	\item Bring up the interface: \newline
	\command{\prompt{\acs{sta}1} ifconfig wlan1 up}
	\item Configure the interface to connect to our \ac{ap}: \newline
	\command{\prompt{\acs{sta}1} iw dev wlan1 connect \ac{wmn}-\acs{gid}-A}
	\item Do the same on \ac{sta}2.

\remark When you run \incommand{iwconfig}, the \incommand{Access Point} parameter should show the \ac{mac} address of the \ac{ap} and show identical frequencies and \acp{essid} on both stations. 
\begin{verbatim}
wlan1     IEEE 802.11abgn  ESSID:"wmn-0-A"  
          Mode:Managed  Frequency:5.24 GHz  Access Point: 00:00:11:22:33:44   
          Bit Rate=1 Mb/s   Tx-Power=5 dBm   
          RTS thr:off   Fragment thr:off
          Encryption key:off
          Power Management:off
          Link Quality=12/70  Signal level=-164 dBm  
          Rx invalid nwid:0  Rx invalid crypt:0  Rx invalid frag:0
          Tx excessive retries:0  Invalid misc:4   Missed beacon:0
\end{verbatim}
	\item What parameters are different in your output compared to the output above?\newline
	\begin{esolution}
	\end{esolution}
\end{enumerate}
\end{exercise}

If the output for both stations confirms the association between the two wireless stations and the \ac{ap}, the actual L2 connection between these three devices is up and running. To verify the connection, we will configure IP addresses on the stations and have them communicate.

\begin{exercise}{Verifying the basic setup}\label{ex:pingtest}

	\begin{enumerate}
		\item Configure the IP address of \ac{sta}1 and \ac{sta}2:\newline
		\command{\prompt{\ac{sta}1} ip addr add fc00::\acs{gid}::1/64 dev wlan1}
		\command{\prompt{\ac{sta}2} ip addr add fc00:\acs{gid}::2/64 dev wlan1}
		\item Start a ping session from \ac{sta}1 to \ac{sta}2. Give the minimum, maximum and average \ac{rtt}:\newline
		\command{\prompt{\ac{sta}1} ping6 -c 30 fc00:\acs{gid}::2}
		\begin{esolution}
		\end{esolution}
		\item Perform a ping between the same devices as before, but this time, use the wired IP addresses. Again write down the same values.\newline
		\begin{esolution}
		\end{esolution}
	\end{enumerate}
\end{exercise}

\remark Remark that we did not (yet) enable L3 capabilities at the \ac{ap}.

\section{Wireless Sniffing}

Using the basic network we constructed in the previous section, we will go through the various sniffing modes available in wireless networks. The main difference between wired and wireless networks is obviously the fact that the transmission medium (radio waves vs. cable) is shared amongst all wireless users. Therefore, sniffing the network offers a lot more opportunities in wireless systems.
\begin{figure}
	\begin{center}
		\includegraphics[width=0.5\textwidth]{images/snif1.pdf} 
		\caption{Wireless packet capture setup} 
		\label{fig:snif1} 
	\end{center}
\end{figure}

The first setup is shown in figure \ref{fig:snif1} and corresponds to the basic setup we created in the previous section.

\begin{exercise}{First scan}\label{ex:firstScan}
\begin{enumerate}
	\item Make sure the \ac{nd} cache of both \acp{sta} is cleared.\label{ex:1-1}\newline
	\command{\prompt{\ac{sta}2} ip neigh flush dev wlan1}
	\command{\prompt{\ac{sta}1} ip neigh flush dev wlan1}
	\remark You can always check the state of the cache using \command{ip neigh show}
	\item On the \ac{ap} and on both \acp{sta}, start a packet capture using \incommand{tcpdump} and save it to \file{snif-location.pcap}. \newline
	\remark This is the time to make sure that you have a remote location mounted with \verb!sshfs! in /mnt!\newline\newline
	\command{\prompt{\acs{ap}} tcpdump -i wlan1 -w \file{\acs{ap}.pcap}}
	\command{\prompt{\acs{sta}1} tcpdump -i wlan1 -w \file{\acs{sta}1.pcap}}
	\command{\prompt{\acs{sta}2} tcpdump -i wlan1 -w \file{\acs{sta}2.pcap}}
	\item Start a ping session from \ac{sta}1 to \ac{sta}2. Limit the ping to only two requests. \label{ex:1-3}\newline
	\command{\prompt{\ac{sta}1} ping6 -c 2 fc00:\acs{gid}::2} 
	\item Which type of packets can be seen in the \acs{ap} trace file? You can open the trace file with wireshark on the remote computer.\newline
	\begin{esolution}
	\end{esolution}
	\item Now, take a closer look at the \ac{mac} headers. Which type of link layer headers show up on the packets?\newline
	\begin{esolution}
	\end{esolution}
\end{enumerate}
\end{exercise}

From the tracefile made on the AP, it is impossible to see if we made a wired or wireless trace. Let's more closely examine the tracefiles made on the stations, where a difference will become apparent.

\begin{exercise}{Scanning other interfaces}\label{ex:scan}

Open the trace files made on both the sending and receiving station.

	\begin{enumerate}
		\item You should observe duplicate entries. Describe which duplicates are observed in each file (use packet numbers!):\newline
		\begin{esolution}
		\end{esolution}
		\item Take a closer look at the \ac{mac} addresses of the duplicate packets using Wireshark. What addresses are used on the frames? Are they identical?\newline
		\begin{esolution}
		\end{esolution}
	\end{enumerate}
\end{exercise}

\incommand{tcpdump} will automatically put the interface in \emph{promiscuous mode}. This means all packets which can be read by the wireless interface will be delivered up the networking stack even though the destination \ac{mac} address does not correspond to the address of the specific card. Each interface receives all packets sent by the \ac{ap}. If the destination \ac{mac} address is not that of the receiving interface, the packet is normally dropped. In promiscuous mode, however, these packets are nevertheless stored in the trace file. Now repeat the previous exercise but disable the promiscuous mode.

\begin{exercise}{Disabling promiscuous mode}

\begin{enumerate}
	\item Make sure the \ac{nd} cache is cleared on \ac{sta}1 and \ac{sta}2. \newline 
	\command{\prompt{\acs{sta}1} ip neigh flush dev wlan1}
	\command{\prompt{\acs{sta}2} ip neigh flush dev wlan1} 
	\item On \ac{sta}1, start a packet capture not using promiscuous mode using \incommand{tcpdump} and save it to \file{\acs{sta}1.pcap}.\newline
	\command{\prompt{\acs{sta}1} tcpdump -i wlan1 -p -w \file{\acs{sta}1.pcap}} 
	\item Start a ping session from  \ac{sta}2 to \ac{sta}1. \newline
	\command{\prompt{\acs{sta}2} ping6 -c 2 fc00:\acs{gid}::1} 
	\item Do you still observe the duplicate entries?\newline
	\begin{esolution}
	\end{esolution} 
\end{enumerate}
\end{exercise}

\begin{figure}[h]
	\begin{center}
		\subfloat[]{\includegraphics[width=0.5\textwidth]{images/snif2b.pdf} 	\label{fig:snif2b} }
		\subfloat[]{\includegraphics[width=0.5\textwidth]{images/snif2c.pdf} 	\label{fig:snif2c} }
		\caption{Third party scanning}

	\end{center}
\end{figure}

\begin{exercise}{Third party scanning}
	\label{ex:secondAP}

So far, we only sniffed the network on nodes which took part of the network activity. Let's introduce another party, which just wants to listen to what is happening on the channel. Imagine a notebook eavesdropping on the traffic of a wireless cell. First we will add it to the same \ac{essid} and later on change it to a different \ac{essid}.  The intended setups are shown in figure \ref{fig:snif2b} and \ref{fig:snif2c}. The first situation then compares to a situation where a laptop is connected to your home network and is sniffing the traffic of other active stations in this network, while the latter situation can be seen as you being connected to your network, trying to sniff the network of a neighbour on the same channel, but using a different network name.

	We'll start from the previous setup.\newline
	\remark It is essential that you bring \incommand{wlan1} down on \ac{sta}1 first.

	\begin{enumerate}
		\item Bring \incommand{wlan1} down on \ac{sta}1.\newline
		\command{\prompt{\ac{sta}1} ifconfig wlan1 down}
		\item Add \incommand{wlan0} of \ac{sta}1 to get the setup of \ref{fig:snif2b}: \newline
		\command{\prompt{\ac{sta}1} ifconfig wlan0 up}
		\command{\prompt{\ac{sta}1} iw dev wlan0 connect essid wmn-\acs{gid}-A}
		\command{\prompt{\ac{sta}1} ip addr add fc00:grID::1/64 dev wlan0}
		\item Now, reintroduce \incommand{wlan1} of \ac{sta}1 to the network.\newline
		\command{\prompt{\ac{sta}1} ifconfig wlan1 up}
		\command{\prompt{\ac{sta}1} iw dev wlan1 connect wmn-\acs{gid}-A}
		\item Configure \incommand{wlan1} with the following IP address:\newline
		\command{\prompt{\ac{sta}1} ip addr add fc00:grID:1::1/64 dev wlan1}
		\remark Remark that this IP address belongs to a different subnet! This is in preparation of the next part of the exercise.
		\item Repeat steps \ref{ex:1-1} to \ref{ex:1-3} from exercise \ref{ex:firstScan} , but in step 2, start the capture only on \incommand{wlan1} of \ac{sta}1 and save it to \file{\acs{sta}1.A.pcap}.
		\item To change the \verb!wlan1! interface to another \ac{essid}, we first need another \ac{ap} managing this \ac{essid} (figure \ref{fig:snif2c}). Configure a second \ac{ap} interface on the \ac{ap} node, this time using ``\ac{wmn}-\ac{gid}-B'' as essid\\
		\remark Copy the previous hostapd.conf file and change the ssid and interface.
		\item Make sure \verb!wlan1! of \ac{sta}1 is connected to the \ac{wmn}-\ac{gid}-B network.
		\item Repeat the same exercise again, saving the trace to \file{\acs{sta}1.B.pcap}. \newline
		\item Compare the results from both tests. How do these two setups compare to a wired setup?\newline
		\begin{esolution}
		\end{esolution}
	\end{enumerate}

\end{exercise}

\begin{exercise}{Pinging the \ac{ap}}

In the previous exercises, we have used the \ac{ap} only as a L2 device. The \ac{ap} can of course also be configured as a L3 device. We will now configure the \ac{ap} as a L3 device and perform the same ping test again, but this time from \ac{sta}1 to \ac{ap}.
	\begin{enumerate}
		\item Add an IP address on \incommand{wlan0} of the \ac{ap}: \newline
		\command{\prompt{\ac{ap}} ip addr add fc00:grID:1::3/64 dev wlan0}
		\item On \ac{sta}1, start a packet capture on \incommand{wlan1} and save it to \file{\acs{sta}1.pcap}. \newline
		\command{\prompt{\ac{sta}1} tcpdump -i wlan1 -w \file{\acs{sta}1.pcap}} 
		\item Start a ping session from \ac{sta}1 to \ac{ap}. Limit the ping to only two requests. \label{ex:1-3}\newline
		\command{\prompt{\ac{sta}1} ping6 -c 2 fc00:grID:1::3} 
		\item In the trace files, do you observe duplicate entries? Why or why not?\newline
		\begin{esolution}
		\end{esolution}
	\end{enumerate}

\end{exercise}

\begin{exercise}{Sniffing in monitor mode}

\begin{figure}[h]
	\begin{center}
		\includegraphics[width=0.5\textwidth]{images/snif4.pdf} 
		\caption{Introducing a monitoring interface.}
		\label{fig:snif4} 
	\end{center}
\end{figure}

Until now, not much difference has been observed compared to network sniffing on a wired network - apart from observing duplicate packets. In the next exercises we will broaden the setups to show the actual differences. Firstly, the monitor interface will be introduced. Configuring a \acl{wnic} in monitor mode will enable you to see all traffic that is present in a certain channel. In figure \ref{fig:snif4} the next setup is shown. We continue from the previous setup and add an extra interface.
\begin{enumerate}
	\item Configure the monitoring device:\newline
	\command{\prompt{\ac{sta}2} iw dev wlan0 set type monitor}
	\item Bring the monitor interface up: \newline
	\command{\prompt{\ac{sta}2} ifconfig wlan0 up} 
	\item Configure the channel: \newline
	\command{\prompt{\ac{sta}2} iw dev wlan0 set channel x} 
	\remark Remark that it is not necessary to configure an \ac{essid}. Why?\newline
	\begin{esolution}
	\end{esolution}
	\item Clear the \ac{nd} caches:\newline
	\command{\prompt{\ac{sta}2} ip neigh flush dev wlan1}
	\command{\prompt{\ac{sta}1} ip neigh flush dev wlan0} 
	\item Start a capture session on the monitor interface and save it to \file{\acs{sta}2.pcap}:\newline
	\command{\prompt{\ac{sta}2} tcpdump -i wlan0 -w \file{\acs{sta}2.pcap}} 
	\item Start a ping session from \ac{sta}1 to \ac{sta}2. \label{ex:3-4}\newline
	\command{\prompt{\ac{sta}1} ping6 -c 2 fc00:\acs{gid}::2} 
	\item Open the capture file in Wireshark. 
	\begin{itemize}
		\item What type of frames are visible within the trace? Give an example (packet number) of each. \newline
		\begin{esolution}
		\end{esolution}
		\item What type of link layer headers  are visible now?\newline
		\begin{esolution}
		\end{esolution}
		\item A radiotap header should also be visible before the link layer header. This header is not actually transmitted, but is used to communicate transmission statistics between the driver and kernel. As such, it reports on various transmission statistics as the signal strength and the used antenna. Select a packet (give the packetID within the trace), and list which items are present. Although the format is standardized, the actual content depends on the driver and hardware used.\newline
		\begin{esolution}
		\end{esolution}
		\item Describe the path a packet takes to be delivered from one \acl{sta} to another. Give a detailed overview of the various addresses in the headers. Identify the frames (packet ID) from the trace you use to illustrate this.\newline
		\begin{esolution}
		\end{esolution}
	\end{itemize}
\end{enumerate}
\end{exercise}

Remember the traces you made containing duplicate packets? The duplicates can now be clearly observed in the traces from a monitor interface. Each transmission from one \acl{sta} to another, both connected to the same \ac{ap} is always relayed over the \ac{ap}. The sender thus first recorded its transmission and then detected the relayed frame on the air. Frames which are sent to an \ac{ap}, are discarded by other stations. This explains why a third station did not show duplicate packets. Finally, an \acl{ap} will only report one occurrence in \incommand{tcpdump} traces, as the relaying of a frame is done transparently to the kernel in the hardware/driver.

\begin{exercise}

Finally, we will repeat exercise \ref{ex:secondAP}. We will be sending traffic on both wireless networks and monitor the channel. The setup remains unchanged (see figure \ref{fig:snif4}).

\begin{enumerate}
	\item Clear the \ac{nd} caches as before. Be sure to delete all \ac{nd} entries on all \acp{wnic}!
	\item Start a scan on the monitor interface of \ac{sta}2 and save it to \file{\acs{sta}2.pcap}: \newline
	\command{\prompt{\ac{sta}2} tcpdump -i wlan0 -w \file{\acs{sta}2.pcap}}
	\item Perform a ping from \ac{sta}1 to \ac{ap} and start a ping from \ac{sta}2 to \ac{sta}1, both on \incommand{wlan1}. This means we will inject traffic on both wireless networks.\newline
	\remark These pings can be performed consecutively. Make sure they are in the same capture session. \newline
	\command{\prompt{\ac{sta}1} ping6 -c 2 fc00:\acs{gid}:1::3}
	\command{\prompt{\ac{sta}2} ping6 -c 2 fc00:\acs{gid}::1}
	\item Which ping session(s) is/are visible in the trace?\newline
	\begin{esolution}
	\end{esolution}
\end{enumerate}

\end{exercise}

\section{Ad-Hoc Networks}

In the next section, we will set up a basic ad-hoc network. Contrary to the infrastructure network we used in the previous section, ad-hoc networks are built from identically configured hosts, without a central entity in charge. Let's start right away to build our first ad-hoc network.

\begin{exercise}{Basic ad-hoc network}

\begin{enumerate}
	\item To ensure all interfaces are in the default state, reboot all devices:\newline
	\command{\prompt{\ac{ap}} reboot}
	\command{\prompt{\ac{sta}1} reboot}
	\command{\prompt{\ac{sta}2} reboot}
	\item  The basic setup is shown in figure \ref{fig:ad-hoc1}. All nodes will have the same configuration: mode, \ac{essid} and channel.\newline
	\command{\prompt{\ac{sta}1} iw dev wlan1 set type ibss}
	\command{\prompt{\ac{sta}2} iw dev wlan1 set type ibss}
	\command{\prompt{\ac{sta}3} iw dev wlan1 set type ibss}
	\begin{figure}
		\begin{center}
			\includegraphics[width=0.5\textwidth]{images/adhoc1.pdf} 
			\caption{Basic ad-hoc network.}
			\label{fig:ad-hoc1} 
		\end{center}
	\end{figure}
	\item Assign IP addresses to all interfaces and activate them:\newline
	\command{\prompt{\ac{sta}1} ip addr add fc00:\acs{gid}::1/64 dev wlan1}
	\command{\prompt{\ac{sta}2} ip addr add fc00:\acs{gid}::2/64 dev wlan1}
	\command{\prompt{\ac{sta}3} ip addr add fc00:\acs{gid}::3/64 dev wlan1}
	\command{\prompt{\ac{sta}1} ifconfig wlan1 up}
	\command{\prompt{\ac{sta}2} ifconfig wlan1 up}
	\command{\prompt{\ac{sta}3} iwconfig wlan1 up}
	\item Configure all interfaces with a \ac{essid} and a channel:\newline
	\command{\prompt{\ac{sta}1} iw dev wlan1 ibss join \ac{wmn}-\acs{gid}-A <frequency>}
	\command{\prompt{\ac{sta}2} iw dev wlan1 ibss join \ac{wmn}-\acs{gid}-A <frequency>}
	\command{\prompt{\ac{sta}3} iw dev wlan1 ibss join \ac{wmn}-\acs{gid}-A <frequency>}
	\remark The frequency you should use here can be found in the introduction section.\newline
	\item To monitor the traffic in the channel, set up a monitor interface on \ac{sta}1:\newline
	\command{\prompt{\ac{sta}1} iw dev wlan0 set type monitor}
	\command{\prompt{\ac{sta}1} ifconfig wlan0 up}
	\command{\prompt{\ac{sta}1} iw dev wlan0 set freq <frequency>}
	\item Scan using the monitor interface and save it to \file{\acs{sta}1.pcap}.\newline
	\command{\prompt{\ac{sta}1} tcpdump -i wlan0 -w \file{\acs{sta}1.pcap}}
	\item Verify that the nodes now all can reach each other:\newline
	\command{\prompt{\ac{sta}1} ping6 -c 2 fc00:\acs{gid}::2}
	\command{\prompt{\ac{sta}2} ping6 -c 2 fc00:\acs{gid}::3}
	\command{\prompt{\ac{sta}3} ping6 -c 2 fc00:\acs{gid}::1}	
	\item Describe how data is exchanged between the various hops. How does this compare to infrastructure mode? Motivate your findings by selecting frames (give the packet ID) from the trace file and describe their \ac{mac} header and addressing scheme.\newline
	\begin{esolution}
	\end{esolution}

	\item Repeat the ping tests from exercise \ref{ex:pingtest} (from \ac{sta}1 to \ac{sta}2 and write down the requested values. Do you observe a significant change in timings?\newline
	\begin{esolution}
	\end{esolution}
\end{enumerate}

\end{exercise}

### 2017-02-15 Lab_1/solutions/L1-2-1.tex
Filled in first three questions of lab 1.

Output of the command \textit{iwconfig wlan1}
\begin{verbatim}
wlan1     IEEE 802.11abgn  ESSID:off/any
          Mode:Managed  Access Point: Not-Associated   Tx-Power=0 dBm
          RTS thr:off   Fragment thr:off
          Encryption key:off
          Power Management:off
\end{verbatim}
\begin{itemize}
\item \textbf{IEEE 802.11agbn}: - This first field shows which wireless MAC technologies the device supports on that interface.\\
%
\item \textbf{ESSID:off/any}	- The second field specifies the network name (if any) of a network consisting of a group of cells connected via repeaters or infrastructure
 where users may roam between the cells. When this field is set to off or any, ESSID checking is disabled.\\
%
\item \textbf{Mode:managed}: Specifies the operating mode of the device and tells us what kind of network it operates in as well as what the function of the node is in the network.\\
I this case the mode is set to Managed which means that the node connects to a network composed of many access points where roaming is enabled.\\
%
\item \textbf{Access Point}: Specifies which access point the card is connected to. When configuring, it is possible to specify which access point the card should try to register to.
An access point can be chosen, it can be set to "any", "auto" or "off".
%
\item \textbf{Tx-power=0dBm}: For network cards supporting multiple transmit powers, this setting allows you to set the transmit power in dBm.
It also allows you to specify power control and radio using "auto", "fixed" and "on" or "off" respectively.\\
%
\item \textbf{RTS thr:off}: This parameter specifies when the RTS/CTS mechanism will be used and can be set to "auto", "fixed" and "off".
 RTS/CTS adds overhead but increases throughput in situaties with hidden nodes or a large number of active nodes.\\
%
\item \textbf{Fragment thr:off}: Similar to the RTS threshold this parameter can be "auto", "fixed" or "off" and allows you to specify the maximum fragment size.
Fragmentation can reduce error penalties in a noisy environment. On some cards, this threshold can be higher than the maximum packet size which will enable Frame Bursting.\\
%
\item \textbf{Encryption key:off}: This parameter allows manipulation of encryption, scrambling keys and security mode. It can be used to change security keys,
disable or enable encryption and change the security mode (implications of this are dependent on the type of card). \\
%
\item \textbf{Power Management:off}: Used to manipulate power management scheme parameters and mode. It allows you to disable or enable power management as well as specify which types
of packets should be received and choose the period between wake ups.\\
\end{itemize}

### 2017-02-22 Lab_1/solutions/L1-12-2.tex
Added some more Lab1 solutions and all Lab1 traces

Minimum: 0.853ms
Maximum: 1.536ms
Average: 1.194

Faster: not via AP but direct

### 2017-02-22 Lab_2/header.tex
Added Lab_2 sources

\documentclass[a4paper,onecolumn,12pt]{report}

\usepackage{a4wide}
\usepackage[printonlyused]{acronym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{array}
\usepackage{boxedminipage}
\usepackage{cite}
\usepackage{color}
\usepackage{fancyvrb}
\usepackage[draft]{fixme}
\usepackage{forloop}
\usepackage[pdftex]{graphicx}
\usepackage{helvet}
\usepackage[colorlinks,linkcolor=black,urlcolor=blue]{hyperref}
\usepackage{ifthen}
\usepackage{listings}
\usepackage{longtable}
\usepackage{subfig}
\usepackage{tikz}
\usepackage{verbatim}
\usepackage{xspace}

\newboolean{Solutions} \setboolean{Solutions}{true}

\renewcommand{\chaptername}{Lab}
\newcommand{\stress}[1]{\emph{\textbf{#1}}} 
\newcommand{\command}[1]{\texttt{#1}\newline} 
\newcommand{\incommand}[1]{\texttt{#1}}
\newcommand{\file}[2]{\texttt{/mnt/L\arabic{chapter}-\arabic{Exercisecount}-\stepcounter{questioncount}\arabic{questioncount}\addtocounter{questioncount}{-1}.#1}} 
\newcommand{\curfile}[2]{\texttt{/mnt/L\arabic{chapter}-\arabic{Exercisecount}-\stepcounter{questioncount}\arabic{questioncount}\addtocounter{questioncount}{-1}.#1}} 
\newcommand{\remark}{\includegraphics{images/remark.pdf}}
\newcommand{\report}{ \includegraphics{images/report.pdf}}

\newcommand{\result}[1]{
\begin{verbatim}
	#1 
\end{verbatim}
} 
\newcommand{\prompt}[1]{#1:\textasciitilde\#}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% exercise environment
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcounter{Exercisecount}[chapter]
\setcounter{Exercisecount}{0}
\newenvironment{exercise}[1]
{% This is the begin code
\refstepcounter{Exercisecount} {\textbf {Exercise \arabic{Exercisecount}}}: #1
}
{% This is the end code
 }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% question counter environment
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcounter{questioncount}[Exercisecount]
\setcounter{questioncount}{0}

\newcommand{\answerlabel}{L\arabic{chapter}-\arabic{Exercisecount}-\arabic{questioncount}}

\newcommand{\answer}{
\begin{tikzpicture}

\node [fill=shade,rounded corners=5pt]
{
\refstepcounter{questioncount}
\color{white}{\textbf{\answerlabel}}
};
\end{tikzpicture}
}

\newcommand{\includeanswer}{\input{solutions/\answerlabel.tex}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% verbatim solution environment
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\newcounter{linescounter}
%\newenvironment{esolution}[1]
%{
%\answer
%\ifthenelse{\boolean{Solutions}}
%{
%\color{blue}
%
%\hrulefill
%
%\includeanswer
%
%}
%{\setlength{\extrarowheight}{0.75cm} 
%
%\forloop{linescounter}{0}{\value{linescounter}<#1}{\quad\newline\vspace{12pt} \dotfill\newline}
%\comment
%}
%}

\newenvironment{esolution}
{
\answer
\ifthenelse{\boolean{Solutions}}
{
\color{blue}

\hrulefill

\includeanswer

}
{\setlength{\extrarowheight}{0.75cm} 
\quad\newline\vspace{12pt} \dotfill\newline
\comment
}
}
{
\ifthenelse{\boolean{Solutions}}
{

\hrulefill

}
{\endcomment}
}

\DefineVerbatimEnvironment{Verbatim}{Verbatim}
{formatcom=\color{blue},fontfamily=courier,fontseries=b,frame=lines,numbers=left}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% shortcuts
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\wifi}{IEEE 802.11\xspace}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% fonts
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\renewcommand{\rmdefault}{phv}
\renewcommand{\sfdefault}{phv}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% colors
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\definecolor{shade}{HTML}{3877A9}	%light blue shade
\setlength{\parskip}{10pt plus 1pt minus 1pt}

\lstset{ %
basicstyle=\footnotesize,       % the size of the fonts that are used for the code
numbers=left,                   % where to put the line-numbers
numberstyle=\footnotesize,      % the size of the fonts that are used for the line-numbers
stepnumber=2,                   % the step between two line-numbers. If it's 1, each line 
                                % will be numbered
numbersep=5pt,                  % how far the line-numbers are from the code
backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
showspaces=false,               % show spaces adding particular underscores
showstringspaces=false,         % underline spaces within strings
showtabs=false,                 % show tabs within strings adding particular underscores
frame=single,                   % adds a frame around the code
tabsize=2,                      % sets default tabsize to 2 spaces
captionpos=b,                   % sets the caption-position to bottom
breaklines=true,                % sets automatic line breaking
breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
title=\lstname,                 % show the filename of files included with \lstinputlisting;
                                % also try caption instead of title
escapeinside={\%*}{*)},         % if you want to add a comment within your code
morekeywords={*,...}            % if you want to add more keywords to the set
}

### 2017-02-22 Lab_2/labo2.tex
Added Lab_2 sources

%!TEX root = labo.tex
\setcounter{chapter}{1}
\chapter{Frequencies and Channels}

In the \wifi standards, various channels are defined. Two main frequency bands are used, namely 2.4 GHz and 5GHz. \wifi{a} networks use the 5 GHz band, while \wifi{b/g} networks operate in the 2.4 GHz band. In both bands, various ``channels'' have been defined. However, channel separation is not as clean cut as one might expect. In the following setups, you will illustrate this using some simple tests.

\section{Available channels}
\begin{exercise}{Frequencies}

Within the \wifi specification, various channels are defined. Depending on the local authorities (\ac{bipt} in Belgium\cite{bipt}), the list of allowed channels can vary. Using \incommand{iw} you can get the list of available channels.\newline
\remark The wireless drivers we use make a difference between the logical interface (\incommand{wlan0} as we have used it so far) and the actual physical radio interface. All things related to physical characteristics are actually handled by the physical interface, which is called \incommand{phy0} or \incommand{phy1}, respectively.\newline
\begin{enumerate}
	\item  To get a correct overview of which frequencies/channels are available in Belgium, use the command \incommand {iw phy1 info}.
	\item List all frequencies/channels that the \incommand{phy1} interface supports.\newline
	\begin{esolution}
	\end{esolution}
\end{enumerate}

You should observe several frequencies that are marked \emph{disabled}. This indicates that the card supports these frequencies, but cannot use them because of Belgian regulations. Also, several frequencies should be marked with \emph{radar detection}. These can be used, but special mechanisms must be put in place in order to avoid interference with radar installatiuons (e.g. airport radar).

\end{exercise}

\begin{exercise}{Available UA hotspots}
\begin{figure}[h]
	\begin{center}

		\includegraphics[width=0.5\textwidth]{images/channels1.pdf}
		\caption{Channel separation setup 1} 
		\label{fig:channels1} 
	\end{center}
\end{figure}

\begin{enumerate}
		\item Start by building the setup as illustrated in figure \ref{fig:channels1}. \newline
		\remark Remark that it is not necessary to configure the SSID or channel, but bring up the interface to activate it.
		\item Find out which \acp{ap} are active using \incommand{iw wlan0 scan}. This command provides an interface to list various settings from the \ac{wnic}. The manpage or command line help should be self-explanatory.
		\item Fill in the following table:\newline
		\begin{esolution}
		\end{esolution}
\ifthenelse{\not \boolean{Solutions}}{
		\renewcommand\arraystretch{2.5}
		\setlength\LTleft{0pt}
		\setlength\LTright{0pt}

		\begin{longtable}{@{\extracolsep{\fill}}ccc}
		%{\textwidth}{@{\extracolsep{\fill}}
		\hline
		SSID & Frequency (MHz) & BSS \\
		\hline
		& & \\
		\hline
		& & \\
		\hline
		& & \\
		\hline
		& & \\
		\hline
		\end{longtable}}{}
%		\item \report Write down the exact command used to obtain your results:
%		\begin{esolution}{2}
%		\end{esolution}
	\end{enumerate}
\end{exercise}

\section{Channel Separation}

\begin{exercise}{Channel overhearing} \label{ex:overhearing}

	\begin{figure}[h]
		\begin{center}
			\includegraphics[width=\textwidth]{images/beacon.pdf}
			\caption{Channels in a captured beacon frame.} 
			\label{fig:beacons} 
		\end{center}
	\end{figure}

In the previous exercise, you made an overview of a lot of \aclp{ap}. The next exercise will illustrate the channel overhearing. The information about which channel an \ac{ap} is working on, is provided in the beacon frames the \ac{ap} periodically transmits. In figure \ref{fig:beacons}, a Wireshark\cite{wireshark} screen shot shows a beacon frame. The Radiotap header indicates the frame was received on channel 5, while the beacon's content shows the \ac{ap} sending this beacon only operates at channel 6. Channels are thus not cleanly separated.  In the next exercise you'll create an overview of the channel overhearing.

\begin{figure}[h!]
		\begin{center}
			\includegraphics[width=0.5\textwidth]{images/channels2.pdf}
			\caption{Setup for exercise \ref{ex:overhearing}.} 
			\label{fig:channels2} 
		\end{center}
\end{figure}

	\begin{enumerate}
		\item Start from the setup as shown in figure \ref{fig:channels2}. \newline
		\remark Note that the \ac{ap} should be configured in channel 4 from the b/g range. To be able to do this, you should change the line \verb!hw_mode=a! to \verb!hw_mode=b! in the hostapd.conf file.
		\item On \ac{sta}1, on every channel, make a capture \file{chanID.pcap}.\newline
		\item Use the following wireshark display filter to only show beacons:\newline
		\command{wlan.fc.type\_subtype == 8}
		\item Fill out the following table using the obtained information.\newline
		\begin{esolution}
		\end{esolution}

\ifthenelse{\not \boolean{Solutions}}{
\begin{longtable}{@{\extracolsep{\fill}}c|l}	
	Selected channel & Observed channels in received beacons \\
	\hline
	1&\\\hline
	2&\\\hline
	3&\\\hline
	4&\\\hline
	5&\\\hline
	6&\\\hline
	7&\\\hline
	8&\\\hline
	9&\\\hline
	10&\\\hline
	11&\\\hline
\end{longtable}	}{}
		\end{enumerate}
\end{exercise}

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=\textwidth]{images/bgchannels.pdf}
		\caption{2.4 GHz channels} 
		\label{fig:channel_separation} 
	\end{center}
\end{figure}

From this exercise, it should be clear that the channels within the \wifi{b/g} range are not strictly separated.  Figure \ref{fig:channel_separation}\footnote{By Michael Gauthier, Wireless Networking in the Developing World [CC-BY-SA-3.0 (http://creativecommons.org/licenses/by-sa/3.0)], via Wikimedia Commons} gives you an idea why: the consecutive channels overlap to a certain extent. Therefore, a careful channel planning is crucial in building and deploying wireless networks on these frequencies. In the next exercise, we will take a look at the channel separation in the \wifi{a} band.

\begin{exercise}{Channel separation in \wifi{a}}\label{ex:separation}

	\begin{enumerate}
		\item Now, change the \ac{ap} so that it is on channel \verb!x!.
		\item Using \incommand{tcpdump} as in the previous exercise, find out in which channels \wifi{a} the beacons of this \ac{ap} can be seen. Save your traces in \file{chanID.pcap}.\newline
		\begin{esolution}
		\end{esolution}
	\end{enumerate}

\end{exercise}

\section{Using the Wireless Channel}

\begin{figure}[h]
		\begin{center}
			\includegraphics[width=0.5\textwidth]{images/adhoc2.pdf} 
			\caption{Basic ad-hoc network.}
			\label{fig:ad-hoc2} 
		\end{center}
	\end{figure}

\begin{exercise}{Beacons}

In the next few exercises, we will again use an ad-hoc network setup. It is best to reboot the devices before proceeding.
\begin{enumerate}
	\item Create the network setup as shown in figure \ref{fig:ad-hoc2}. Perform the following commands on all three stations. Substitute \incommand{nodeNumber} with a different number (e.g. 1, 2 and 3) on each node.\newline
		\command{\prompt{wmn} iw dev wlan0 set type ibss}
		\command{\prompt{wmn} ip addr add fc00:\acs{gid}::nodeNumber/64 dev wlan0}
		\command{\prompt{wmn} ifconfig wlan0 up}
		\command{\prompt{wmn} iw dev wlan0 ibss join \acs{wmn}-\acs{gid}-A <frequency>}
	\item Put interface wlan1 of \ac{sta}1 in monitor mode on the same frequency.\newline
		\command{\prompt{\ac{sta}1} iw dev wlan1 set type monitor}
		\command{\prompt{\ac{sta}1} ifconfig wlan1 up}
		\command{\prompt{\ac{sta}1} iw dev wlan1 set freq <frequency>}
	\item Check if all stations can reach each other.
	\item Perform a \incommand{tcpdump} on the monitor interface. Save it to \file{\acs{sta}1.pcap}. Let it run for a few seconds and then stop the trace.
	\item In your trace file, you should observe beacon frames. Carefully inspect one of the beacon frames in this trace and compare it to a beacon frame captured in the previous exercise (exercise \ref{ex:separation}). Give the packet IDs of the beacon frames you are comparing and identify the differences in the management frame part.\newline
	\begin{esolution}
	\end{esolution}
	\item Start a new trace on the monitor interface and save it to \file{STA1.pcap}.
	\item While the trace is running, perform two ping tests. You may perform the simultaneously, but it will be easier to answer the next question if you perform them consecutively.\newline
	\command{\prompt{\ac{sta}2} ping6 -c 5 fc00:\acs{gid}::3}
	\command{\prompt{\ac{sta}3} ping6 -s 1400 -c 5 fc00:\acs{gid}::1}
	\item Open your trace file and filter out the beacon frames with the following display filter: \verb#!(wlan.fc.type_subtype == 0x08)#. Apart from the data frames, which kind of frames do you observe?\newline
	\begin{esolution}
	\end{esolution}
	\item Explain the purpose of those frames.\newline
	\begin{esolution}
	\end{esolution}
\end{enumerate}
\end{exercise}

\begin{exercise}{\ac{rts}/\ac{cts}}

For this exercise, you will study the \ac{rts}/\ac{cts} mechanism. This mechanism will clear the channel for each transmission, minimizing the chance of collisions in the channel. A threshold value is used to determine if \ac{rts}/\ac{cts} should be used. Larger packets will receive \ac{rts}/\ac{cts} protections while smaller packet will not. Using \ac{rts}/\ac{cts} for small packets adds a lot of overhead and hence degrades performance.

\begin{enumerate}
	\item Start from the setup as for the previous exercise.
	\item The \ac{rts}/\ac{cts} threshold will be set with \incommand{iwconfig wlan0 rts 1000}. Perform this command on all nodes.
	\item Performing the same \incommand{ping6} commands as in the previous exercise. Start a capture session on the monitor interface and save it to \file{STA1.pcap}.
	\item You should observe \ac{rts}/\ac{cts} packets. Which packets are protected by this mechanism? Give an example (packet ID).\newline
	\begin{esolution}
	\end{esolution}
	\item Compare an \ac{rts} and \ac{cts} frame. How do they differ? Illustrate using frames from your last trace file.
	\begin{esolution}
	\end{esolution}
\end{enumerate}

As you can see in the \ac{rts}/\ac{cts} frames, they do not contain any information about the network on which they are transmitted. The \ac{rts}/\ac{cts} is meant to avoid collisions on the wireless medium, so any node that receives a \ac{cts} frame is required to remain silent for the duration included in the \ac{cts} frame. The only exception, for obvious reasons, is the node to which the \ac{cts} frame is sent. Capturing \ac{rts}/\ac{cts} frames (or acknowledgement frames) on a certain channel is always a good indication that some activity is going on in that channel, even if it is impossible to overhear the actual data transmission.
\end{exercise}

\section{\wifi{n}}

Until now, we have only used the \acp{wnic} in \wifi{a} or b/g mode. ``b'' is the oldest mode. ``g'' improves upon ``b'' by offering significantly higher maximal throughput (54 Mbps versus 11 Mbps) .``a'' operates at different frequencies and offers the same throughput as ``g''.

Another mode of operation is called ``n''. \wifi{n} is an amendment to the \wifi standard in order to improve throughput over both ``a'' and ``g''. It operates at both frequency bands and is defined for bit rates up to 600 Mbps, but very few setups will be able to obtain these speeds.

\wifi{n} can use different optimizations, such as the \ac{mimo} principle, 40 MHz wide channels (versus 20 MHz in \wifi{a/b/g}), and frame aggregation to improve throughput. The following exercise will demonstrate that \wifi{n} channels can indeed be twice as wide as those in \wifi{a}, and thus interfere with adjacent \wifi{a} channels.

\begin{exercise}{\wifi{n}}

\begin{figure}[h]
		\begin{center}
			\includegraphics[width=0.5\textwidth]{images/adhoc2.pdf} 
			\caption{\wifi{n} ad-hoc network.}
			\label{fig:ad-hoc-n} 
		\end{center}
	\end{figure}

\stress{As this exercise will use both channels allocated to the mobile and wireless lab, two groups cannot perform this exercise at the same time! Check if no other group is currently working on this course!}

\begin{enumerate}
	\item Reboot all devices. In order to enable \wifi{n} mode without any problems, we will start from a blank setup.
	\item After they have been rebooted, perform the following on all devices in order to create an ad-hoc network with \wifi{n} support in the 5 GHz band:\newline
		\command{\prompt{\ac{sta}} iw dev wlan0 set type ibss}
		\command{\prompt{\ac{sta}} ip addr add fc00:\acs{gid}::nodeNumber/64 dev wlan0}
		\command{\prompt{\ac{sta}} ifconfig wlan0 up}
		\command{\prompt{\ac{sta}} iw dev wlan0 ibss join wmn-\acs{gid}-A 5180 HT20}
	\item Check that all nodes can reach each other.
	\item Set up a monitor interface on channel 36 (5180 MHz) on \ac{sta}1.
	\item Start a trace on the monitor interface and save it to \file{\acs{sta}1.pcap}.
	\item Start a ping from \ac{sta}2 to \ac{sta}3 and from \ac{sta}3 to \ac {sta}1. Like before, use different ping sizes:\newline
		\command{\prompt{\ac{sta}2} ping6 -c 5 fc00:\acs{gid}::3}
		\command{\prompt{\ac{sta}3} ping6 -s 1400 -c 5 fc00:\acs{gid}::1}
	\item Filter out the beacon frames from your trace. Do you observe packets from both ping sessions? Did you see all packets from these sessions? Why or why not?\newline
	\begin{esolution}
	\end{esolution}
	\item Set the monitor interface to channel 40 (5200 MHz).
	\item Perform the same ping test again, but now save your trace to \file{\acs{sta}1.pcap}.
	\item Did you capture any packets?\newline
	\begin{esolution}
	\end{esolution}
\end{enumerate}

\end{exercise}

\begin{exercise}{40 MHz channels}

In the previous exercise, we enabled \wifi{n}, but we did not enable 40 MHz wide channels yet. We will do that now.
\begin{enumerate}
	\item Deconfigure all \incommand{wlan0} interfaces.\newline
	\command{\prompt{\ac{sta}} iw dev wlan0 ibss leave}
	\item Now, create an ad-hoc network that uses 40 MHz wide channels:\newline
	\command{\prompt{\ac{sta}} iw dev wlan0 ibss join wmn-0-A 5180 HT40+}
	\item As in the previous exercise, set the monitor interface of \ac{sta}1 on channel 36 and perform a capture while sending both pings. Save it to \file{\acs{sta}1.36.pcap}.
	\item Do the same again, but this time monitor channel 40.\newline Save your trace in \file{\acs{sta}1.40.pcap}.
	\item Do you observe any different behaviour in the trace file on channel 36, compared to the previous exercise?\newline
	\begin{esolution}
	\end{esolution}
	\item In the trace file on channel 40, which packets have you captured? Why?\newline
	\begin{esolution}
	\end{esolution}
\end{enumerate}
\end{exercise}

### 2017-02-23 Lab_3/labo3.tex
Made some preparations for the exercises in lab 3.
Former-commit-id: e4bfe43ad9d9fbbcb468d8b6b20611ef3282197f

%!TEX root = labo.tex
\setcounter{chapter}{2}
\chapter{Performance Measurements}

In this lab, the performance and throughput in wireless networks will be investigated. Using tools like \incommand{iperf}, we will record the maximum throughput which can be achieved in wireless networks and have a look at the parameters influencing this throughput.

\section{Bit Rates}

\begin{exercise}{Basic throughput in \wifi{a}}\label{ex:basicTput}
\label{ex:tput1}

	This first exercise will give you an insight into the difference in usable throughput and the available bit rate. To determine the throughput, we will be using a tool called \incommand{iperf}. This is a client-server based tool which sends \ac{tcp} or \ac{udp} traffic and reports the measured throughput. Consult the man pages for the details about this tool. A second tool to be used is \incommand{gnuplot}, in order to plot your results in a graph. More info about \incommand{gnuplot} can be found in \cite{gnuplotHome} and a nice tutorial in \cite{gnuplotTut}. A basic gnuplot script to generate your first plots is provided on the course website. The \incommand{iperf} tool is preinstalled on the wireless nodes, while \incommand{gnuplot} is available on the lab PCs.

	\begin{figure}[h!]
		\begin{center}
			\includegraphics[width=0.5\textwidth]{images/perf1.pdf}
			\caption{Basic throughput setup.} 
			\label{fig:perf1} 
		\end{center}
	\end{figure}

	\begin{enumerate}
		\item Start by configuring the setup as shown in figure \ref{fig:perf1}. Use \incommand{fc00:\acs{gid}::3/64} for the \ac{ap} and \incommand{fc00:\acs{gid}::1/64} for \ac{sta}1.
		\item \label{item:rates}\wifi supports various bit rates. Using \incommand{iw list} you can query the supported rates:\newline
		Which rates are supported in \wifi{a}?\newline
		\begin{esolution}
		\end{esolution}
		Which rates are supported in \wifi{b/g}?\newline
		\begin{esolution}
		\end{esolution}
		\item \incommand{iperf} can be used to measure the maximum throughput between two stations. Therefore, a server is started on one end and a client connects on the other end to this server. A connection is set up and as \ac{tcp} tries to maximize the throughput on a connection, an estimate of the maximum throughput on a link can be calculated. When using \incommand{iperf} with the basic parameters, it will perform a 10 seconds test and report the achieved throughput at the client side.
		\item Start a basic \incommand{iperf} session between the \ac{sta} and \ac{ap}, with the \incommand{iperf} server on the \ac{ap}: \newline
		\command{\prompt{AP} iperf -V -s}
		\command{\prompt{STA1} iperf -V -c fc00:\acs{gid}::3 }
		Copy the output of the \incommand{iperf} client:\newline
		\begin{esolution}
		\end{esolution}

		\item Now, using \incommand{iperf}, collect the throughput for each available rate and write the results in a file \file{tcp.txt}. On each line, first put the rate followed by a space and then the result in Mbit/s obtained from \incommand{iperf}, e.g. \texttt{54 30} denotes that a 30Mbit/s was measured when using a rate of 54 Mbps. You can change the rate used at the \ac{sta} using \incommand{iw}, e.g. to 54Mbps, as follows:\newline
		\command{\prompt{STA1} iw dev wlan0 set bitrates legacy-5 54}
		You can check the actual used bit rate from the output of \incommand{iwconfig}:\newline
		\command{\prompt{STA1} iwconfig wlan0}
		\remark As we will be repeating the collection of these results in the following exercises, it will be easier to use some bash scripting to speed up this process. 
		A helper script can be found in the file \texttt{iperf-tcp.sh}. Change this script so it loops over the correct bitrates, and use it with the command \incommand{./iperf-tcp.sh fc00:grID::3}.\newline
		This script generates output that can be directly copied to \file{tcp.txt}

		\item \incommand{iperf} will by default use \ac{tcp} to check the connection, but it is also possible to use a unidirectional \ac{udp} stream. Therefore, one can try to feed more data to the network than the network can support and as such measure how much can be actually delivered. Thus, repeat the previous scenario and collect the maximum achievable throughput using \incommand{iperf} in \ac{udp} mode for each available bit rate. Save these results in the same format as in the previous item in a file \file{udp.txt}.\newline Again, a script called \texttt{udp.sh} is provided that automates this process. The script will try to send a UDP stream that fills the wireless link.\newline
		\command{\prompt{AP} iperf -V -s -u -l 1452}
		\command{\prompt{STA1} ./iperf-udp fc00:grID::3}
		\remark \texttt{-l} is the letter l, not the number one!

		\item Now plot these results using \incommand{gnuplot}. On the course website, a \incommand{gnuplot} script, \texttt{tput1.gnuplot}, is provided to plot the throughput and the relative link usage over the various available rates. The used commands are straightforward and the script is inline commented so should be self-explanatory. Make sure you understand the various commands in the file. The script produces two PDF files which can be viewed with any regular PDF viewer. Place the \texttt{.txt} files you created in the previous steps in the same directory as the \incommand{gnuplot} script. Save the generated files in your lab report as \texttt{L3-1-4-tput.pdf} and \texttt{L3-1-4-usage.pdf}. Add the plots here and shortly discuss what can be observed. Also shortly discuss the difference between \ac{udp} and \ac{tcp}. To generate the PDF files, use:\newline
		\command{gnuplot tput.gnuplot}
		\begin{esolution}
		\end{esolution}
		\end{enumerate}
\end{exercise}

\begin{exercise}{Client to client throughput}
\label{ex:tput2}
	\begin{figure}[h!]
		\begin{center}

			\includegraphics[width=0.5\textwidth]{images/perf2.pdf}
			\caption{Client to client throughput setup.} 
			\label{fig:perf2} 
		\end{center}
	\end{figure}

	In the previous exercise, the throughput was measured between a \ac{sta} and the \ac{ap} it was connected to. In this exercise, we are interested in the throughput between two \acp{sta} associated with the same \ac{ap}.
	\begin{enumerate}
		\item Create the test setup as shown in figure \ref{fig:perf2}. Configure \acs{sta}2 with IP address \texttt{fc00:grID::2/64}.
		\item Collect the \ac{tcp} and \ac{udp} throughput measures as in the previous exercise and save them to \file{tcp.txt}\\ and \file{udp.txt}.
		\item Adapt the provided \incommand{gnuplot} script and create graphs for the new measurements. Include them in your report.
		\item Comment on the obtained results and compare the results from this exercise with those from the previous exercise. Explain the difference.\newline
		\begin{esolution}
		\end{esolution} 
	\end{enumerate}

\end{exercise}

\section{Network Settings}
\begin{exercise}{\acs{rts}/\acs{cts}}

In lab 2, we enabled the \ac{rts}/\ac{cts} mechanism to show that in \wifi, stations can reserve the medium in order to avoid collisions. In this exercise, we will study the effect of \ac{rts}/\ac{cts} on throughput.

\begin{enumerate}
	\item Enable \ac{rts}/\ac{cts} on all devices.\newline
	\command{\prompt{} iw phy phy0 set rts 1000}
	\item Repeat the iperf tests from \ac{sta}1 to \ac{ap}. Save your logs to \file{tcp.txt}\\ and \file{udp.txt}. Again modify the gnuplot script to create new graphs and include them in your report. To make the comparison more easy, modify the gnuplot script so that you plot both the results from exercise \ref{ex:tput1} and this exercise on the same graph. What do you observe? Why?\newline
	\begin{esolution}
	\end{esolution}

\end{enumerate}

\end{exercise}

\begin{exercise}{Frame length}

	The standard \ac{mtu} for ethernet frames is 1500 bytes. \wifi however allows an \ac{mtu} up to 2274 bytes. In this exercise, you will measure the effect of frame size on throughput.

	\begin{enumerate}
		\item Continue from the previous setup, but make sure \ac{rts}/\ac{cts} is turned off on all nodes:\newline
		\command{iw phy phy0 set rts off}
		\item On \ac{sta}1, reset rate control to its default (automatic) value:\newline
		\command{\prompt{\ac{sta}1} iw dev wlan0 set bitrates}
		\item Perform the following tests with 4 different maximum frame sizes: 1500, 1758, 2016 and 2274. The frame size can be controlled by changing the \ac{mtu}. \emph{If you do this, do this on all nodes!}\newline
		\command{ifconfig wlan0 mtu 1758}
		\item For each of the \acp{mtu} mentioned, start with a \ac{tcp} \incommand{iperf} test between \ac{sta} and \ac{ap} as in the previous exercises and make plots for each frame size. Generate a results file called \file{tcp.txt}\\ containing \ac{mtu} and bit rate achieved. The file should have the same format as the ones produced this far, except that the first column now contains your \ac{mtu} setting rather than the link speed.\newline
		\command{\prompt{\ac{sta}1} iperf -V -c fc00:grID::3}
		\item Repeat the tests, now using \ac{udp}. For \incommand{iperf}, use the \incommand {-l}  parameter, followed by the current MTU setting minus 48 (e.g. 1452 if the \ac{mtu} is set to 1500) to generate packets that are large enough to fill the MTU. \stress{This must be done on both client and server!} Save your results to \file{udp.txt}\\.\newline
		\command{\prompt{\ac{sta}1} iperf -V -u -l 1452 -c fc00:grID::3 -b 54M}
    		\item Modify the \incommand{gnuplot} script to generate the same graph as before. Plotting only the throughput versus \ac{mtu} suffices. A relative link usage graph is not required. What do you observe? Why? Be as precise as possible.\newline
		\begin{esolution}
		\end{esolution}	 
	\end{enumerate}

\end{exercise}

\section{Packet Size}
\begin{exercise}{IP Fragmenting}

	Changing frame lengths has an effect on the throughput, as you have shown in the previous exercise. However, in that case, traffic was flowing between segments with the  same fragment size. When we consider the default setup of an \ac{ap} which is connected to the Internet, the \ac{wan} link is most likely limited to 1500 bytes per frame and thus IP fragmenting comes into play. In IPv6, fragmenting in intermediate hops is not allowed. The end hosts may, however, fragment the IP packets end-to-end to make sure they fit on the link with the smallest \ac{mtu}. Depending on the original payload length, fragments of various lengths will be created. In this exercise we will have a look at the impact of fragmenting on the overall throughput.

	To do so, we will run the \incommand{iperf} server on the third node, which is only connected by wire to the \ac{ap}. Using additional routes, we will create a setup where \ac{sta}1 and \ac{sta}2 communicate via the \ac{ap}. Traffic between \ac{sta}1 and the \ac{ap} will be transmitted on the wireless link, while traffic between the \ac{ap} and \ac{sta}2 will be transmitted on the wired link.

	\begin{figure}[h!]
		\begin{center}
			\includegraphics[width=0.5\textwidth]{images/perf1.pdf}
			\caption{Fragmenting setup.} 
			\label{fig:fragmenting} 
		\end{center}
	\end{figure}

	\begin{enumerate}
		\item Start by configuring your setup as shown in figure \ref{fig:fragmenting}. \ac{sta}2 will only be used as \incommand{iperf} server and does not need an active \ac{wnic}.
		\item Set \texttt{fc00:\acs{gid}::3/64} as IP for the \ac{ap} and \texttt{fc00:\acs{gid}::1/64} for \ac{sta}1.
		\item Traffic from \ac{sta}1 to \ac{sta}2 will by default use the wired network. To force a wireless hop, we will add some new routes:\newline
		\command{\prompt{\ac{sta}2} ip route add fc00:\acs{gid}::/64 via <wired IP address of \acs{ap}>}
		\command{\prompt{\ac{sta}1} ip route add <wired IP address of \acs{sta}2>/128 via fc00:\acs{gid}::3}
		\item IPv6 forwarding is by default not enabled on the lab nodes. We must enable it on the \ac{ap} for the setup to work. However, due to a bug, the \ac{ap} will clear its default route when we do this, making it instantly inaccessible over IPv6. Perform the following as a workaround:\newline
		\command{\prompt{\ac{ap}} ip -6 route show | grep default}
		You should get output like this:\newline
		\texttt{default via fe80::225:90ff:fe61:efac dev eth0  proto kernel  metric 1024  expires 1670sec}
		\item Enable IPv6 routing functionality on the \ac{ap}:\newline
		\command{\prompt{\ac{ap}} sysctl -w net.ipv6.conf.all.forwarding=1 \&\& ip route add default via fe80::225:90ff:fe61:efac dev eth0}\newline
		The IP address in that command should be the same as that in the output of the previous step.
		\item Perform a \incommand{traceroute6} from \ac{sta}1 to \ac{sta}2 to check if the routes are set up correctly. Include your traceroute6 output below:\newline
		\command{\prompt{\ac{sta}1} traceroute6 <wired IP address of \acs{sta}2>}\newline
		\begin{esolution}
		\end{esolution}
		\item Now, for each \ac{mtu} (1500, 1758, 2016 and 2274), perform the following:\newline
		\begin{itemize}
			\item Set the \ac{mtu} on the wireless link.
			\item Perform a TCP \incommand{iperf} from \ac{sta}1 to \ac{sta}2.
			\item Perform a UDP \incommand{iperf} from \ac{sta}1 to \ac{sta}2. In each trace, set your frame size to the \ac{mtu} of the wireless link minus 48. Send at a bit rate of 54 Mbps.
		\end{itemize}
		\item The results of your tests should once again be saved to text files like in the previous exercise. Save them to \file{tcp.txt}\\ and \file{udp.txt}\\. Create a throughput graph as in the previous exercise.
		\item You should also make \texttt{.pcap} trace files from these experiments. However, as the link gets fully saturated, the trace file would become very large. Therefore, we will repeat the experiments sending far less traffic. This of course means that you will not see a difference in \incommand{iperf} results. The traces will help you, however, to understand what has happened in the previous exercise.
		\item In order to limit the amount of traffic, we will do the following on \ac{sta}1:
		\begin{itemize}
		\item Set the rate of \texttt{wlan0} to 6 Mbps.
		\item for both TCP and UDP \incommand{iperf}, add the \texttt{-t 2} parameter. This limits the \incommand{iperf} trace to two seconds.
		\item for UDP \incommand{iperf}, omit the \texttt{-b 54M} parameter. This way, \incommand{iperf} will only transmit at 1 Mbps.
		\end{itemize}
		\item Now, do the same \incommand{iperf} tests again, using the above parameters for \incommand{iperf}. For each run, make a \incommand{tcpdump} trace on interface \texttt{wlan0} of the \ac{ap}. Save the traces to \file{tcp.<MTU size>.pcap}\\ and \file{udp.<MTU size>.pcap}\\, respectively.
		\item What do you observe? Distinguish between the behaviour for TCP and UDP. Explain your findings by including your graphs and refering to the trace files you made. Be as precise as possible.\newline
		\begin{esolution}

		\end{esolution}

	\end{enumerate}	
\end{exercise}

### 2017-02-23 Lab_3/iperf.sh
Made some preparations for the exercises in lab 3.

@@ -1,0 +1,6 @@
+#!/bin/bash
+for i in seq 2 4 6 8 10 12 14 16 18 20; do
+    echo $1
+    iw dev wlan0 set bitrates legacy-5 i
+    iperf -V -c $1
+done

### 2017-02-23 Lab_1/solutions/L1-5-1.tex
Filling in parts of Lab_1, some questions remain about L1-5-1 and L1-6-2, L1-10-2 onward still to be done.
Former-commit-id: 338fad62136bf05ef12a7e88fdfaaee88eba11b4

\begin{itemize}

\item Neighbor Solicitation:  STA1 is trying to find STA2 by broadcasting these
\item Neighbor Advertisement: STA2 telling STA1 it's L2 address
\item Ping Request from STA1 to STA2
\item Ping Reply from STA2 to STA1
%\item More Neighbor Solicitations and Advertisements follow, refreshing the previous bindings. Interestingly, these have source and destination different.
\end{itemize}

### 2017-02-24 Lab_3/solutions/L3-2-1.tex
2nd lab session, working on lab3 and lab4

the tcp connection waits for acks while sending which leads to less collisions than in the udp case which makes it better for low bitrates.

as the bitrates go up, collisions and retransmissions increase and tcp starts to suffer and udp pulls out ahead (affects tcp more than it does udp).

because both tcp and udp have to be sent twice, first from STA1 to the AP and then from the AP to STA2, the achieved throughput is about half of the achieved throughput in the first experiment.

### 2017-02-24 Lab_3/solutions/L3-3-1.tex
2nd lab session, working on lab3 and lab4

The resulting udp and tcp streams (with rts/cts) have a throughput that is consistently lower than their non-rts/cts counterparts. Adding the rts/cts mechanism does not really help because the theoretical datarate is hard to achieve under normal circumstances.

in this case it is no different and we see that the added overhead of rts/cts slows down the realised throughput of the stream even more.

### 2017-02-24 Lab_4/solutions/L4-1-2.tex
2nd lab session, working on lab3 and lab4

278: broadcasted probe request for the ssid wmn-1-A sent by STA1
279: the answer is from AP to STA1 and contains the ssid as well

in the association request <pkt id> the ssid is still listed. 
further communication such as association requests also contain the ssid

only broadcasts are now gone such as pkt 5 in the first L411

-- only authentication seen

### 2017-02-24 Lab_3/solutions/L3-1-4.tex
2nd lab session, working on lab3 and lab4
Former-commit-id: 39aa658f16b0f0f14f691a0cf38b4b29e0343fa4

udp is hoger : minder overhead, gn congestion control dus resultaten zijn als verwacht
throughput stijgt omdat de bitrate ook omhoog gaat

usage verlaagt: sneller sturen -> meer retransmits en minder usage

### 2017-02-24 Lab_3/solutions/L3-4-1.tex
2nd lab session, working on lab3 and lab4
Former-commit-id: 39aa658f16b0f0f14f691a0cf38b4b29e0343fa4

udp hogere frame length helpt minder, meer botsingen dan verliest er meer dus die capt op een bepaald moment en begint bij 2274 ook weer te dalen

tcp hogere frame length zorgt er voor dat voor constante hoeveelheid overhead er meer data verzonden kan worden dus die blijft in stijgende lijn gaan, ook al is de bereikte tput nog steeds lager dan udp

### 2017-02-24 Lab_4/solutions/L4-1-1.tex
2nd lab session, working on lab3 and lab4
Former-commit-id: 39aa658f16b0f0f14f691a0cf38b4b29e0343fa4

5 broadcast beacon frame

2528 probe 
2529 probe reply
2694 association request van sta1 naar ap waar ssid in staat

### 2017-02-24 Lab_4/solutions/L4-2-1.tex
2nd lab session, working on lab3 and lab4
Former-commit-id: 39aa658f16b0f0f14f691a0cf38b4b29e0343fa4

in de eerdere kreeg authentication als reply succesful, nu is het unspecified failure
<pkt ids>

### 2017-02-24 Lab_4/solutions/L4-2-2.tex
2nd lab session, working on lab3 and lab4
Former-commit-id: 39aa658f16b0f0f14f691a0cf38b4b29e0343fa4

Mac address spoof. either two mac addresses will be on the network causing problems and crashing the network or the spoofed mac address will not be present at the time granting the attacker access to the network.

### 2017-02-26 Lab_1/solutions/L1-10-2.tex
Finished Lab1 answers. Added .gitignore for latex files
Former-commit-id: db97bdc772bc07c8d7e33e5fe4f66942f69d6ec3

There are three main types of frames observed, but each can be divided into separate subtypes. 
\begin{itemize}
\item Management frames
\begin{itemize}

\item Acknowledgement (e.g. 9)

\item Probe Request (e.g. 15, for SSID UAntwerpen)
\item Beacon Frame (e.g. 29, for SSID UAntwerpen)

\item Probe response (e.g. 313, for SSID wmn-1-A)
\item Action (e.g. 1172)
\end{itemize}

\item Control frames
\begin{itemize}
\item Request-to-send (e.g. 1)
\item Clear-to-send (e.g. 7)
\item 802.11 Block Ack (e.g. 4)
\item Acknowledgement (e.g. 9)
\end{itemize}
\item Data frames
\begin{itemize}
\item QoS Null function (e.g. 47)
\item Null function (e.g. 1178)
\item Data (e.g. 1967, these are the Neighbor soliciations/advertisements and pings)
\end{itemize}
\end{itemize}
There were also some 802.11ac-specific frames such as the VHT NDP Announcement (e.g. 12)

### 2017-02-26 Lab_1/solutions/L1-10-3.tex
Finished Lab1 answers. Added .gitignore for latex files
Former-commit-id: db97bdc772bc07c8d7e33e5fe4f66942f69d6ec3

Now we do see 802.11 headers, and no more Ethernet headers. In monitor mode the driver or network stack no longer converts the headers to Ethernet. Each frame contains the radio information header with some network statistics. This seems to contain roughly the same info as the Radiotap header described below, and is also a pseudo-header. \\
What follows is the 802.11 header. The precise set of fields it contains depends on the (sub)type of the frame. The data frame headers seem to contain all possible fields: The frame control field containing a type + subtype byte, a duration field, source and destination MAC addresses of initial sender and final receiver, source and destination MAC address of current transmission, the AP's MAC, the STA's MAC, a fragment number, a sequence number and a frame check sequence. \\
Only the data frames also have a Logical Link Control header. It identifies used L2 protocols and the type of L3 data contained.

### 2017-02-26 Lab_1/solutions/L1-12-1.tex
Finished Lab1 answers. Added .gitignore for latex files
Former-commit-id: db97bdc772bc07c8d7e33e5fe4f66942f69d6ec3

Let's again consider the first ping request (packet 3878). It contains the following addresses.
\begin{lstlisting}
    Receiver address: Sparklan_26:13:c2 (00:0e:8e:26:13:c2)
    Destination address: Sparklan_26:13:c2 (00:0e:8e:26:13:c2)
    Transmitter address: Sparklan_20:83:df (00:0e:8e:20:83:df)
    Source address: Sparklan_20:83:df (00:0e:8e:20:83:df)
    BSS Id: 8a:89:24:d2:26:b8 (8a:89:24:d2:26:b8)
\end{lstlisting}

The source and transmitter addresses and the destination and receiver addresses are the same. This indicates the frame was sent directly from original source to final destination without any intermediary stops. There is no longer an AP. The BSS Id is still filled in but is a random address. %http://sourcedaddy.com/networking/bssid-ssid-and-essid.html
The STA address is no longer part of the header. \\ \\
Note that there are no longer any duplicates as there is no AP to repeat the frames.

### 2017-02-26 Lab_1/solutions/L1-2-1.tex
Finished Lab1 answers. Added .gitignore for latex files
Former-commit-id: db97bdc772bc07c8d7e33e5fe4f66942f69d6ec3

@@
 \item \textbf{Mode:managed}: Specifies the operating mode of the device and tells us what kind of network it operates in as well as what the function of the node is in the network.\\
-I this case the mode is set to Managed which means that the node connects to a network composed of many access points where roaming is enabled.\\
+In this case the mode is set to Managed which means that the node connects to a network composed of many access points where roaming is enabled.\\
 %
@@
 \item \textbf{RTS thr:off}: This parameter specifies when the RTS/CTS mechanism will be used and can be set to "auto", "fixed" and "off".
- RTS/CTS adds overhead but increases throughput in situaties with hidden nodes or a large number of active nodes.\\
+ RTS/CTS adds overhead but increases throughput in situations with hidden nodes or a large number of active nodes.\\
 %

### 2017-02-26 Lab_1/solutions/L1-6-1.tex
Finished Lab1 answers. Added .gitignore for latex files
Former-commit-id: db97bdc772bc07c8d7e33e5fe4f66942f69d6ec3

STA1's trace file (traces/L1-5-1.STA1.pcap) shows duplicate ping requests (e.g. packets 3 and 4) and some duplicate neighbor advertisements (e.g. 10 and 11) and solicitations (e.g. 12 and 13).\\
According to the src MAC addresses, these are all packets sent by STA1.\\ \\
Similarly, for the STA2 dump (traces/L1-5-1.STA2.pcap), everything that is sent by STA2 is duplicated. \\
This includes the neighbor advertisement in packets 2 and 3; the ping replies in 5 and 6; the neighbor solicitation in 10 and 11.

### 2017-02-26 Lab_1/solutions/L1-10-4.tex
Finished Lab1 answers. Added .gitignore for latex files

This is the fully expanded radiotap header of packet 1977 (ping request).

\begin{lstlisting}
Radiotap Header v0, Length 26
    Header revision: 0
    Header pad: 0
    Header length: 26
    Present flags
        .... .... .... .... .... .... .... ...1 = TSFT: Present
        .... .... .... .... .... .... .... ..1. = Flags: Present
        .... .... .... .... .... .... .... .1.. = Rate: Present
        .... .... .... .... .... .... .... 1... = Channel: Present
        .... .... .... .... .... .... ...0 .... = FHSS: Absent
        .... .... .... .... .... .... ..1. .... = dBm Antenna Signal: Present
        .... .... .... .... .... .... .0.. .... = dBm Antenna Noise: Absent
        .... .... .... .... .... .... 0... .... = Lock Quality: Absent
        .... .... .... .... .... ...0 .... .... = TX Attenuation: Absent
        .... .... .... .... .... ..0. .... .... = dB TX Attenuation: Absent
        .... .... .... .... .... .0.. .... .... = dBm TX Power: Absent
        .... .... .... .... .... 1... .... .... = Antenna: Present
        .... .... .... .... ...0 .... .... .... = dB Antenna Signal: Absent
        .... .... .... .... ..0. .... .... .... = dB Antenna Noise: Absent
        .... .... .... .... .1.. .... .... .... = RX flags: Present
        .... .... .... .0.. .... .... .... .... = Channel+: Absent
        .... .... .... 0... .... .... .... .... = MCS information: Absent
        .... .... ...0 .... .... .... .... .... = A-MPDU Status: Absent
        .... .... ..0. .... .... .... .... .... = VHT information: Absent
        ...0 0000 00.. .... .... .... .... .... = Reserved: 0x00000000
        ..0. .... .... .... .... .... .... .... = Radiotap NS next: False
        .0.. .... .... .... .... .... .... .... = Vendor NS next: False
        0... .... .... .... .... .... .... .... = Ext: Absent
    MAC timestamp: 66921462
    Flags: 0x10
        .... ...0 = CFP: False
        .... ..0. = Preamble: Long
        .... .0.. = WEP: False
        .... 0... = Fragmentation: False
        ...1 .... = FCS at end: True
        ..0. .... = Data Pad: False
        .0.. .... = Bad FCS: False
        0... .... = Short GI: False
    Data Rate: 48,0 Mb/s
    Channel frequency: 5180 [A 36]
    Channel flags: 0x0140, Orthogonal Frequency-Division Multiplexing (OFDM), 5 GHz spectrum
        .... .... ...0 .... = Turbo: False
        .... .... ..0. .... = Complementary Code Keying (CCK): False
        .... .... .1.. .... = Orthogonal Frequency-Division Multiplexing (OFDM): True
        .... .... 0... .... = 2 GHz spectrum: False
        .... ...1 .... .... = 5 GHz spectrum: True
        .... ..0. .... .... = Passive: False
        .... .0.. .... .... = Dynamic CCK-OFDM: False
        .... 0... .... .... = Gaussian Frequency Shift Keying (GFSK): False
        ...0 .... .... .... = GSM (900MHz): False
        ..0. .... .... .... = Static Turbo: False
        .0.. .... .... .... = Half Rate Channel (10MHz Channel Width): False
        0... .... .... .... = Quarter Rate Channel (5MHz Channel Width): False
    SSI Signal: -50 dBm
    Antenna: 1
    RX flags: 0x0000
        .... .... .... .... .... ..0. = Bad PLCP: False
\end{lstlisting}
We'll only go over the actual statistics in the header and not every field. \\
There are some flags identifying the network's configuration in the Flags field (no WEP, fragmentation disabled, ...)\\
The data rate (48Mbps) and channel frequency (5180MHz, channel 36) are listed. \\
Then there are some channel flags identifying channel properties such as used multiplexing and modulating techniques. \\
Finally there is some info on signal strength and the used antenna.

### 2017-02-26 Lab_1/solutions/L1-10-5.tex
Finished Lab1 answers. Added .gitignore for latex files

The ping requests clearly show how the packets are transmitted. Consider the first request in packets 1975 and 1977. The ping is from STA1 to STA2.

The first contains these addresses
\begin{lstlisting}
    Receiver address: Sparklan_20:83:df (00:0e:8e:20:83:df)
    Destination address: Sparklan_26:13:b0 (00:0e:8e:26:13:b0)
    Transmitter address: Routerbo_12:1b:a8 (d4:ca:6d:12:1b:a8)
    Source address: Routerbo_12:1b:a8 (d4:ca:6d:12:1b:a8)
    BSS Id: Sparklan_20:83:df (00:0e:8e:20:83:df)
    STA address: Routerbo_12:1b:a8 (d4:ca:6d:12:1b:a8)
\end{lstlisting}
The source and destination addresses are of STA1 and STA2 respectively. These show the initial source and the final destination of the frame. Conversely, the transmitter and receiver addresses are of the interface that actually transmitted this frame and the interface that should receive it. Here they are STA1 and the AP, respectively. The BSS Id is the AP's address. Here it equals the receiver address so other stations know to ignore this frame. Finally the STA address is that of the involved station, here being STA1. \\ \\

These are the addresses for the second occurrence of the same request.
\begin{lstlisting}
    Receiver address: Sparklan_26:13:b0 (00:0e:8e:26:13:b0)
    Destination address: Sparklan_26:13:b0 (00:0e:8e:26:13:b0)
    Transmitter address: Sparklan_20:83:df (00:0e:8e:20:83:df)
    Source address: Routerbo_12:1b:a8 (d4:ca:6d:12:1b:a8)
    BSS Id: Sparklan_20:83:df (00:0e:8e:20:83:df)
    STA address: Sparklan_26:13:b0 (00:0e:8e:26:13:b0)
\end{lstlisting}
As this is still the same request logically, the source, destination and BSS Id addresses are the same as before. What did change are the transmitter and receiver addresses: these are now the AP's and STA2's respectively. The STA address is now that of STA2. This shows that the frame was first sent to the AP by STA1, and then repeated to STA2 by the AP.

### 2017-02-26 Lab_1/solutions/L1-12-2.tex
Finished Lab1 answers. Added .gitignore for latex files

@@
-Minimum: 0.853ms
-Maximum: 1.536ms
-Average: 1.194
+Minimum RTT: 0.853ms\\
+Maximum RTT: 1.536ms\\
+Average RTT: 1.194ms\\\\
 
-Faster: not via AP but direct
+Looking at the average RTT, this is 12\% faster than in managed mode.
+The frame no longer travels via the AP.
+The additional physical distance to be covered by the signal, plus the processing time at the AP explains why the RTT in exercise 4 is higher.
+\\ \\
+Note that the physical distance between STA1 and AP, and between STA1 and STA2 are roughly the same.
+This means the direct transmission in ad-hoc mode for this specific ping does not have to travel significantly farther than when the AP was involved.

### 2017-02-26 Lab_3/solutions/L3-2-1.tex
Added Lab3 answers and questions file

@@
-the tcp connection waits for acks while sending which leads to less collisions than in the udp case which makes it better for low bitrates.
-
-as the bitrates go up, collisions and retransmissions increase and tcp starts to suffer and udp pulls out ahead (affects tcp more than it does udp).
-
-because both tcp and udp have to be sent twice, first from STA1 to the AP and then from the AP to STA2, the achieved throughput is about half of the achieved throughput in the first experiment.
+\includegraphics[width=0.5\textwidth]{traces/L3-2-1-tput.pdf}
+\includegraphics[width=0.5\textwidth]{traces/L3-2-1-usage.pdf}
+For low bitrates TCP performs better than UDP.
+Due to TCP's congestion control there are fewer collisions compared to UDP.
+For lower bitrates this makes TCP perform better than UDP.
+For higher bitrates the influence of TCP's additional overhead becomes larger and UDP is more performant than TCP again.
+\\ \\
+For both TCP and UDP, throughput in this case is roughly half of the previous test.
+Each packet is transmitted twice (to the AP and from the AP), halving the total throughput between source and destination.
+\\ \\
+ The effect of increasing bitrate on the usage is similar to the previous test.

### 2017-02-26 Lab_3/solutions/L3-4-1.tex
Added Lab3 answers and questions file

@@
-udp hogere frame length helpt minder, meer botsingen dan verliest er meer dus die capt op een bepaald moment en begint bij 2274 ook weer te dalen
-
-tcp hogere frame length zorgt er voor dat voor constante hoeveelheid overhead er meer data verzonden kan worden dus die blijft in stijgende lijn gaan, ook al is de bereikte tput nog steeds lager dan udp
+\includegraphics[width=0.5\textwidth]{traces/L3-4-1-tput.pdf} \\
+UDP has a fairly small header per datagram.
+As overhead per datagram is already fairly limited, there is not much room for optimization by putting more data in each datagram.
+Additionally a larger datagram is more likely to get lost in a wireless network than a smaller one.
+We observe this when increasing the MTU from 2016 to 2274: the additional packet loss due to the larger frame size outweighs the decrease in per-bit overhead, leading to a lower throughput.
+ \\ \\
+In the case of TCP there is more per-bit overhead, meaning there is larger room for improvement by putting more data in each segment.
+Increasing the MTU consistently increases throughput for the tested range: additional packet loss never outweighs the performance increase of reducing the per-bit overhead.
+This is however never enough to overtake UDP performance.

### 2017-02-26 Lab_3/solutions/L3-1-4.tex
Added Lab3 answers and questions file
Former-commit-id: 46eacfcfc8e8444bb6fd165479eac20335ab3f98

@@
-udp is hoger : minder overhead, gn congestion control dus resultaten zijn als verwacht
-throughput stijgt omdat de bitrate ook omhoog gaat
-
-usage verlaagt: sneller sturen -> meer retransmits en minder usage
+\includegraphics[width=0.5\textwidth]{traces/L3-1-4-tput.pdf}
+\includegraphics[width=0.5\textwidth]{traces/L3-1-4-usage.pdf}
+UDP's throughput is consistenly higher than TCP's.
+TCP has larger headers so more overhead.
+More importantly, TCP's congestion control mechanisms lead to reduced throughput.
+Throughput grows together with bitrate: the network supports sending more bits per second with higher bitrates.
+\\ \\
+The throughput does not grow as quickly as the bitrate: with higher bitrates the bits are sent faster and bits are more likely to get lost.
+Due to this additional loss doubling the bitrate does not double the observed throughput.
+This translates to a decreasing usage with increasing bitrate.

### 2017-02-26 Lab_3/solutions/L3-3-1.tex
Added Lab3 answers and questions file
Former-commit-id: 46eacfcfc8e8444bb6fd165479eac20335ab3f98

@@
-The resulting udp and tcp streams (with rts/cts) have a throughput that is consistently lower than their non-rts/cts counterparts.
-Adding the rts/cts mechanism does not really help because the theoretical datarate is hard to achieve under normal circumstances.
+\includegraphics[width=0.5\textwidth]{traces/L3-3-1-tput.pdf}
+\includegraphics[width=0.5\textwidth]{traces/L3-3-1-usage.pdf}
 
-in this case it is no different and we see that the added overhead of rts/cts slows down the realised throughput of the stream even more.
+For both TCP and UDP throughput is consistently lower with RTS/CTS enabled.
+The only communication in our tests is direct traffic between STA1 and AP (only STA1 $\rightarrow$ AP in case of UDP).
+There were few collisions even without RTS/CTS.
+This means RTS/CTS mainly adds overhead, leading to lower throughput.
+\\ \\ Even with RTS/CTS enabled UDP still outperforms TCP: TCP's lower performance is caused by its additional overhead and congestion control, which is not changed by having RTS/CTS enabled.

### 2017-02-27 Lab_2/solutions/L2-6-2.tex
Working on Lab 2, filled in up to 4, half-assed 5 and 6: still has to be looked at for finishing. 802.11 (7-8) left as is for now because the results were inconclusive.
Former-commit-id: bd8cc1a2c8a31bdc0890d43ce4838e97a8e3ab36

Comparing the packets in 1611 (RTS) and 1612 (CTS), we immediately see that the RTS is targetted to the desired receiver while the CTS is broadcast. The RTS message is 6 bytes longer than the CTS message and the important differences are in the final field: for RTS it is 802.11 Request To Send and for CTS it is IEEE 802.11 Clear To Send.
The difference in size is because the L2 address of the transmitter of the CTS packet is not included in the CTS, while it is included for the RTS. Aside from that, not much differences can be observed: subtypes and such are different as expected.

### 2017-02-27 Lab_2/solutions/L2-3-1.tex
Working on Lab 2, filled in up to 4, half-assed 5 and 6: still has to be looked at for finishing. 802.11 (7-8) left as is for now because the results were inconclusive.
Former-commit-id: 37ac7684f000734feb181df1407cfca635225d9d [formerly 37ac7684f000734feb181df1407cfca635225d9d [formerly bd8cc1a2c8a31bdc0890d43ce4838e97a8e3ab36]]
Former-commit-id: 70f9d6ffe5623aee9e82549a39cc1f24d93efe24
Former-commit-id: 1f552c2cdc9567cf305dcbeca05b35bb858cfc6f

\begin{tabular}{l | l}
Selected channels & Observed channels in received beacons \\ \hline
1 & 1,4\\ \hline
2 & 1,4\\ \hline
3 & 1,4,6\\ \hline
4 & 4,6\\ \hline
5 & 4,6\\ \hline
6 & 4,6\\ \hline
7 & 6\\ \hline
8 & 4,6\\ \hline
9 & 11,4,6\\ \hline
10 & 11,4\\ \hline
11 & 11,4\\
\end{tabular}

### 2017-02-27 Lab_2/solutions/L2-5-1.tex
Working on Lab 2, filled in up to 4, half-assed 5 and 6: still has to be looked at for finishing. 802.11 (7-8) left as is for now because the results were inconclusive.
Former-commit-id: 37ac7684f000734feb181df1407cfca635225d9d [formerly 37ac7684f000734feb181df1407cfca635225d9d [formerly bd8cc1a2c8a31bdc0890d43ce4838e97a8e3ab36]]
Former-commit-id: 70f9d6ffe5623aee9e82549a39cc1f24d93efe24
Former-commit-id: 1f552c2cdc9567cf305dcbeca05b35bb858cfc6f

For this exercise, we will be comparing packet 2 from traces/L2-4-1.36.pcap and packet 2 from traces/L2-5-1.STA1.pcap.
For the purpose of comparing the WLAN management frames, I have included them below:

For the beacon frame captured in  exercise 4:
\begin{lstlisting}
IEEE 802.11 wireless LAN management frame
    Fixed parameters (12 bytes)
        Timestamp: 0x00000000082f503b
        Beacon Interval: 0,102400 [Seconds]
        Capabilities Information: 0x0001
            .... .... .... ...1 = ESS capabilities: Transmitter is an AP
            .... .... .... ..0. = IBSS status: Transmitter belongs to a BSS
            .... ..0. .... 00.. = CFP participation capabilities: No point coordinator at AP (0x0000)
            .... .... ...0 .... = Privacy: AP/STA cannot support WEP
            .... .... ..0. .... = Short Preamble: Not Allowed
            .... .... .0.. .... = PBCC: Not Allowed
            .... .... 0... .... = Channel Agility: Not in use
            .... ...0 .... .... = Spectrum Management: Not Implemented
            .... .0.. .... .... = Short Slot Time: Not in use
            .... 0... .... .... = Automatic Power Save Delivery: Not Implemented
            ...0 .... .... .... = Radio Measurement: Not Implemented
            ..0. .... .... .... = DSSS-OFDM: Not Allowed
            .0.. .... .... .... = Delayed Block Ack: Not Implemented
            0... .... .... .... = Immediate Block Ack: Not Implemented
    Tagged parameters (28 bytes)
        Tag: SSID parameter set: wmn-1-A
            Tag Number: SSID parameter set (0)
            Tag length: 7
            SSID: wmn-1-A
        Tag: Supported Rates 6(B), 9, 12(B), 18, 24(B), 36, 48, 54, [Mbit/sec]
            Tag Number: Supported Rates (1)
            Tag length: 8
            Supported Rates: 6(B) (0x8c)
            Supported Rates: 9 (0x12)
            Supported Rates: 12(B) (0x98)
            Supported Rates: 18 (0x24)
            Supported Rates: 24(B) (0xb0)
            Supported Rates: 36 (0x48)
            Supported Rates: 48 (0x60)
            Supported Rates: 54 (0x6c)
        Tag: DS Parameter set: Current Channel: 36
            Tag Number: DS Parameter set (3)
            Tag length: 1
            Current Channel: 36
        Tag: Traffic Indication Map (TIM): DTIM 1 of 0 bitmap
            Tag Number: Traffic Indication Map (TIM) (5)
            Tag length: 4
            DTIM count: 1
            DTIM period: 2
            Bitmap control: 0x00
                .... ...0 = Multicast: False
                0000 000. = Bitmap Offset: 0x00
            Partial Virtual Bitmap: 00
\end{lstlisting}
%
for the beacon frame captured in exercise 5:
\begin{lstlisting}
IEEE 802.11 wireless LAN management frame
    Fixed parameters (12 bytes)
        Timestamp: 0x000000000a53c0aa
        Beacon Interval: 0,102400 [Seconds]
        Capabilities Information: 0x0002
            .... .... .... ...0 = ESS capabilities: Transmitter is a STA
            .... .... .... ..1. = IBSS status: Transmitter belongs to an IBSS
            .... ..0. .... 00.. = CFP participation capabilities: Station is not CF-Pollable (0x0000)
            .... .... ...0 .... = Privacy: AP/STA cannot support WEP
            .... .... ..0. .... = Short Preamble: Not Allowed
            .... .... .0.. .... = PBCC: Not Allowed
            .... .... 0... .... = Channel Agility: Not in use
            .... ...0 .... .... = Spectrum Management: Not Implemented
            .... .0.. .... .... = Short Slot Time: Not in use
            .... 0... .... .... = Automatic Power Save Delivery: Not Implemented
            ...0 .... .... .... = Radio Measurement: Not Implemented
            ..0. .... .... .... = DSSS-OFDM: Not Allowed
            .0.. .... .... .... = Delayed Block Ack: Not Implemented
            0... .... .... .... = Immediate Block Ack: Not Implemented
    Tagged parameters (32 bytes)
        Tag: SSID parameter set: wmn-1-A
            Tag Number: SSID parameter set (0)
            Tag length: 7
            SSID: wmn-1-A
        Tag: Supported Rates 6(B), 9, 12(B), 18, 24(B), 36, 48, 54, [Mbit/sec]
            Tag Number: Supported Rates (1)
            Tag length: 8
            Supported Rates: 6(B) (0x8c)
            Supported Rates: 9 (0x12)
            Supported Rates: 12(B) (0x98)
            Supported Rates: 18 (0x24)
            Supported Rates: 24(B) (0xb0)
            Supported Rates: 36 (0x48)
            Supported Rates: 48 (0x60)
            Supported Rates: 54 (0x6c)
        Tag: IBSS Parameter set: ATIM window 0x0
            Tag Number: IBSS Parameter set (6)
            Tag length: 2
            Atim Windows: 0x0000
        Tag: Vendor Specific: Microsof: WMM/WME: Information Element
            Tag Number: Vendor Specific (221)
            Tag length: 7
            OUI: 00-50-f2
            Vendor Specific OUI Type: 2
            Type: WMM/WME (0x02)
            WME Subtype: Information Element (0)
            WME Version: 1
            WME QoS Info: 0x00
                0... .... = U-APSD: Disabled
                .... 0000 = Parameter Set Count: 0x00
                .000 .... = Reserved: 0x00
\end{lstlisting}
%
A first difference we observed is that the capabilities information beacon frame captured
in exercise 5 contains different capabilities information than the one captured in exercise 4:
the ex5 beacon frame shows that the sender of the beacon frame is a STA and that the network is
an ad-hoc network, while in exercise 4 the capabilities reflect that the transmitter is an AP
and the network is not ad-hoc.\\
For the tagged parameters section, the ex5 beacon is bigger. It contains some IBSS parameters (ATIM) and vendor specifici information about the ad-hoc network.
 In the ex4 beacon, the tagged parameters contain information about the channel and the Traffic Indication Map.

### 2017-02-27 Lab_2/solutions/L2-5-3.tex
Working on Lab 2, filled in up to 4, half-assed 5 and 6: still has to be looked at for finishing. 802.11 (7-8) left as is for now because the results were inconclusive.

The purpose of each of these frames:
\begin{itemize}
\item Request-To-Send packet
\item Clear-To-Send packet
\item 802.11 Block Ack
\item Acknowledgement, ie 2259
\item Probe Response ie 2601
\item Probe Request ie 2740
\end{itemize}

### 2017-03-01 Lab_4/labo4.tex
Preparing for Lab 4 lab session
Former-commit-id: 049acdc0e1549609e708b3f6f0ae3e9b32f4a579

%!TEX root = labo.tex
\setcounter{chapter}{3}
\chapter{Wireless Security}

In this lab, you will have a look at the impact of various security measures which are available for wireless networks. The main goal is to show what the effect is of a certain security method and what a possible attacker/sniffer can do on the protected network. In the first lab, you already performed some sniffing using the monitor mode. It should be clear that unless we take some security measures, it is quite easy to intercept and read any management information and data from open networks. It is even quite easy to perform \ac{dos} attacks on open networks as you will see at the end of this lab.

In the first part of the lab, you will investigate the more basic security features like hiding a network and \ac{mac} filtering. In the second part, you will have a look at encryption mechanisms and try to hack some of these. In the last part, we will introduce some basic attacks. These attacks are introduced to illustrate the mechanisms behind wireless networks and should only be used for this purpose.

\section{Basic Security Measures}

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=0.5\textwidth]{images/macfilter.pdf}
		\caption{MAC filter setup} 
		\label{fig:macfilter} 
	\end{center}
\end{figure}

\begin{exercise}{Network hiding}

A first step in preventing someone to use your wireless network is to hide it from being detected by a regular scan. Hiding a \ac{ssid} ensures it will not show up in the list of available \acp{ap} when a scan is performed. You will however show that the clever sniffer is still able to detect the network.

\begin{enumerate}
	\item Set up your network as illustrated in figure \ref{fig:macfilter}, but do not set up the \ac{wnic} of \ac{sta}1 yet. \label{ex:ssidHiding1}
	\item Start a trace on the monitor interface (\file{\acs{sta}2.pcap}) of \ac{sta}2. While the trace is running, bring up the \ac{wnic} from \ac{sta}1. \label{ex:ssidHiding2} Make sure the station is associated with the \ac{ap} before you stop your trace.
	\item Indicate in which frames the \ac{ssid} is visible. Give a packet ID of an example packet in your trace file for each.\newline
	\begin{esolution}
	\end{esolution} 
	\item The next step is to hide the \ac{ssid}. This can be done adding the following line to the hostapd.conf file: \incommand{ignore\_broadcast\_ssid=1}. Bring the interface down on \ac{sta}1, restart \incommand{hostapd} on the \ac{ap} and repeat steps \ref{ex:ssidHiding1} and \ref{ex:ssidHiding2}. Save your trace to \file{STA2.pcap}.
	\item Compare the two trace files and indicate how the \ac{ssid} is hidden. In what way can a \ac{ssid} still be detected?\newline
	\begin{esolution}
	\end{esolution}
\end{enumerate}

\end{exercise}

\begin{exercise}{\ac{mac} filtering}

Continuing from the setup you used in the previous exercise, a possible intruder has still complete access to our network. In the next step, you will prevent him from joining the network by \ac{mac} filtering.

\begin{enumerate}
	\item The management of the \ac{mac} filtering is also performed by modifying the \incommand{hostapd.conf} file at the \ac{ap}.  The entries needed for \ac{mac} filtering are:
	\begin{description}
		\item[macaddr\_acl=] controls the behaviour of the mac filtering  and expects an integer value from the following list:
		\begin{description}
			\item[0] accept unless in deny list,
			\item[1] deny unless in accept list,
			\item[2] use external RADIUS server (accept/deny lists are searched first
		\end{description}
		\item[accept\_mac\_file=] append the name of the file here that contains the list of \ac{mac} addresses to accept.
		\item[deny\_mac\_file=] append the name of the file here that contains the list of \ac{mac} addresses to deny.
	\end{description} 
	\item Configure the \ac{ap} in such a way that the \ac{mac} address of the \texttt{wlan0} interface of \ac{sta}1 is blocked from the network. SSID broadcasting may be enabled again.
	\item Now bring down the \ac{wnic} at \ac{sta}1 again, restart a capture session on \ac{sta}2 and save it to \file{\acs{sta}2.pcap}\\ and bring the \ac{wnic} back up at \ac{sta}1 when this capture session is active. Compare the results from this capture file with the first one you made and discuss the differences between both. Illustrate with packet IDs.\newline
	\begin{esolution}
	\end{esolution}
	\item Of course, in a real world setup, blocking unwanted MAC addresses is not feasible. One would use the option with the \texttt{accept\_mac\_file} parameter, allowing only known \ac{mac} addresses to connect. How would you circumvent this security measure? If an attacker has joined the network, will this have an effect on legitimate stations on that network?\newline
	\begin{esolution}
	\end{esolution}
\end{enumerate}

\end{exercise}

\section{Encryption}
\subsection{\acs{wep}}

The security techniques from the previous section will ``prevent'' a possible attacker from joining the network by hiding the network or preventing him access to the network. However, it is still possible to read any communication over the network by using a \ac{wnic} in monitor mode. Encryption mechanisms will counter this and obfuscate the network traffic for an eavesdropper. You will take a look at \ac{wep} here as this method is still sometimes being used although it is very vulnerable to attacks. \ac{wpa}2 is far more resilient.

\begin{exercise}{\ac{wep} encryption}
\label{ex:wep}

	\ac{wep} was the first available encryption method, standardised together with the \wifi standard. It was however quickly been proven insecure. In fact, as of 2004, \ac{wep} was declared deprecated by IEEE because of its flaws. In this exercise, you will first set up a \ac{wep} encrypted network to check what is exactly encrypted. In the next exercise you will crack the encryption key.

Because \ac{wep} is quite old, the tools you will use assume that IPv4 is used, as they will use \ac{arp} messages. Therefore, for this exercise, you will configure IPv4 addresses instead of IPv6 addresses.

\begin{figure}[h!]
		\begin{center}
			\includegraphics[width=0.5\textwidth]{images/wepcrack.pdf}
			\caption{WEP setup} 
			\label{fig:wepcrack} 
		\end{center}
	\end{figure}

	\begin{enumerate}
		\item Start from the network setup as shown in figure \ref{fig:wepcrack}. Disable any \ac{mac} filter that is still in place.
		\item In order to avoid capturing needless IPv4 traffic, disable the \texttt{avahi-daemon} on all three nodes:\newline
		\command{\prompt{\acs{wmn}} killall avahi-daemon}
		\item Create a tracefile on the monitor interface of \ac{sta}2 (\file{\ac{sta}2.cap}\\), containing the scan, authentication and association phase of \ac{sta}1 and a 5 packet ping session between \ac{sta}1 and \ac{sta}2. Use \texttt{10.0.\acs{gid}.1/24} for \ac{sta}1 and \texttt{10.0.\acs{gid}.2/24} for \ac{sta}2.\newline
		\command{\prompt{\acs{sta}} ip addr add 10.0.\acs{gid}.X/24 broadcast 10.0.\acs{gid}.255 dev wlan1}
		\item Make sure that \ac{sta}2 is connected to the \ac{ap} and that the \ac{wnic} of \ac{sta}1 is down before the start of the capture session. To ping, use \incommand{ping} instead of \incommand{ping6}. This trace will be our reference trace. \label{wep1} 
		\item Activate the \ac{wep} encryption on the \ac{ap}. You can choose a random key as long as it is either 5 or 13 characters long ("arandomwepkey" in this example):\newline
		Change the \incommand{hostapd.conf} file, so that it includes the following lines:\newline
		\command{wep\_key0="arandomwepkey"}
		\command{wep\_default\_key=0}
		\remark The WEP key must be \stress{exactly 5 or 13} characters long!
		\item Make sure the \ac{ap} is running. Connect \ac{sta}2 to the \ac{ap}:\newline
		\command{\prompt{\acs{sta}2} iw dev wlan1 connect wmn-0-A key 0:arandomwepkey}
		\item Make sure the interface of \ac{sta}1 is down and then start a new trace on the monitor interface of \ac{sta}2. Save it to \file{\acs{sta}2.wep.pcap}\\.
		\item Perform the same test as in step \ref{wep1}, but be sure to add \texttt{key 0:arandomwepkey} when you connect \ac{sta}1 to the network.
		\item Compare both trace files. In the second trace file, you should no longer be able to see the \incommand{ping} session in the clear.
		\begin{itemize}
			\item Which frame types are encrypted?\newline
			\begin{esolution}
			\end{esolution}
			\item Which part of a frame is encrypted and which part is still readable?\newline
			\begin{esolution}
			\end{esolution}
		\end{itemize} 
		Refer to specific frames from the second trace file to illustrate your answers.

	\end{enumerate}

\end{exercise}

\begin{exercise}{Cracking the \ac{wep} keys}

	Without going into details about the flaws in  \ac{wep}, the main problem arises from the fact that traffic keys are easily repeated on a busy network. A traffic key is the key actually used by the encryption algorithm RC4.  As RC4 is a stream cipher, the same key should not be used twice. Therefore, the shared \ac{wep} key, which is configured in each \ac{ap} and \ac{sta}, is combined with a random \ac{iv}. This \ac{iv} is transmitted in plaintext between two wireless clients so the receiver can recreate the encryption key used for a specific packet, by combining the chosen (shared) network key with the received \ac{iv}. As the \ac{iv} has a length of only 3 bytes, the number of possibilities is relatively low and in busy networks, the same \acp{iv} are frequently reused, weakening the cryptographic strength of \ac{wep}. Furthermore, statistical relations between used keys and encrypted text can be exploited to find the used key.

	Various methods have been developed to crack a \ac{wep} key, but you will just illustrate the mere ease with which a \ac{wep} encrypted network can be hacked using the aircrack tool \cite{aircrackNG}. This tool implements various approaches, but you will limit ourselves to \ac{ptw} method. The attack is based on the fact that the first 16 bytes of an \ac{arp} packet are fixed. This is also the reason we switched to IPv4 for this exercise\ldots For more details read \cite{cryptoeprint:2007:120}, specifically sections 1 and 5 for a general overview. Exploiting the relation between cleartext and the captured \acp{iv} makes it possible to quite easily determine the network key.

	\begin{figure}[h!]
		\begin{center}
			\includegraphics[width=0.5\textwidth]{images/wepcrack2.pdf}
			\caption{WEP cracking setup} 
			\label{fig:wepcrack2} 
		\end{center}
	\end{figure}

	\begin{enumerate}
		\item Before you start to crack the configured \ac{wep} key, first have a second look at the traces from the previous exercise. Using Wireshark\cite{wireshark}, you can decrypt an encrypted trace. Using the menu Edit$\rightarrow$ Preferences $\rightarrow$ Protocols $\rightarrow$ IEEE 802.11 you can specify the key in HEX. To convert your key to HEX, use e.g. the following website: \url{http://www.dirtymonday.net/key_convert.html} For our example it will be \incommand{6172616E646F6D7765706B6579}.\newline
Make sure you also check the `decrypt packets' box. Select an ICMP request in both trace files (unencrypted and encrypted, but decrypted by Wireshark) and identify the extra information added to a frame in order to encrypt it. Also indicate where in the frame this information is added.\newline
		\begin{esolution}
		\end{esolution}
		\item Start from the \ac{wep} encrypted network you created in the previous exercise.
		\item Add an additional monitor interface on \ac{sta}1, resulting in the network scheme shown in figure \ref{fig:wepcrack2}. The monitor interface on \ac{sta}2 will be used to perform the actual attack. \ac{arp} packets will be injected into the network in order to trigger \ac{arp} responses which will contain new \acp{iv} with every response . These packets will be captured and stored in files. In parallel with this capturing of new \ac{arp} responses and the associated \acp{iv}, we can start the process of analyzing these packages and searching for the key. The monitor interface on \ac{sta}1 will be used to monitor the channel so we can afterwards take a look at what happened.
		\item Start a packet capture to \file{\acs{sta}1.pcap}\\ on the monitor interface of \ac{sta}1.
		\item Start the \ac{iv} capture process at \ac{sta}2. This command will store its captured frames in files \incommand{output-xy.cap} and \incommand{output-xy.txt}. Make sure to run this command with your active directory set to the \texttt{/mnt} folder, in which a remote directory is mounted.\newline
		\command{\prompt{STA2} airodump-ng --band a --channel x --bssid <\acs{mac} addr \acs{ap}> -w output wlan0}
		This should give you an overview screen where at least the incoming beacon count is rising. If no activity is shown, bring the interface down and then bring it up again and retry. This command will store its captured frames in files \incommand{output-xy.cap} and \incommand{output-xy.txt}. Make sure to run this command in \texttt{/mnt}, where a remote location is mounted.\newline
		\item On a new terminal, perform a fake authentication to the \ac{ap} from the attacking machine (\ac{sta}2). This ensures the frames we will be injecting for \texttt{wlan1} will be accepted by the \ac{ap}:\newline
		\command{\prompt{STA2} aireplay-ng --fakeauth 0 -e wmn-\acs{gid}-A -a <\acs{mac} addr \acs{ap}> -h <\acs{mac} addr wlan0> wlan0.}	
		\item It is now time to actively inject packets into the wireless network. This is done by the following command:
		 \command{\prompt{STA2} aireplay-ng --arpreplay -b <MAC addr AP> -h <MAC addr wlan0> -x 1024 -o 512 wlan0}\newline
		 This command will monitor incoming packets and when an \ac{arp} request passes, it will re-inject it at a rate of several 100 frames/s.  You should be able to see this in the output from \incommand{airodump-ng}.\newline
		 \remark Due to a bug in the driver, the packet injection rate will not be high enough. You should run 10 of the above \incommand{aireplay-ng} commands in parallel to generate enough packets. You can do so easily by editing the \texttt{wepcrack.sh} script found on the nodes. Insert the correct \ac{mac} addresses in that script and run it. To stop the replay attack, simply type \incommand{killall aireplay-ng}.
		  \item In order to get an \ac{arp} request, just start a ping from \ac{sta}1 to \ac{ap}. This should trigger an \ac{arp} request. If the \ac{arp} tables already contained the IP addresses from both hosts (check it with \incommand{\prompt{STA1} ip neigh}), remove the entry using \incommand{ip neigh flush dev wlan1}.
		 \item To crack the key, the tool \incommand{aircrack-ng} can be used. This tool processes the files generated by \incommand{airodump-ng}. The needed number of packets to crack the key depends on the bit length of the key. In general, you need at least 20,000 packets for a 64-bit key and between 40,000 and 85,000 for a 128-bit key. To start the analysis, run the following command in the folder containing the output from \incommand{airodump-ng}:\newline
		\command{\prompt{STA2} aircrack-ng -b  <\acs{mac} addr \acs{ap}> output*.cap -q}
		\item If you captured enough packets, after a while, the \incommand{aircrack-ng} tool should provide you with the \ac{wep} key used in your network. Copy its output here:\newline
		\begin{esolution}
		\end{esolution}
		\item Using the following \incommand{tshark} command on a lab PC, you can get a list of \acp{iv} used during your session. \newline
		\command{tshark -r <trace file> -T fields -e wlan.wep.iv}
		\item How many \acp{iv} did you capture? And how many of them occured at least twice?\newline
		\begin{esolution}
		\end{esolution}	
	\item When handing in the report, you must include the \texttt{output-*.csv} and \texttt{*.netxml} files. \texttt{output-*.cap} is not required, as it will be quite large. Limit the trace file of \ac{sta}1 to only a few (e.g. a hundred) of packets that show the \ac{arp} injection in progress.
	\end{enumerate}

\end{exercise}

\subsection{\acs{wpa}2}

As you have demonstrated in the previous exercise, \ac{wep} encryption is easily circumvented. You will now take a look at \ac{wpa}2 encryption. You will no try to crack it, as \ac{wpa}2 uses the \ac{ccmp} algorithm with \ac{aes} encryption, which is extremely resilient to attacks. To give you an idea, the best known method until now to crack a 128 bit \ac{aes} key, which is used by \ac{wpa}2, still takes at least \( 2^{126.1} \) computations\cite{cryptoeprint:2011:449}.

A security flaw concerning \ac{wps} has been discovered that makes \acp{ap} running \ac{wps} vulnerable to a relatively easy attack (which would still take far too much time for this lab exercise)\cite{wps}. This, of course, does not hold true for \acp{ap} that do not have \ac{wps} enabled. If you want to read up on how \ac{wpa}2 works and on the evolution in security from \ac{wep} over \ac{wpa} to \ac{wpa}2, you should read \cite{wifisecurity:2012}.

\begin{exercise}{\acs{wpa}2}

\begin{figure}[h!]
		\begin{center}
			\includegraphics[width=0.5\textwidth]{images/wepcrack.pdf}
			\caption{\acs{wpa} setup} 
			\label{fig:wpa} 
		\end{center}
	\end{figure}

\begin{enumerate}
	\item Start from the \ac{wpa}2 enabled network setup as seen in figure \ref{fig:wpa}.
	\item On the \ac{ap}, to enable \ac{wpa}2, make sure that the line \texttt{wpa=0} in the \texttt{hostapd.conf} file is changed to \texttt{wpa=2}, and add the following lines to \texttt{hostapd.conf}:
	\begin{verbatim}
		wpa_passphrase=aRandomPassphrase
		wpa_key_mgmt=WPA-PSK
		rsn_pairwise=CCMP
	\end{verbatim}
	\remark Use your own passphrase if you like. Unlike \ac{wep}, it is not restricted to a certain amount of characters.
	\item Set the IP addresses of \texttt{wlan1} on \ac{sta}1 and \ac{sta}2 to \texttt{fc00:\acs{gid}::1/64} and \texttt{fc00:\acs{gid}::2/64}, respectively.
	\item Connect \ac{sta}2 to the network. In order to do so, modify the \texttt{wpa\_supplicant.conf} file so that it includes the correct \acs{essid} and passhprase. Afterwards, run:\newline
	\command{\prompt{\acs{sta}2} wpa\_supplicant -B -c/root/wpa\_supplicant.conf -iwlan1}
	\remark You might get an error like \texttt{ioctl[SIOCSIWENCODEEXT]: Invalid argument}. You can safely ignore this error.
	\item Make sure the monitor interface on \ac{sta}2 is configured and start a trace. Save it in \file{\acs{sta}2.pcap}\\.
	\item Now bring the interface on \ac{sta}1 up an do \texttt{ping6 -c 5} to \ac{sta}2.
	\item Which frame types are encrypted? Give examples.\newline
		\begin{esolution}
		\end{esolution}
	\item Which part of a frame is encrypted and which part is still readable?\newline
		\begin{esolution}
		\end{esolution}
		\item Compare this trace file with the one made in exercise \ref{ex:wep}. Compare the authentication/association phase of \ac{sta}1. What are the similarities and what are the differences? Refer to packet IDs!\newline
		\begin{esolution}
		\end{esolution}
		\item Now, compare an encrypted frame in both traces. Indicate where they differ.\newline
		\begin{esolution}
		\end{esolution}
\end{enumerate}

\end{exercise}

\section{Denial of Service Attacks}

\begin{figure}[h!]
		\begin{center}
			\includegraphics[width=0.5\textwidth]{images/wepcrack2.pdf}
			\caption{DoS setup} 
			\label{fig:dos} 
		\end{center}
	\end{figure}

In the previous section, you have gained access to a \ac{wep} encrypted network by actively injecting packets and cracking the key. This kind of attack makes it possible that you, the attacker, can afterwards associate with the network and sniff all traffic on that network. Alternatively, you can also use the network yourself as if you were a normal user.

Within this section, you will illustrate a \ac{dos} attack that shows the vulnerability of a wireless network. You will not gain access to the network as such, but you will disturb normal network operation. Tools to do so are freely available and current encryption techniques cannot prevent these attacks as they make use of management frame injection. As you have seen, management frames are not encrypted and thus can be easily injected into a wireless network.

\begin{exercise}{Disassociation Attack}

	The principle behind a disassociation attack is quite easy. A disassociation frame is sent by either a \ac{sta} or \ac{ap} to signal the disconnection from the wireless network. For instance, it is sent by a \ac{sta} just before you disable an associated \ac{wnic}. Although the sending of a disassociation message is not mandatory by the standard, at reception of such frame, a client should consider its connection lost and an \ac{ap} will remove a \ac{sta} from the list of connected stations. Injecting a malicious disassociation frame for a wireless node will thus force this node to scan for and associate again to its \ac{ap}, resulting in a connectionless period. If the disassociation frame is sent periodically, this can render the wireless network useless to the attacked \ac{sta}.

\begin{enumerate}
	\item Start from the network scheme as illustrated in figure \ref{fig:dos}.
	\item Be sure to have configured your network with \ac{wpa}2 encryption.
	\item Configure the IP addresses of \ac{sta}1 and \ac{sta}2  to \texttt{fc00:\acs{gid}::1/64} and\newline \texttt{fc00:\acs{gid}::2/64} respectively.
	\item On the monitor interface of \ac{sta}1, start a capture session and save it to\newline \file{\acs{sta}1.pcap}\\.
	\item Start a \incommand{ping6} from \ac{sta}1 to \ac{sta}2.
	\item On \ac{sta}2, use the following command to send a disassociation to \ac{sta}1:\newline
	\command{\prompt{STA2}  aireplay-ng --deauth 1 wlan0 -a <MAC addr AP> -c <MAC addr STA1>}
	\item What is the result for the ping?\newline
	\begin{esolution}
	\end{esolution}
	\item Open the trace file and explain using this trace what exactly happened.\newline
	\begin{esolution}
	\end{esolution}

\end{enumerate}

\end{exercise}

### 2017-03-01 Lab_4/solutions/L4-4-1.tex
Lab 4 exercises done, traces to come.

Difference : WEP parameters 4 bytes + 4 bytes that dont show up when you look at wep params in the pkt
IV en key index voor data, WEP ICV na data maar voor fcs

### 2017-03-01 Lab_2/solutions/L2-7-1.tex
Attempt at further completing Lab_2 report. TODO: review.
Former-commit-id: 2d360e1e49356ed1878920c2d29d20af597bf555

We observed packets of both ping sessions, from the first session only the replies were caught on the monitor interface and for the second session, all ping packets were seen by the monitor.\\
%
Not seeing the ping requests of the first ping session could be due to failed demodulations at the monitor, causing it to be not be able to detect the modulated
messages sent by STA1 in the first ping session. Because the second ping transmission does originate at STA2, it is less of an issue and that ping traffic is captured without issue.
%
The sniffer machine is not in range of the target wireless devices, or there's too much interference, so the sniffer machine's radio
can't successfully demodulate the transmissions. Note that the increasing use of beamforming helps devices steer their RF energy right
to where the intended receiver is, which means less of that energy goes to where the sniffer machine is, meaning the sniffer will
get a worse signal:noise ratio, and less likelihood of demodulating packets sent at the highest data rates.

### 2017-03-01 Lab_2/solutions/L2-8-1.tex
Attempt at further completing Lab_2 report. TODO: review.
Former-commit-id: 2d360e1e49356ed1878920c2d29d20af597bf555

No, the behaviour is largely the same: we again capture some icmp traffic but it is again not the complete conversation.
Because the stations are using the 36 channel (albeit with 40 MHz bands) and we are capturing on channel 36, we capture some icmp packets.
%
The 40 MHz wide bands indicate that the network has a "main" channel, in this case being 36, and extra space to be used if need be.
Because our icmp traffic is not enough to fill up the main channel, the extra space remains (largely) unused and we thus observe the same behaviour as in the previous exercise where we only capture part of the icmp exchange.

### 2017-03-01 Lab_2/solutions/L2-5-3.tex
Attempt at further completing Lab_2 report. TODO: review.
Former-commit-id: a602a8aac6f200eb09671acfd35ed294a12bdd7a [formerly 4bb19f2939594df7590ee676ee1dffbd10f93aa3 [formerly 4bb19f2939594df7590ee676ee1dffbd10f93aa3 [formerly 2d360e1e49356ed1878920c2d29d20af597bf555]]]
Former-commit-id: 290b42e00fb01a08084458b8079fb2136b9c82d4
Former-commit-id: 78b53e3af44825658c0ba968adbf66a6dd8d5bad
Former-commit-id: 475a6d34e4d1bc5ffbd59fb146dc2c133d900a0f

@@
 \begin{itemize}
-\item Request-To-Send packet
-\item Clear-To-Send packet
-\item 802.11 Block Ack
-\item Acknowledgement, ie 2259
-\item Probe Response ie 2601
-\item Probe Request ie 2740
+\item Action, ie 4380 : Action Frames are a type of  management frame used to trigger an action in the cell.
+In this case, a request to group ACK's (add block ack request) into frames before sending them.
+\item Request-To-Send packet 2294 : RTS frames are used in the RTS/CTS mechanism and allow stations to request the medium for an amount of time.
+During that time, the station will be able to send uninterrupted and / or without another station becoming active and causing interference.
+\item Clear-To-Send packet 2295 : CTS confirms a RTS request, granting a station the medium for some amount of time.
+\item 802.11 Block Ack 2544 : Acknowledgements of multiple frames, grouped together into one block ack packet.
+\item Acknowledgement, ie 2259 : used to acknowledge the reception of a frame or packet.
+\item Probe Response ie 2601 : Response to a probe request, contains the SSID of the network, information about rates, the kind of network (ad hoc or managed by AP) and the mac address of the AP if there is one, along with various other statistics about the network, in this case about eduroam.
+\item Probe Request ie 2740 : Packet broadcast by a station that wants to know what SSID's are available.
+It is possible for the wanted SSID to be specified in the probe request, in which case we would call it a directed probe request.
+In this case, the request was simply broadcast.
+\item data = icmp ??
+\item neighbour discovery stuff?
+\item VHT NDP Announcement ie 716 : 802.11ac specific, used in the VHT sounding procedure and beamforming.
 \end{itemize}

### 2017-03-02 Lab_3/solutions/L3-5-2.tex
Finished/improvied L3 answers, fixed L3-4 graph. Cleaned up L3 files
Former-commit-id: fc51ba109bff8af870c143cfa139df67e8dd7573

\includegraphics[width=0.5\textwidth]{traces/L3-5-2-tput.pdf}

UDP shows a very clear drop in throughput when increasing the frame length over 1452 bytes. While the first trace (/traces/L3-5-2.udp.1500.pcap) just shows regular UDP datagrams, the following three dumps show IPv6 fragmentation. For example in traces/L3.5.2.udp.1758.pcap, the iperf client first tries to send a UDP packet of length 1772 (packet 1). An ICMPv6 message indicating this packet exceeded the MTU is sent back (packet 2). Each following frame containing 1710 bytes of application data is now divided across two packets (e.g. packet 3+4) each respecting the Path MTU. First an IPv6 fragment is filled with as much data as the path MTU will allow. The remaining data, along with the UDP header is sent in a second packet. This means each 1710 byte frame now requires two sets of layer 2 and 3 headers, significantly increasing the overhead (relative to the amount of application data) and explaining the drop in throughput. We can calculate the relative overhead (assuming Ethernet at layer 2). The 1458 byte frames had 14 (ethernet) + 40 (IPv6) + 8 (UDP) = 62 bytes of headers, meaning ~4.1\% of transmitted data was overhead. For the 1710 byte frames this becomes 2 * 14 (ethernet) + 2 * 48 (IPv6 with fragmentation header) + 8 (UDP) = 132 bytes of headers, leading to ~7.2\% overhead. \\ \\ For further increases in frame length, the throughput starts to rise again. Each 1710 byte frame was divded into parts of 1448 and 262 bytes and each sent in a separate packet. Each packet could contain up to 1452 bytes with this path MTU. For the 1968 and 2226 frame lengths, each frame is still fragmented into only two parts, but the second fragment is bigger (respectively 520 and 778 bytes of application data). By sending fuller packets, the relative overhead decreases and the throughput increases. \\ \\
For TCP the throughput remains stable while increasing the wireless link's MTU. In UDP's case the fragmentation was triggered by also increasing the frame length. For TCP no such option exists. Even if we set the Maximum Segment Size (MSS) to something exceeding the Path MTU, the kernel will first discover the Path MTU and lower the segment size to not exceed the Path MTU. In each of the traces the data is sent to the iperf server in packets of 1514 bytes. Having an MTU higher than the actual packet size somewhere in the data path makes no difference, there is no real difference between the four test cases with TCP.

### 2017-03-02 Lab_3/solutions/L3-2-1.tex
Finished/improvied L3 answers, fixed L3-4 graph. Cleaned up L3 files
Former-commit-id: e2d9a5629b298fdd5f25bbca670651d596d6a449 [formerly a957ecc47cebd3f78a38d769264f36af02aa977c [formerly a957ecc47cebd3f78a38d769264f36af02aa977c [formerly fc51ba109bff8af870c143cfa139df67e8dd7573]]]
Former-commit-id: 9ad44cf59776ac9a2453877d9c225a3462773819
Former-commit-id: 98f291c9e89e1c8647b54ffef7ba3c97147b2df5
Former-commit-id: 00d79017118fbe7a6e9310e9051be078fe6d10fb

@@
 Due to TCP's congestion control there are fewer collisions compared to UDP.
+With lower bitrates the attainable maximum throughput is lower.
+This means TCP's slow start will reach it faster.
+The shorter the slow start phase, the smaller its negative effect on throughput can be.
 For lower bitrates this makes TCP perform better than UDP.
-For higher bitrates the influence of TCP's additional overhead becomes larger and UDP is more performant than TCP again.
+For higher bitrates the influence of TCP's additional overhead and negative effect of congestion control become larger and UDP is more performant than TCP again.
 \\ \\

### 2017-03-02 Lab_4/solutions/L4-2-1.tex

@@
+In the first dump, when the station sent an authentication frame (pkt ID 2666) it was
+followed by an authentication frame (pkt ID 2668) indicating the succesful authentication
+of STA with AP in the WLAN management frame status code.\\
+In the second case, with mac filtering enabled and configured to not allow STA1's wlan0 to connect,
+the authentication frame(pkt ID 464) sent by STA1 is met with an authentication frame(pkt ID 466) sent by the AP
+with a status code 0x0001 signalling an unspecified failure which prevents the establishment of STA's connection with AP.
 in de eerdere kreeg authentication als reply succesful, nu is het unspecified failure

### 2017-03-02 Lab_4/solutions/L4-1-1.tex

@@
-5 broadcast beacon frame
-
-2528 probe 
-2529 probe reply
-2694 association request van sta1 naar ap waar ssid in staat
+\begin{itemize}
+\item PacketID example 5   , Beacon frame broadcast by the AP (or distributed amongst stations in an IBSS network) contains the SSID, in this case, wmn-1-A.
+\item PacketID example 2528, Probe Request, in case of directed probe request contains the SSID of the network the station wants more information about.
+The other case is when no SSID is specified and "Broadcast" is used as SSID, in that case no SSID name is visible in this packet.
+\item PacketID example 2529, Probe Response, response from AP containing capability information, supported data rates, etc., sent after receiving a probe request.
+The SSID is listed under tagged parameters and is equal to wmn-1-A.
+\item PacketID example 2694, Association Request, part of the process where a station registers and connects to an AP.
+This first step contains the SSID with which the STA wants to connect and is sent to the corresponding AP.
+\end{itemize}

### 2017-03-02 Lab_4/solutions/L4-1-2.tex

@@
-278: broadcasted probe request for the ssid wmn-1-A sent by STA1
-279: the answer is from AP to STA1 and contains the ssid as well
-
-in the association request <pkt id> the ssid is still listed. 
-further communication such as association requests also contain the ssid
-
-only broadcasts are now gone such as pkt 5 in the first L411
-
--- only authentication seen
+The SSID itself can still be seen in following packet / frame types, despite SSID hiding:
+\begin{itemize}
+\item PacketID 455  , Probe Request
+\item PacketID 456  , Probe Response
+\item PacketID 617  , Association Request
+\end{itemize}
+The beacon frames sent by our AP can no longer be used to determine the SSID but other traffic still works.
+With the parameter being "ignore_broadcast_ssid", it was to be expected only the beacon frames would no longer contain the SSID.

### 2017-03-02 Lab_4/solutions/L4-4-1.tex
Writing Lab_4 report, finished up to and including WEP. Also limited the large tcpdump file as requested in the assignment.
Former-commit-id: 7fc75470399e852cd4643990a68d792c4b53d991

@@
-Difference : WEP parameters 4 bytes + 4 bytes that dont show up when you look at wep params in the pkt
-IV en key index voor data, WEP ICV na data maar voor fcs
+The difference in size between the decrypted encrypted packet with ID 434 in traces/L4-3-1.STA2.wep.pcap and the regular (non-WEP) packet with ID 353 in traces/L4-3-1.STA2.pcap is 8 bytes.
+The extra 8 bytes that the decrypted encrypted packet has are listed in it's IEEE 802.11 Data header, as the WEP parameters. These WEP parameters include an initialisation vector
+that is 3 bytes in length, a key index for 1 byte and also a WEP ICV that is 4 bytes long and is added after the encrypted data in the packet and before the frame check sequence.

### 2017-03-02 Lab_4/solutions/L4-2-1.tex
Writing Lab_4 report, finished up to and including WEP. Also limited the large tcpdump file as requested in the assignment.
Former-commit-id: 2a11c5b2f9b9fb43ef3661baec594f748e630830 [formerly d546ff20c73c904d4e0664d4c3f9585b6471a25c [formerly 3ee733a48a147ab31686d313d13ef3aabeb27e5f] [formerly 3ee733a48a147ab31686d313d13ef3aabeb27e5f [formerly 7fc75470399e852cd4643990a68d792c4b53d991]]]
Former-commit-id: 321cb65de5c4cba73812455b6519f7c29ef1b513
Former-commit-id: 9cf2e3208c311f8ec1756e80fe0865e862a1800e
Former-commit-id: 2c06f5e18aa0db19e5d3cf09151c70a741420314

--- 2017-02-24 Lab_4/solutions/L4-2-1.tex
@@
-in de eerdere kreeg authentication als reply succesful, nu is het unspecified failure
-<pkt ids>
+In the first dump, when the station sent an authentication frame (pkt ID 2666) it was
+followed by an authentication frame (pkt ID 2668) indicating the succesful authentication
+of STA with AP in the WLAN management frame status code.\\
+In the second case, with mac filtering enabled and configured to not allow STA1's wlan0 to connect,
+the authentication frame(pkt ID 464) sent by STA1 is met with an authentication frame(pkt ID 466iperf) sent by the AP
+with a status code 0x0001 signalling an unspecified failure which prevents the establishment of STA's connection with AP.

### 2017-03-02 Lab_4/solutions/L4-2-2.tex
Writing Lab_4 report, finished up to and including WEP. Also limited the large tcpdump file as requested in the assignment.
Former-commit-id: 2a11c5b2f9b9fb43ef3661baec594f748e630830 [formerly d546ff20c73c904d4e0664d4c3f9585b6471a25c [formerly 3ee733a48a147ab31686d313d13ef3aabeb27e5f] [formerly 3ee733a48a147ab31686d313d13ef3aabeb27e5f [formerly 7fc75470399e852cd4643990a68d792c4b53d991]]]
Former-commit-id: 321cb65de5c4cba73812455b6519f7c29ef1b513
Former-commit-id: 9cf2e3208c311f8ec1756e80fe0865e862a1800e
Former-commit-id: 2c06f5e18aa0db19e5d3cf09151c70a741420314

@@
-Mac address spoof.
-either two mac addresses will be on the network causing problems and crashing the network or the spoofed mac address will not be present at the time granting the attacker access to the network.
+This security measure can be circumvented by spoofing our MAC-address to be among those listed in the accepted MAC addresses. \\ \\In the case that the original device
+as well as the one we're currently spoofed as are present in the network, problems will arise in the network and things will break (the manner in which they break is
+dependent on the type of APs and drivers used.). In the second case, when the original device is no longer present in the network
+(this could be ensured by executing a deauthentication attack on the original device), the attacker will be able to access the network using the spoofed MAC address.

### 2017-03-02 Lab_3/solutions/L3-3-1.tex

@@
 For both TCP and UDP throughput is consistently lower with RTS/CTS enabled.
-The only communication in our tests is direct traffic between STA1 and AP (only STA1 $\rightarrow$ AP in case of UDP).
-There were few collisions even without RTS/CTS.
-This means RTS/CTS mainly adds overhead, leading to lower throughput.
+Looking at the full iperf  output, we noticed UDP packet loss was, surprisingly, consistently higher with RTS/CTS enabled.
+Furthermore the RTS/CTS mechanism adds overhead, leading to lower throughput.
+\\ \\
+Some additional research indicated that packets dropped before they are ever sent, are also counted towards packet loss.
+By analyzing the number of actually sent packets for UDP with tcpdump, we noticed all 'packet loss' was due to packets being dropped at the sender: every packet that was actually sent also arrived.
+Any time spent sending RTS/CTS packets could be considered 'wasted' in this case: there is no packet loss for it to prevent, and the time could have been spent sending additional data.
+Assuming there was no packet loss after transmission for TCP either, this explanation holds for TCP as well.
 \\ \\ Even with RTS/CTS enabled UDP still outperforms TCP: TCP's lower performance is caused by its additional overhead and congestion control, which is not changed by having RTS/CTS enabled.

### 2017-03-02 Lab_3/solutions/L3-4-1.tex

--- 2017-02-24 Lab_3/solutions/L3-4-1.tex
@@
-udp hogere frame length helpt minder, meer botsingen dan verliest er meer dus die capt op een bepaald moment en begint bij 2274 ook weer te dalen
-
-tcp hogere frame length zorgt er voor dat voor constante hoeveelheid overhead er meer data verzonden kan worden dus die blijft in stijgende lijn gaan, ook al is de bereikte tput nog steeds lager dan udp
+\includegraphics[width=0.5\textwidth]{traces/L3-4-1-tput.pdf} \\
+Increasing the MTU and thus the size of the packets leads to increasing throughput for both UDP and TCP.
+Every bump in MTU leads to a throughput increase of 1.7 to 2 Mbps.
+This can partially be explained by the fact that Iperf reports the throughput looking only at the application data.
+Header sizes are 32bytes for 802.11, 40bytes for IPv6, 8bytes for UDP and 20bytes for TCP, leading to a per-packet overhead of 80 and 92 bytes per packet for UDP and TCP respectively.
+The larger the packet size, the lower the relative overhead of the headers.
+In a sense, iperf subtracts the header overhead from the actual packet throughput by only reporting the application data throughput.
+\\ \\ This alone is not enough to explain the increase in throughput though.
+Let's consider UDP for the first two MTUs.
+At 1452 bytes per packet, the headers are ~5.51\% of the application data in size.
+At 1710 bytes per packet this drops to ~4.68\%.
+If we consider the throughput regarding the entire packet, the throughputs rise to 30.9 and 32.7 Mbps respectively.
+The relative gap between them is only slightly smaller than before, and thus still largely unexplained.
+Another interesting observation is that for UDP each increase in MTU decreases the reported packet loss by roughly 4\%.
+By looking at a trace of the tests, it is clear that (almost) all packet loss occurs before the packets are actually sent, as opposed to in the air or at the receiver.
+This seems to indicate the packets are dropped deliberately at send queues.
+A detailed explanation of why the percentage of dropped packets is so significantly different depending on the MTU would require thorough knowledge of how the queues work (both at TCP/UDP and wnic level).
+Some quick Googling indicated more people have noticed this effect, but nobody was able to offer a detailed explanation for why it occurs.
+\\ \\
+(While all other data was gathered during the same session, this exercise was performed at another time. This explains why throughputs in this graph are higher than in the others).

### 2017-03-02 Lab_3/solutions/L3-1-4.tex

@@
 More importantly, TCP's congestion control mechanisms lead to reduced throughput.
+TCP will occasionally not send as fast as the network could handle (mainly during slow start), while UDP is constantly transmitting as fast as the wnic will allow it to.
 Throughput grows together with bitrate: the network supports sending more bits per second with higher bitrates.
@@
 The throughput does not grow as quickly as the bitrate: with higher bitrates the bits are sent faster and bits are more likely to get lost.
-Due to this additional loss doubling the bitrate does not double the observed throughput.
+For TCP this triggers retransmissions and congestion control.
+For UDP the lost data is simply not counted towards the throughput.
+Both have a negative effect on throughput.
+This means doubling the bitrate does not double the observed throughput.
 This translates to a decreasing usage with increasing bitrate.

### 2017-03-02 Lab_2/solutions/L2-6-2.tex
Modified some L2 answers, cleaned up the L2 traces. Apart from 2-5-2 L2 is roughly finished

@@
-Comparing the packets in 1611 (RTS) and 1612 (CTS), we immediately see that the RTS is targetted to the desired receiver while the CTS is broadcast.
-The RTS message is 6 bytes longer than the CTS message and the important differences are in the final field: for RTS it is 802.11 Request To Send and for CTS it is IEEE 802.11 Clear To Send.
-The difference in size is because the L2 address of the transmitter of the CTS packet is not included in the CTS, while it is included for the RTS.
-Aside from that, not much differences can be observed: subtypes and such are different as expected.
+Comparing the packets in 1611 (RTS) and 1612 (CTS), we notice that the RTS contains both a transmitter and a receiver address, while the CTS contains only the latter, making it 6 bytes shorter.
+When the AP receives a RTS, it has to know which station sent it.
+When a station receives a CTS after sending a RTS, it can just assume it was sent from the AP the RTS was sent to.
+The transmitter address is omitted to save 6 bytes.
+Less importantly, the type/subtype field is obviously different.
+Unrelated to this being a RTS and a CTS, the FCS and duration fields differ.

### 2017-03-02 Lab_2/solutions/L2-8-1.tex
Modified some L2 answers, cleaned up the L2 traces. Apart from 2-5-2 L2 is roughly finished

@@
-No, the behaviour is largely the same: we again capture some icmp traffic but it is again not the complete conversation.
-Because the stations are using the 36 channel (albeit with 40 MHz bands) and we are capturing on channel 36, we capture some icmp packets.
+No, the behaviour is largely the same: we again capture some ICMP traffic but it is again not the complete conversation.
+Note that we did manage to capture one ICMP packet (\#262) from STA2 this time, showing that this is not impossible.
+Also interesting, but unrelated to the 40MHz channel is that one request in the second exchange was captured twice (\#1193 and \#1194).
+The second instance indicates that it is a retransmission, meaning the monitor was able to capture the first transmission while the intended receiver was not.
+\\ \\
 %
-The 40 MHz wide bands indicate that the network has a "main" channel, in this case being 36, and extra space to be used if need be.
-Because our icmp traffic is not enough to fill up the main channel, the extra space remains (largely) unused and we thus observe the same behaviour as in the previous exercise where we only capture part of the icmp exchange.
+Even though we now in theory only monitor half of the used channel, we did not capture any fewer ICMP packets.
+This is because when combining two 20MHz channels into one 40MHz channel, one of the two is used as the main channel and the other becomes a secondary channel.
+The secondary channel is only used when needed.
+'5180 HT40+' means 'use 5180 as the main channel and the one above it as secondary'.
+%
+Because our ICMP traffic is by far not enough to saturate the main channel, the secondary channel remains (largely) unused and we thus observe a capture similar to the previous exercise.

### 2017-03-02 Lab_2/solutions/L2-7-1.tex
Modified some L2 answers, cleaned up the L2 traces. Apart from 2-5-2 L2 is roughly finished
Former-commit-id: 147b063b2b4b0cc6af7827f13c3c03686ff0b711

@@
 %
-Not seeing the ping requests of the first ping session could be due to failed demodulations at the monitor, causing it to be not be able to detect the modulated
-messages sent by STA1 in the first ping session. Because the second ping transmission does originate at STA2, it is less of an issue and that ping traffic is captured without issue.
+Not seeing the ping requests of the first ping session could be due to failed demodulations at the monitor.
+Across multiple attemps almost all of STA2's ICMP packets were missed while the other STAs' were almost all captured.
+This could be related to STA1 and STA3 being close to each other and STA2 being farther away.
 %
-The sniffer machine is not in range of the target wireless devices, or there's too much interference, so the sniffer machine's radio
-can't successfully demodulate the transmissions. Note that the increasing use of beamforming helps devices steer their RF energy right
-to where the intended receiver is, which means less of that energy goes to where the sniffer machine is, meaning the sniffer will
-get a worse signal:noise ratio, and less likelihood of demodulating packets sent at the highest data rates.
+Failure to capture/demodulate STA2's packets may be related to the use of beamforming in 802.11n. This helps devices steer their RF energy right
+to where the intended receiver is, and not to the monitor. This means the monitor will
+get a worse signal to noise ratio, making is harder to demodulate packets sent at the highest data rates. We are however not sure whether the devices use/support beamforming.

### 2017-03-02 Lab_2/solutions/L2-5-1.tex
Modified some L2 answers, cleaned up the L2 traces. Apart from 2-5-2 L2 is roughly finished
Former-commit-id: e23d15b4d046db2d1a7efe0d385cd2063bea80e3 [formerly 667ac0bcf380efc65a8adcd7028db2ed790746ce [formerly 667ac0bcf380efc65a8adcd7028db2ed790746ce [formerly 147b063b2b4b0cc6af7827f13c3c03686ff0b711]]]
Former-commit-id: 9335664ea7e24b951bcf88594bcb3acc14c11ada
Former-commit-id: 0880c005726e10a7200c71b6dd62eef2a777fe49
Former-commit-id: 2e53566381fd089a1354c56a8389e71d68277fe5

@@
 For this exercise, we will be comparing packet 2 from traces/L2-4-1.36.pcap and packet 2 from traces/L2-5-1.STA1.pcap.
-For the purpose of comparing the WLAN management frames, I have included them below:
+They are included below for convenience:
 
@@
 %
-for the beacon frame captured in exercise 5:
+For the beacon frame captured in exercise 5:
 \begin{lstlisting}
@@
 %
-A first difference we observed is that the capabilities information beacon frame captured
-in exercise 5 contains different capabilities information than the one captured in exercise 4:
-the ex5 beacon frame shows that the sender of the beacon frame is a STA and that the network is
-an ad-hoc network, while in exercise 4 the capabilities reflect that the transmitter is an AP
-and the network is not ad-hoc.\\
-For the tagged parameters section, the ex5 beacon is bigger. It contains some IBSS parameters (ATIM) and vendor specifici information about the ad-hoc network.
- In the ex4 beacon, the tagged parameters contain information about the channel and the Traffic Indication Map.
+A first difference we observed is in the capabilities information parameters:
+the ad-hoc beacon frame shows that the sender of the beacon frame is a STA and that the network is
+an ad-hoc network, while the AP's capabilities reflect that the transmitter is an AP
+and the network is not in an IBSS (i.e. not ad-hoc).\\
+For the tagged parameters section, the two contain different parameter sets, making the ad-hoc one 4 bytes longer.
+The ad-hoc frame contains an IBSS parameter set with the Announcement Traffic Indication Message (ATIM) window and vendor specificic information parameters concerning Quality of Service.
+ In the AP beacon, the tagged parameter sets contain information about the channel and the Traffic Indication Map.

### 2017-03-03 Lab_4/solutions/L4-2-2.tex
Changed some previous L4 answers, added L4-5 anwers with new trace file, cleaned up L4 traces
Former-commit-id: 5511c468bc01f5a2f070aae601f4fa113eea62a9 [formerly 44bdf5977a2e43760e4e50d86b41a27e37a221e3 [formerly cd8d4c681d2b595cbdffce27a6fdc369899f8747] [formerly cd8d4c681d2b595cbdffce27a6fdc369899f8747 [formerly d095ed18bcc71f5876daaae0cd962c7a5891cf27]]]
Former-commit-id: 0377acbb4cc17645053cd1131ef969e69a56490a
Former-commit-id: 6eda004ed558a2a57893992f193415d08eff9107
Former-commit-id: fcb4274de68a0f9471ff8ca397eafd6473101ed9

--- 2017-02-24 Lab_4/solutions/L4-2-2.tex
@@
-Mac address spoof.
-either two mac addresses will be on the network causing problems and crashing the network or the spoofed mac address will not be present at the time granting the attacker access to the network.
+An attacker can easily circumvent this security measure by spoofing their MAC-address to be among those listed in the accepted MAC addresses.
+The easiest way to find such an address is to monitor the network and obtain MAC addresses of currently connected devices.
+\\ \\In the case that the attacker uses a MAC address for which the genuine device is also connected, problems may arise in the network.
+MAC addresses are supposed to be unique within a LAN.
+The precise issues to arise depend on network configuration and the specific devices used.
+In a wired LAN one of the two devices would probably become inaccissble due to their entry being overwritten in the MAC table of the switch.
+This obviously cannot occur in a wireless LAN.

### 2017-03-03 Lab_4/solutions/L4-1-1.tex
Changed some previous L4 answers, added L4-5 anwers with new trace file, cleaned up L4 traces
Former-commit-id: d095ed18bcc71f5876daaae0cd962c7a5891cf27

--- 2017-02-24 Lab_4/solutions/L4-1-1.tex
@@
-5 broadcast beacon frame
-
-2528 probe 
-2529 probe reply
-2694 association request van sta1 naar ap waar ssid in staat
+\begin{itemize}
+\item Beacon frame (packet 5): broadcast by the AP (or distributed amongst stations in an IBSS network) contains the SSID, in this case, wmn-1-A.
+\item Probe Request (packet 2528): the directed probe request contains the SSID of the network the station wants more information about.
+There are also broadcast probe requests which do not contain an SSID.
+\item Probe Response (packet 2529): response from AP containing capability information, supported data rates, etc., sent after receiving a probe request. The SSID is listed under tagged parameters.
+\item Association Request (packet 2694): part of the process where a station registers and connects to an AP.
+This first step contains the SSID with which the STA wants to connect and is sent to the corresponding AP.
+\end{itemize}

### 2017-03-03 Lab_4/solutions/L4-2-1.tex
Changed some previous L4 answers, added L4-5 anwers with new trace file, cleaned up L4 traces
Former-commit-id: d095ed18bcc71f5876daaae0cd962c7a5891cf27

--- 2017-02-24 Lab_4/solutions/L4-2-1.tex
@@
-in de eerdere kreeg authentication als reply succesful, nu is het unspecified failure
-<pkt ids>
+In the first dump, when the station sent an authentication frame (packet 2666) the AP replied with an authentication frame (packet 2668) indicating the station was succesfully authenticated (status code 0: successful).\\
+In the second case, with MAC filtering enabled and configured to not allow STA1's wlan0 to connect,
+the authentication frame(packet 464) sent by STA1 is met with an authentication frame(packet 466)
+with status code 1.
+This indicates an unspecified failure.
+STA1 does not attempt an association as authentication failed.
+Note that STA1 is not informed that authentication failed because it was blacklisted.
+There is no reason to let a potential attacker know why exactly they could not connect.

### 2017-03-03 Lab_4/solutions/L4-5-3.tex
Changed some previous L4 answers, added L4-5 anwers with new trace file, cleaned up L4 traces
Former-commit-id: d095ed18bcc71f5876daaae0cd962c7a5891cf27

There are no significant differences between the authentication phases (packets 409 and 411 for WEP, 176 and 178 for WPA2). For the assocation phase (packets 424 and and 436 for WEP, 189 and 191 for WPA2), an additional 22 byte RSN (Robust Security Network) Information tag is present in WPA2's Request. This identifies the requested protocols. Other than this there are no significant differences. \\ \\The main difference is that this is followed with 4 EAPOL (Extensible Authentication Protocol over LAN) packets which perform the actual WPA2 authentication. This is known as EAPOL's 4-way handshake. Station and AP exchange some randomly generated values which are used in combination with the passphrase to generate the Pairwise Transient Key (PTK) they will use for encryption. Someone overhearing this cannot recreate the PTK unless they also know the passphrase. A step like this is completely absent in WEP (with Open System authentication): the AP assumes the station knows the passphrase, knowing the station will not be able to encrypt/decrypt communication without it.

### 2017-03-03 Lab_4/solutions/L4-5-4.tex
Changed some previous L4 answers, added L4-5 anwers with new trace file, cleaned up L4 traces
Former-commit-id: d095ed18bcc71f5876daaae0cd962c7a5891cf27

We compare two encrypted Router Solicitation packets (packet 596 in L4-3-1.STA2.wep.pcap, packet 256 in L4-5-1.STA2.pcap). Looking at the encrypted packets, we notice that while the WEP packet has 8 bytes of WEP parameters, the WPA2 packet has 8 bytes of CCMP parameters. CCMP is the default encryption method in WPA2. This consists of an 8 byte extended IV. All 8 bytes are at the end of the 802.11 header, as opposed to the WEP parameters where 4 bytes were between the encrypted data and the FCS. We furthermore notice the encrypted data is 8 bytes larger for WPA2. After decrypting the WPA2 packets (wpa-pwd aRandomPassphrase:wmn-1-A) the 8 additional bytes are gone. We suspect these 8 bytes were a side-effect (e.g. padding) of the encryption algorithm.

### 2017-03-03 Lab_4/solutions/L4-1-2.tex
Writing Lab_4 report, finished up to and including WEP. Also limited the large tcpdump file as requested in the assignment.
Changed some previous L4 answers, added L4-5 anwers with new trace file, cleaned up L4 traces

@@
 \begin{itemize}
-\item PacketID 455  , Probe Request
-\item PacketID 456  , Probe Response
-\item PacketID 617  , Association Request
+\item Probe Request (packet 455)
+\item Probe Response (packet 456)
+\item Association Request (packet 617)
 \end{itemize}
-The beacon frames sent by our AP can no longer be used to determine the SSID but other traffic still works.
-With the parameter being "ignore_broadcast_ssid", it was to be expected only the beacon frames would no longer contain the SSID.
+The SSID was only removed from the broadcast beacon frames, the others are unchanged.
+Note that the beacon frames are still sent.
+In Wireshark's packet info field the SSID is shown as "Broadcast" but further inspection of the packet shows the SSID field is simply empty.
+With the parameter being "ignore\_broadcast\_ssid", it was to be expected only the beacon frames would no longer contain the SSID.

### 2017-03-03 Lab_4/solutions/L4-4-1.tex
Changed some previous L4 answers, added L4-5 anwers with new trace file, cleaned up L4 traces

--- 2017-03-01 Lab_4/solutions/L4-4-1.tex
@@
-Difference : WEP parameters 4 bytes + 4 bytes that dont show up when you look at wep params in the pkt
-IV en key index voor data, WEP ICV na data maar voor fcs
+The difference in size between the encrypted packet with ID 1415 in traces/L4-3-1.STA2.wep.pcap and the regular (non-WEP) packet with ID 1077 in traces/L4-3-1.STA2.pcap is 8 bytes (154 vs 146).
+The extra 8 bytes are the WEP parameters in the IEEE 802.11 Data header.
+These WEP parameters are an initialization vector (3 bytes), a key index (1 byte) and a WEP Integrity Check Value (4 bytes).
+This ICV is inserted between the encrypted data and the frame check sequence.

### 2017-03-03 Lab_4/solutions/L4-6-1.tex
report on DoS (after finally understanding the whole spoof-everything concept) completed. Wrote report using current L4-6-1.pcap, the result of the second experiment and stored old result in L4-6-1.old_poging1.pcap.

The deassociation attack caused the ping session to be interrupted. After 5 seconds, the station was able to reassociate with the AP and resume the ping traffic.
Repeating the attack some time later again caused a 5 second interruption in ping traffic. \\ \\A second attempt at the exercise showed a 22 second delay and not a 5 second delay.
%
We found that reestablishment of the connection is a function of how aggressively the client attempts to reauthenticate and any higher-level timeouts or backoffs and varying network quality
that may suppress the demand for communication and conclude that that is what leads to the difference in the time needed to be able to resume the ping session.\\ \\
% %source: http://users.csc.calpoly.edu/~bellardo/pubs/usenix-sec03-80211dos-html/aio.html -- deauthentication header
The main thing to note for the ping session is that it was interrupted for a short time but when the attack ceased, the ping session resumed after the STA reassociated with the AP.

### 2017-03-03 Lab_4/solutions/L4-6-2.tex
report on DoS (after finally understanding the whole spoof-everything concept) completed. Wrote report using current L4-6-1.pcap, the result of the second experiment and stored old result in L4-6-1.old_poging1.pcap.
Former-commit-id: da603afe73d91ae825aeca6e459d3daa4ad6c60e [formerly 48cb0ef5c04b73297cdb92835e6d67f2dba342d5 [formerly 48cb0ef5c04b73297cdb92835e6d67f2dba342d5 [formerly 4d011b2488b0b55e97cbc75631b3b69c1f87dcb9]]]
Former-commit-id: 609929a03727fa35fb17407ac17095e00a8d61c8
Former-commit-id: d04614249d1112874b046a31e4e923f05396f4cb
Former-commit-id: c6faf457912e017d2c7b6cbf0d053daee

After seeing the first set of ping messages (though, we did not manage to decrypt them all using Wireshark), we see 128 de-authentication messages in the dump file (traces/L4-6-1.STA1.pcap).
Out of those 128 packets, 64 are sent to the AP by what appears to be STA, but it is actually the attacker that is spoofing it's MAC address to be equal to STA1's wlan1 interface and the
remaining 64 are sent to the STA1 by what appears to be the AP, but it is again the attacker spoofing it's MAC address to be equal to AP's.\\ \\
%
In our dump file (/traces/L4-6-1.STA1.pcap) we can see a reassociation that starts at packetID 502 (after a first attack, but the packets before the first attack are less clear
because we could not decrypt the first ping session.) that is followed by a few pings exchanges with the last reply arriving in packetID 765. After this packet 765 we again started
a de-association attack, which starts with the packet with ID 791 that is sent by the attacker (spoofing as the AP) to STA1. It demonstrates the attack because STA1 stops sending traffic to AP after
receiving this deauthentication message, thinking AP no longer wants to receive any traffic from STA1. After this first de-authentication packet, a second de-authentication packet is sent by
the attacker (spoofing as STA1) to the AP, making it so neither side suspects anything, and will simply try to re-associate. This sending of de-authentication packets conttinues
with packets being sent by the attacker to STA1 and AP (still spoofing as AP for STA1 and as STA1 for AP) until 64 packets have been sent in both directions (thus, 128 total).\\ \\
%
The continued sending of de-authentication packets makes it so that the STA1 cannot re-associate with the AP. After the 128 packets the de-association attack stops and in packet 1104 we see
a new association start up between STA1 and AP which succeeds, as the attack has ceased. After this association succeeds, pings resume as normal and we see that the amount of time that our
attack interrupted the pings for was roughly equal to 5 seconds.

### 2017-03-05 Lab_4/solutions/L4-6-1.tex
Modified L4-6 solutions

@@
-The deassociation attack caused the ping session to be interrupted. After 5 seconds, the station was able to reassociate with the AP and resume the ping traffic.
-Repeating the attack some time later again caused a 5 second interruption in ping traffic. \\ \\A second attempt at the exercise showed a 22 second delay and not a 5 second delay.
-%
-We found that reestablishment of the connection is a function of how aggressively the client attempts to reauthenticate and any higher-level timeouts or backoffs and varying network quality
-that may suppress the demand for communication and conclude that that is what leads to the difference in the time needed to be able to resume the ping session.\\ \\
-% %source: http://users.csc.calpoly.edu/~bellardo/pubs/usenix-sec03-80211dos-html/aio.html -- deauthentication header
-The main thing to note for the ping session is that it was interrupted for a short time but when the attack ceased, the ping session resumed after the STA reassociated with the AP.
+The disassociation/deauthentication attack caused the ping session to be interrupted. After 5 seconds, the station was able to reestablish connection with the AP and resume the ping traffic.
+Repeating the attack some time later again caused a  5 second interruption in ping traffic.  \\ \\
+The main thing to note for the ping session is that it was interrupted for a short time but when the attack ceased, the ping session resumed quickly after the STA reassociated with the AP.

### 2017-03-05 Lab_4/solutions/L4-6-2.tex
Modified L4-6 solutions

@@
-After seeing the first set of ping messages (though, we did not manage to decrypt them all using Wireshark), we see 128 de-authentication messages in the dump file (traces/L4-6-1.STA1.pcap).
-Out of those 128 packets, 64 are sent to the AP by what appears to be STA, but it is actually the attacker that is spoofing it's MAC address to be equal to STA1's wlan1 interface and the
-remaining 64 are sent to the STA1 by what appears to be the AP, but it is again the attacker spoofing it's MAC address to be equal to AP's.\\ \\
+Wireshark could not decrypt the earliest ping packets.
+It is possible that the multiple EAPOL handshakes between STA1 and AP due to the attack confused Wirehsark.
+The others can again be decrypted with wpa-pwd aRandomPassphrase:wmn-
+1-A \\ \\
+
+After seeing the first set of ping messages (not decrypted by Wireshark), we see 128 deauthentication messages in the dump file (traces/L4-6-1.STA1.pcap).
+Out of those 128 packets, 64 are sent to the STA by what appears to be AP (e.g.
+203), and the other 64 are sent in the opposite direction (e.g.
+205).
+Both of these are actually sent by the attacker, who is spoofing the source MAC address to be equal to STA1's or AP's wlan1 interface.
+These are the MAC addresses we added to the command.
+By sending deauth packets to both STA and AP, both are aware of the deauthentication, preparing both for the following reauthentication.
+As the purpose of a deauthentication attack is often to capture the following handshake, this is useful.
+By sending many deauth packets, the attacker increases the likelihood of at least one being received even when reception is poor.
+The attack works as soon as one is received.
+One pair of deauthorization packets is sent every 50ms.
+There is a 20ms interval between the AP to STA and the STA to AP packet.
+This means the attacker sends spoofed deauthorizations for about 3.15 seconds.
+\\ \\
 %
-In our dump file (/traces/L4-6-1.STA1.pcap) we can see a reassociation that starts at packetID 502 (after a first attack, but the packets before the first attack are less clear
-because we could not decrypt the first ping session.) that is followed by a few pings exchanges with the last reply arriving in packetID 765. After this packet 765 we again started
-a de-association attack, which starts with the packet with ID 791 that is sent by the attacker (spoofing as the AP) to STA1. It demonstrates the attack because STA1 stops sending traffic to AP after
-receiving this deauthentication message, thinking AP no longer wants to receive any traffic from STA1. After this first de-authentication packet, a second de-authentication packet is sent by
-the attacker (spoofing as STA1) to the AP, making it so neither side suspects anything, and will simply try to re-associate. This sending of de-authentication packets conttinues
-with packets being sent by the attacker to STA1 and AP (still spoofing as AP for STA1 and as STA1 for AP) until 64 packets have been sent in both directions (thus, 128 total).\\ \\
+We see the STA ceases the pings as soon as a deauthentication is received, as it thinks the AP no longer wants to receive any traffic from it.
+A reauthentication (packet 491) + reassociation (packet 502) start roughly 0.4 seconds after the attack ceased.
+Once the full authentication + assocation process is completed, pings resume at packet 522.
+Afterwards we started another two deauthentication attacks with the same effect.
 %
-The continued sending of de-authentication packets makes it so that the STA1 cannot re-associate with the AP. After the 128 packets the de-association attack stops and in packet 1104 we see
-a new association start up between STA1 and AP which succeeds, as the attack has ceased. After this association succeeds, pings resume as normal and we see that the amount of time that our
-attack interrupted the pings for was roughly equal to 5 seconds.

### 2017-03-05 Lab_4/solutions/L4-2-1.tex
Review of Lab_4, added refs to tracefiles where I saw they were missing. Fixed some typo's.
Former-commit-id: ff6b7c514b952aebc4a289413d0b6ba65828b4d6 [formerly 51435a56835d0eeb7babc90f9f0c25085709dd68 [formerly 51435a56835d0eeb7babc90f9f0c25085709dd68 [formerly 180a98254dcf255543ff72f101deac34a2312d85]]]
Former-commit-id: 69524c6d84a28ded87660ec5ba33aa32f5569b11
Former-commit-id: 06de46f32825fa1527f5d38c73593a2ba5a6a55f
Former-commit-id: b4ef786d70e62b66a72c3809b3ab492ea32a8df7

@@
-In the first dump, when the station sent an authentication frame (packet 2666) the AP replied with an authentication frame (packet 2668) indicating the station was succesfully authenticated (status code 0: successful).\\
-In the second case, with MAC filtering enabled and configured to not allow STA1's wlan0 to connect,
+In the first dump (traces/L4-1-1.STA2.pcap), when the station sent an authentication frame (packet 2666) the AP replied with an authentication frame (packet 2668) indicating the station was succesfully authenticated (status code 0: successful).\\
+In the second case (found in traces/L4-2-1.STA2.pcap), with MAC filtering enabled and configured to not allow STA1's wlan0 to connect,
 the authentication frame(packet 464) sent by STA1 is met with an authentication frame(packet 466)

### 2017-03-05 Lab_2/solutions/L2-7-1.tex
Unplagiarized beamforming explanation, removed some unneeded traces from L2

@@
 %
-Failure to capture/demodulate STA2's packets may be related to the use of beamforming in 802.11n. This helps devices steer their RF energy right
-to where the intended receiver is, and not to the monitor. This means the monitor will
-get a worse signal to noise ratio, making is harder to demodulate packets sent at the highest data rates. We are however not sure whether the devices use/support beamforming.
+Failure to capture/demodulate STA2's packets may be related to the use of beamforming in 802.11n.
+With beamforming the device can 'aim' its transmission at the intended receiver.
+When applied here, the monitor would receive a weaker signal, making is harder to demodulate packets, especially when combined with a high data rate.
+We are however not sure whether the devices use/support beamforming.

### 2017-03-05 Lab_4/solutions/L4-5-3.tex

@@
-There are no significant differences between the authentication phases (packets 409 and 411 for WEP, 176 and 178 for WPA2).
+There are no significant differences between the authentication phases, using /traces/L4-3-1.STA2.wep.pcap for WEP and /traces/L4-5-1.STA2.pcap for WPA2 (packets 409 and 411 for WEP, 176 and 178 for WPA2).
 For the assocation phase (packets 424 and and 436 for WEP, 189 and 191 for WPA2), an additional 22 byte RSN (Robust Security Network) Information tag is present in WPA2's Request.

### 2017-03-05 Lab_4/solutions/L4-1-2.tex
Review of Lab_4, added refs to tracefiles where I saw they were missing. Fixed some typo's.
Former-commit-id: 180a98254dcf255543ff72f101deac34a2312d85
Small changes to L4
Former-commit-id: 85572e302b2054e9df8addb61e1b05a242c19184

@@
-The SSID itself can still be seen in following packet / frame types, despite SSID hiding:
+The SSID itself can still be seen in following packet / frame types in our dump file traces/L4-1-2.STA2.pcap, despite SSID hiding:
 \begin{itemize}
@@
 The SSID was only removed from the broadcast beacon frames, the others are unchanged.
-Note that the beacon frames are still sent.
+Note that the beacon frames are still sent (e.g.
+packet 1).
 In Wireshark's packet info field the SSID is shown as "Broadcast" but further inspection of the packet shows the SSID field is simply empty.

### 2017-03-05 Lab_4/solutions/L4-2-1.tex
Small changes to L4
Former-commit-id: 51d0f2b3db5e9158fedd338386d5e2e5972869e0 [formerly 2b5f1f1c0fb6825ae69dc645d22b140faa770073 [formerly 2b5f1f1c0fb6825ae69dc645d22b140faa770073 [formerly 85572e302b2054e9df8addb61e1b05a242c19184]]]
Former-commit-id: ecd77de1f9afec5cae939969c855182e2b9d2389
Former-commit-id: 52891089232b5d215ced4a8b3ee8043519173568
Former-commit-id: 21c20f630ade97911bbb9fd85a85221c0011d317

--- 2017-03-03 Lab_4/solutions/L4-2-1.tex
@@
-In the first dump, when the station sent an authentication frame (packet 2666) the AP replied with an authentication frame (packet 2668) indicating the station was succesfully authenticated (status code 0: successful).\\
-In the second case, with MAC filtering enabled and configured to not allow STA1's wlan0 to connect,
-the authentication frame(packet 464) sent by STA1 is met with an authentication frame(packet 466)
+In the first dump (traces/L4-1-1.STA2.pcap), when the station sent an authentication frame (packet 2666) the AP replied with an authentication frame (packet 2668) indicating the station was succesfully authenticated (status code 0: successful).\\
+In the second case (found in traces/L4-2-1.STA2.pcap), with MAC filtering enabled and configured to not allow STA1's wlan0 to connect,
+the authentication frame (packet 464) sent by STA1 is met with an authentication frame (packet 466)
 with status code 1.

### 2017-03-05 Lab_4/solutions/L4-6-2.tex
Review of Lab_4, added refs to tracefiles where I saw they were missing. Fixed some typo's.
Former-commit-id: ff6b7c514b952aebc4a289413d0b6ba65828b4d6 [formerly 51435a56835d0eeb7babc90f9f0c25085709dd68 [formerly 51435a56835d0eeb7babc90f9f0c25085709dd68 [formerly 180a98254dcf255543ff72f101deac34a2312d85]]]
Former-commit-id: 69524c6d84a28ded87660ec5ba33aa32f5569b11
Former-commit-id: 06de46f32825fa1527f5d38c73593a2ba5a6a55f
Former-commit-id: b4ef786d70e62b66a72c3809b3ab492ea32a8df7
Small changes to L4

@@
 Wireshark could not decrypt the earliest ping packets.
-It is possible that the multiple EAPOL handshakes between STA1 and AP due to the attack confused Wirehsark.
+It is possible that the multiple EAPOL handshakes between STA1 and AP due to the attack confused Wireshark.
 The others can again be decrypted with wpa-pwd aRandomPassphrase:wmn-
@@
 %
-We see the STA ceases the pings as soon as a deauthentication is received, as it thinks the AP no longer wants to receive any traffic from it.
-A reauthentication (packet 491) + reassociation (packet 502) start roughly 0.4 seconds after the attack ceased.
+We see the STA ceases the pings as soon as a deauthentication is received, as it believes the connection to be terminated.
+A reauthentication (packet 491) and reassociation (packet 502) start roughly 0.4 seconds after the attack ceased.
 Once the full authentication + assocation process is completed, pings resume at packet 522.

### 2017-03-15 Lab_2/solutions/L2-5-3.tex
Added new answers for 2-5-2/3, tgz'd lab1
Former-commit-id: 8e36ca0a9c506066ff0b686ecc2d22658683382d

--- 2017-02-27 Lab_2/solutions/L2-5-3.tex
@@
-The purpose of each of these frames:
-\begin{itemize}
-\item Request-To-Send packet
-\item Clear-To-Send packet
-\item 802.11 Block Ack
-\item Acknowledgement, ie 2259
-\item Probe Response ie 2601
-\item Probe Request ie 2740
-\end{itemize}
+The purpose of the Acknowledgement is to let the sender of an IEEE 802.11 Data frame know it was received properly and no retransmission will be needed.
+For example in the packets mentioned above, the Acknowledgement's receiver address is the Data frame's transmitter address: the receiver of the Data frame lets the sender know it was received.
+Note that there is no transmitter address in the Acknowledgements.
+This is related to interframe spaces in IEEE 802.11.
+There is a well-defined window within which the Acknowledgement for a specific frame must be sent, and it ends before a new data frame can be sent.
+This means a valid Acknowledgement must always be for the latest Data frame sent: no further identification is necessary.
+%The purpose of each of these frames:
+%\begin{itemize}
+%\item Action, ie 4380 : Action Frames are a type of  management frame used to trigger an action in the cell.
+In this case, a request to group ACK's (add block ack request) into frames before sending them.
+%\item Request-To-Send packet 2294 : RTS frames are used in the RTS/CTS mechanism and allow stations to request the medium for an amount of time.
+During that time, the station will be able to send uninterrupted and / or without another station becoming active and causing interference.
+%\item Clear-To-Send packet 2295 : CTS confirms a RTS request, granting a station the medium for some amount of time.
+%\item 802.11 Block Ack 2544 : Acknowledgements of multiple frames, grouped together into one block ack packet.
+%\item Acknowledgement, ie 2259 : used to acknowledge the reception of a frame or packet.
+%\item Probe Response ie 2601 : Response to a probe request, contains the SSID of the network, information about rates, the kind of network (ad hoc or managed by AP) and the mac address of the AP if there is one, along with various other statistics about the network, in this case about eduroam.
+%\item Probe Request ie 2740 : Packet broadcast by a station that wants to know what SSID's are available.
+It is possible for the wanted SSID to be specified in the probe request, in which case we would call it a directed probe request.
+In this case, the request was simply broadcast.
+%\item data = icmp ??
+%\item neighbour discovery stuff?
+%\item VHT NDP Announcement ie 716 : 802.11ac specific, used in the VHT sounding procedure and beamforming.
+%\end{itemize}

### 2017-03-15 Lab_1/solutions/L1-5-1.tex
Finished Lab1 answers. Added .gitignore for latex files
Final changes to lab1 report, updated zip file to reflect changes.

@@
+Looking at trace file traces/L1-5-1.AP.pcap, we observe the following message types:
 \begin{itemize}
-
-\item Neighbor Solicitation:  STA1 is trying to find STA2 by broadcasting these
-\item Neighbor Advertisement: STA2 telling STA1 it's L2 address
-\item Ping Request from STA1 to STA2
-\item Ping Reply from STA2 to STA1
-%\item More Neighbor Solicitations and Advertisements follow, refreshing the previous bindings. Interestingly, these have source and destination different.
+\item Neighbor Solicitation (ie packet nr 1):  STA1 is trying to find STA2 by broadcasting these
+\item Neighbor Advertisement (ie packet nr 3): STA2 telling STA1 its L2 address
+\item Ping Request (ie packet nr 4) from STA1 to STA2
+\item Ping Reply (ie packet nr 5) from STA2 to STA1
+%\item More Neighbor Solicitations and Advertisements follow, refreshing the previous bindings. Interestingly, these have different source and destination addresses.
 \end{itemize}

### 2017-03-15 Lab_1/solutions/L1-5-1.tex

@@
 \begin{itemize}
-\item Neighbor Solicitation (ie packet nr 1):  STA1 is trying to find STA2 by broadcasting these
-\item Neighbor Advertisement (ie packet nr 3): STA2 telling STA1 its L2 address
-\item Ping Request (ie packet nr 4) from STA1 to STA2
-\item Ping Reply (ie packet nr 5) from STA2 to STA1
+\item Neighbor Solicitation (e.g. packet 1):  STA1 is trying to find STA2 by broadcasting these
+\item Neighbor Advertisement (e.g. packet 3): STA2 telling STA1 its L2 address
+\item Ping Request (e.g. packet 4) from STA1 to STA2
+\item Ping Reply (e.g. packet 5) from STA2 to STA1
 %\item More Neighbor Solicitations and Advertisements follow, refreshing the previous bindings. Interestingly, these have different source and destination addresses.

### 2017-05-01 advnetwpaper/paper.tex
Added initial version of paper including first drafts of introduction and measurements

%% bare_conf.tex
%% V1.4b
%% 2015/08/26
%% by Michael Shell
%% See:
%% http://www.michaelshell.org/
%% for current contact information.
%%
%% This is a skeleton file demonstrating the use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.8b or later) with an IEEE
%% conference paper.
%%
%% Support sites:
%% http://www.michaelshell.org/tex/ieeetran/
%% http://www.ctan.org/pkg/ieeetran
%% and
%% http://www.ieee.org/

%%*************************************************************************
%% Legal Notice:
%% This code is offered as-is without any warranty either expressed or
%% implied; without even the implied warranty of MERCHANTABILITY or
%% FITNESS FOR A PARTICULAR PURPOSE!
%% User assumes all risk.
%% In no event shall the IEEE or any contributor to this code be liable for
%% any damages or losses, including, but not limited to, incidental,
%% consequential, or any other damages, resulting from the use or misuse
%% of any information contained here.
%%
%% All comments are the opinions of their respective authors and are not
%% necessarily endorsed by the IEEE.
%%
%% This work is distributed under the LaTeX Project Public License (LPPL)
%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
%% distributed and modified. A copy of the LPPL, version 1.3, is included
%% in the base LaTeX documentation of all distributions of LaTeX released
%% 2003/12/01 or later.
%% Retain all contribution notices and credits.
%% ** Modified files should be clearly indicated as such, including  **
%% ** renaming them and changing author support contact information. **
%%*************************************************************************

% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
% *** with production work. The IEEE's font choices and paper sizes can   ***
% *** trigger bugs that do not appear when using other class files.       ***                          ***
% The testflow support page is at:
% http://www.michaelshell.org/tex/testflow/

\documentclass[conference]{IEEEtran}
% Some Computer Society conferences also require the compsoc mode option,
% but others use the standard conference format.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[conference]{../sty/IEEEtran}

% Some very useful LaTeX packages include:
% (uncomment the ones you want to load)

% *** MISC UTILITY PACKAGES ***
%
%\usepackage{ifpdf}
% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
% compilation based on whether the output is pdf or dvi.
% usage:
% \ifpdf
%   % pdf code
% \else
%   % dvi code
% \fi
% The latest version of ifpdf.sty can be obtained from:
% http://www.ctan.org/pkg/ifpdf
% Also, note that IEEEtran.cls V1.7 and later provides a builtin
% \ifCLASSINFOpdf conditional that works the same way.
% When switching from latex to pdflatex and vice-versa, the compiler may
% have to be run twice to clear warning/error messages.

% *** CITATION PACKAGES ***
%
\usepackage{cite}
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of the IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off
% such as if a citation ever needs to be enclosed in parenthesis.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 5.0 (2009-03-20) and later if using hyperref.sty.
% The latest version can be obtained at:
% http://www.ctan.org/pkg/cite
% The documentation is contained in the cite.sty file itself.

\usepackage{siunitx}
\usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
\graphicspath{{../images/}}

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  % \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation
% can be obtained at:
% http://www.ctan.org/pkg/graphicx
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found at:
% http://www.ctan.org/pkg/epslatex
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). The IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.
%
% You can find documentation about the pdfTeX application at:
% http://www.tug.org/applications/pdftex

% *** MATH PACKAGES ***
%
%\usepackage{amsmath}
% A popular package from the American Mathematical Society that provides
% many useful and powerful commands for dealing with mathematics.
%
% Note that the amsmath package sets \interdisplaylinepenalty to 10000
% thus preventing page breaks from occurring within multiline equations. Use:
%\interdisplaylinepenalty=2500
% after loading amsmath to restore such page breaks as IEEEtran.cls normally
% does. amsmath.sty is already installed on most LaTeX systems. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/pkg/amsmath

% *** SPECIALIZED LIST PACKAGES ***
%
%\usepackage{algorithmic}
% algorithmic.sty was written by Peter Williams and Rogerio Brito.
% This package provides an algorithmic environment fo describing algorithms.
% You can use the algorithmic environment in-text or within a figure
% environment to provide for a floating algorithm. Do NOT use the algorithm
% floating environment provided by algorithm.sty (by the same authors) or
% algorithm2e.sty (by Christophe Fiorio) as the IEEE does not use dedicated
% algorithm float types and packages that provide these will not provide
% correct IEEE style captions. The latest version and documentation of
% algorithmic.sty can be obtained at:
% http://www.ctan.org/pkg/algorithms
% Also of interest may be the (relatively newer and more customizable)
% algorithmicx.sty package by Szasz Janos:
% http://www.ctan.org/pkg/algorithmicx

% *** ALIGNMENT PACKAGES ***
%
%\usepackage{array}
% Frank Mittelbach's and David Carlisle's array.sty patches and improves
% the standard LaTeX2e array and tabular environments to provide better
% appearance and additional user controls. As the default LaTeX2e table
% generation code is lacking to the point of almost being broken with
% respect to the quality of the end results, all users are strongly
% advised to use an enhanced (at the very least that provided by array.sty)
% set of table tools. array.sty is already installed on most systems. The
% latest version and documentation can be obtained at:
% http://www.ctan.org/pkg/array

% IEEEtran contains the IEEEeqnarray family of commands that can be used to
% generate multiline equations as well as matrices, tables, etc., of high
% quality.

% *** SUBFIGURE PACKAGES ***
%\ifCLASSOPTIONcompsoc
%  \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
%\else
%  \usepackage[caption=false,font=footnotesize]{subfig}
%\fi
% subfig.sty, written by Steven Douglas Cochran, is the modern replacement
% for subfigure.sty, the latter of which is no longer maintained and is
% incompatible with some LaTeX packages including fixltx2e. However,
% subfig.sty requires and automatically loads Axel Sommerfeldt's caption.sty
% which will override IEEEtran.cls' handling of captions and this will result
% in non-IEEE style figure/table captions. To prevent this problem, be sure
% and invoke subfig.sty's "caption=false" package option (available since
% subfig.sty version 1.3, 2005/06/28) as this is will preserve IEEEtran.cls
% handling of captions.
% Note that the Computer Society format requires a larger sans serif font
% than the serif footnote size font used in traditional IEEE formatting
% and thus the need to invoke different subfig.sty package options depending
% on whether compsoc mode has been enabled.
%
% The latest version and documentation of subfig.sty can be obtained at:
% http://www.ctan.org/pkg/subfig

% *** FLOAT PACKAGES ***
%
%\usepackage{fixltx2e}
% fixltx2e, the successor to the earlier fix2col.sty, was written by
% Frank Mittelbach and David Carlisle. This package corrects a few problems
% in the LaTeX2e kernel, the most notable of which is that in current
% LaTeX2e releases, the ordering of single and double column floats is not
% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
% single column figure to be placed prior to an earlier double column
% figure.
% Be aware that LaTeX2e kernels dated 2015 and later have fixltx2e.sty's
% corrections already built into the system in which case a warning will
% be issued if an attempt is made to load fixltx2e.sty as it is no longer
% needed.
% The latest version and documentation can be found at:
% http://www.ctan.org/pkg/fixltx2e

%\usepackage{stfloats}
% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
% the ability to do double column floats at the bottom of the page as well
% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
% LaTeX2e). It also provides a command:
%\fnbelowfloat
% to enable the placement of footnotes below bottom floats (the standard
% LaTeX2e kernel puts them above bottom floats). This is an invasive package
% which rewrites many portions of the LaTeX2e float routines. It may not work
% with other packages that modify the LaTeX2e float routines. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/pkg/stfloats
% Do not use the stfloats baselinefloat ability as the IEEE does not allow
% \baselineskip to stretch. Authors submitting work to the IEEE should note
% that the IEEE rarely uses double column equations and that authors should try
% to avoid such use. Do not be tempted to use the cuted.sty or midfloat.sty
% packages (also by Sigitas Tolusis) as the IEEE does not format its papers in
% such ways.
% Do not attempt to use stfloats with fixltx2e as they are incompatible.
% Instead, use Morten Hogholm'a dblfloatfix which combines the features
% of both fixltx2e and stfloats:
%
% \usepackage{dblfloatfix}
% The latest version can be found at:
% http://www.ctan.org/pkg/dblfloatfix

% *** PDF, URL AND HYPERLINK PACKAGES ***
%
%\usepackage{url}
% url.sty was written by Donald Arseneau. It provides better support for
% handling and breaking URLs. url.sty is already installed on most LaTeX
% systems. The latest version and documentation can be obtained at:
% http://www.ctan.org/pkg/url
% Basically, \url{my_url_here}.

% *** Do not adjust lengths that control margins, column widths, etc. ***
% *** Do not use packages that alter fonts (such as pslatex).         ***
% There should be no need to do such things with IEEEtran.cls V1.6 and later.
% (Unless specifically asked to do so by the journal or conference you plan
% to submit to, of course. )

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{Bare Demo of IEEEtran.cls\\ for IEEE Conferences}

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Michael Shell}
\IEEEauthorblockA{School of Electrical and\\Computer Engineering\\
Georgia Institute of Technology\\
Atlanta, Georgia 30332--0250\\
Email: http://www.michaelshell.org/contact.html}
\and
\IEEEauthorblockN{Homer Simpson}
\IEEEauthorblockA{Twentieth Century Fox\\
Springfield, USA\\
Email: homer@thesimpsons.com}
\and
\IEEEauthorblockN{James Kirk\\ and Montgomery Scott}
\IEEEauthorblockA{Starfleet Academy\\
San Francisco, California 96678--2391\\
Telephone: (800) 555--1212\\
Fax: (888) 555--1212}}

% conference papers do not typically use \thanks and this command
% is locked out in conference mode. If really needed, such as for
% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
% after \documentclass

% for over three affiliations, or if they all won't fit within the width
% of the page, use this alternative format:
%
%\author{\IEEEauthorblockN{Michael Shell\IEEEauthorrefmark{1},
%Homer Simpson\IEEEauthorrefmark{2},
%James Kirk\IEEEauthorrefmark{3},
%Montgomery Scott\IEEEauthorrefmark{3} and
%Eldon Tyrell\IEEEauthorrefmark{4}}
%\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Electrical and Computer Engineering\\
%Georgia Institute of Technology,
%Atlanta, Georgia 30332--0250\\ Email: see http://www.michaelshell.org/contact.html}
%\IEEEauthorblockA{\IEEEauthorrefmark{2}Twentieth Century Fox, Springfield, USA\\
%Email: homer@thesimpsons.com}
%\IEEEauthorblockA{\IEEEauthorrefmark{3}Starfleet Academy, San Francisco, California 96678-2391\\
%Telephone: (800) 555--1212, Fax: (888) 555--1212}
%\IEEEauthorblockA{\IEEEauthorrefmark{4}Tyrell Inc., 123 Replicant Street, Los Angeles, California 90210--4321}}

% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

% make the title area
\maketitle

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
The abstract goes here.
\end{abstract}

% no keywords

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

\section{Introduction}
% no \IEEEPARstart
Most of the radio spectrum requires a license to operate on. Devices emitting Radio Frequency (RF) energy without a license are limited to using one of the unlicensed radio bands, known as the Industrial, Scientific and Medical (ISM) radio bands. The most well-known of these bands is the 100MHz wide band centered around 2.45GHz, often called the 2.4GHz band. Many consumer devices such as wireless mice and keyboards, bluetooth headphones and Radio Controlled (RC) cars communicate in the 2.4GHz band. Furthermore most IEEE 802.11 (Wi-Fi) routers operate in the 2.4GHz band, along with the 5GHz band (which actually ranges from 5.725GHz to 5.875GHz). These ISM bands were originally intended and still used for devices emitting RF energy for purposes other than communication. The most common such application is the microwave oven. All this leads to high interference in these bands. In this paper, we investigate how bad this 2.4GHz and 5GHz interference actually is in the city of Antwerp. \\ \\
To measure this interference we use the City of Things (CoT) network, operated by research institute imec. This network consists of hundreds of sensors and wireless gateways positioned around the city of Antwerp. On TODO (20?) of these gateways, we sampled activity on the 2.4GHz and 5GHz bands every 10 minutes for TODO days. The main motivation of this paper was in fact to investigate whether interference around the gateways is significant enough to pose a problem for the network. Next to the general magnitude of the interference, we look for patterns in daily and weekly variation of the interference.

Some citing \cite{adams4242save} % At least one citation is needed, otherwise IEEEtran gives a nice error.

\hfill mds

\hfill August 26, 2015

\section{Measuring Interference}
All used nodes have a COMPEX WLE900VX-7A network adapter, containing a Qualcomm Atheros QCA9880 wireless chipset. This 802.11ac chipset, along with all other 802.11ac and 802.11n chips by Qualcomm Atheros support a mode called \textit{spectral scan}. In this mode the chip can scan the activity on all frequencies it supports. This activity is not limited to 802.11 traffic, but is influenced by any received signal. In this mode, the chip acts as a simple spectrum analyzer. The Free and Open-Source Software (FOSS) ath9k and ath10k drivers, respectively for the 802.11n and 802.11ac Qualcomm Atheros chipsets, support this mode. Each scan is performed over a period of \SI{4}{\micro\second}. Every scanned channel is divided into 64, 128 or 256 equally wide bins. For the full channel the noise floor and Received Signal Strength Indicator (RSSI) are reported, along with a magnitude for each bin. This magnitude incidates  how the power within the channel was divided across the bins. We provide an overview of how to interpret these values in TODO. \\
\subsection{Settings}
The ath10k version of spectral scan offers fewer options than the ath9k version. For instance the option to increase the scan time from \SI{4}{\micro\second} to \SI{2044}{\micro\second} is absent in ath10k. The only missing option relevant to our case is the \textit{chanscan} mode of spectral scan. When this mode is triggered the driver gathers a configurable number of samples for every channel it supports. We emulate this behaviour using the \textit{background} mode, which continuously scans the chipset's currently configured channel as long as it is not sending or receiving. We combine this with an \textit{iw scan} command, which listens for access point beacons for a provided list of frequencies. TODO PASSIVE When providing \textit{iw scan} with a list of all supported frequencies, this closely emulates the \textit{chanscan} behaviour.

\subsection{Interpreting spectral scans}
Before going into the interpretation of this specific data, we provide a quick overview of the dBm unit, often used to express power ratios. The regular decibel (dB) unit is a dimensionless unit. In contrast dBm expresses absolute power in reference to watt. 0dBm is defined as 1 milliwatt (mW). For each increase of 3dBm, the power in mW doubles, and for each increase of 10dBm, the power is multiplied by 10. Milliwatt-to-dBm conversion is calculated using the formula $10 log_{10}(\frac{x}{\SI{1}{\milli\watt}})$ with $x$ the power in mW. A negative dBm power simply means the power is less than \SI{1}{\milli\watt}. By using dBm, very small or large power values are often avoided. As power attenuates quadratically with distance, useful power values can vary dramatically.\\\\
For each channel scan, the driver reports the noise floor and RSSI value. The noise floor value is hard-coded for every channel and ranges between -95dBm and -106dBm. This value is the expected power of all noise in a channel combined. The chipset can only receive a singal whose power exceeds this noise floor. The RSSI value is a (unitless) integer indicating, as the name implies, the received power in the channel. While the existence of this value is part of the 802.11 specification, its calculation is not. As a result every manufacturer calculates the RSSI differently. Atheros chips follow a rather simple formula: subtracting the noise floor from the current signal strength in dBm results in the RSSI. These two values are enough to estimate the signal strength in a channel: a noise floor of -96dBm and an RSSI of 40 indicate a received signal of -56dBm. \\
The per-bin magnitudes indicate how power is divided within the channel. To understand these values, some background about radio signaling is needed. Signals in radio communication are usually created using a technique called \textit{IQ modulation}. With this technique, the final signal is a combination of two separate signals (the in-phase or \textit{I} and the quadrature or \textit{Q} signals) whose phases are exactly 90 degrees apart. By changing only the amplitude of the \textit{I} and \textit{Q} signals, the resulting signal's amplitude and phase can attain any value. The reported bin magnitude $b(i)$ for bin $i$ is the sum of the absolute values of the \textit{I} and \textit{Q} signals' magnitudes. This magnitude in turn is the amplitude squared. The per-bin power scales quadratically with this magnitude value: if bin $i$ has twice the magnitude of bin $j$, bin $i$'s share of the channel's power is four times as large as bin $j$'s. We can compute a coefficient $c(i)$ for every bin $i \in [1,n]$ as follows: $c(i) = \frac{b(i)^2}{\sum\limits_{j=1}^{n}b(j)^2}$. Note that every $c(i) \in [0,1]$ and $\sum\limits_{c=1}^{n}{c(i)} = 1$. \\
As the dBm power values computed from the noise floor and RSSI are not linear but rather logarithmic, we cannot simply multiply by $c(i)$ to get each bin's power value. Instead we first convert $c(i)$ to a logarithmic value: $c_{log}(i) = 10 log_{10}(c(i))$. As $log(a*b) = log(a) + log(b)$, we add this coefficient to the total logarithmic power to calculate each share's power: $nf + RSSI + c_{log}(i)$. Noting that $log(a/b) = log(a) - log(b)$, writing the formula using only the parameters returned by the spectral scan results in $nf + RSSI  + 10 log_{10}(b(i)^2) - 10 log_{10}(\sum\limits_{j=1}^{n}{b(j)^2})$.
The earliest reference to this formula being used with the FOSS Atheros drivers was in an email on the ath9k-devel mailing list by Zefir Kurtisi (https://www.spinics.net/lists/linux-wireless/msg101011.html TODO cite), in reply to the Request For Comments (RFC) on spectral scan support by Simon Wunderlich. As far as we know neither this formula nor the calculation of its parameters were ever confirmed to be correct by Qualcomm Atheros. This remains an educated guess by the community.
\subsection{Subsection Heading Here}
Subsection text here.

\subsubsection{Subsubsection Heading Here}
Subsubsection text here.

% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.
% For figures, \caption should occur after the \includegraphics.
% Note that IEEEtran v1.7 and later has special internal code that
% is designed to preserve the operation of \label within \caption
% even when the captionsoff option is in effect. However, because
% of issues like this, it may be the safest practice to put all your
% \label just after \caption rather than within \caption{}.
%
% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
% option should be used if it is desired that the figures are to be
% displayed while in draft mode.
%
%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{myfigure}
% where an .eps filename suffix will be assumed under latex,
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
%\caption{Simulation results for the network.}
%\label{fig_sim}
%\end{figure}

% Note that the IEEE typically puts floats only at the top, even when this
% results in a large percentage of a column being occupied by floats.

% An example of a double column floating figure using two subfigures.
% (The subfig.sty package must be loaded for this to work.)
% The subfigure \label commands are set within each subfloat command,
% and the \label for the overall figure must come after \caption.
% \hfil is used as a separator to get equal spacing.
% Watch out that the combined width of all the subfigures on a
% line do not exceed the text width or a line break will occur.
%
%\begin{figure*}[!t]
%\centering
%\subfloat[Case I]{\includegraphics[width=2.5in]{box}%
%\label{fig_first_case}}
%\hfil
%\subfloat[Case II]{\includegraphics[width=2.5in]{box}%
%\label{fig_second_case}}
%\caption{Simulation results for the network.}
%\label{fig_sim}
%\end{figure*}
%
% Note that often IEEE papers with subfigures do not employ subfigure
% captions (using the optional argument to \subfloat[]), but instead will
% reference/describe all of them (a), (b), etc., within the main caption.
% Be aware that for subfig.sty to generate the (a), (b), etc., subfigure
% labels, the optional argument to \subfloat must be present. If a
% subcaption is not desired, just leave its contents blank,
% e.g., \subfloat[].

% An example of a floating table. Note that, for IEEE style tables, the
% \caption command should come BEFORE the table and, given that table
% captions serve much like titles, are usually capitalized except for words
% such as a, an, and, as, at, but, by, for, in, nor, of, on, or, the, to
% and up, which are usually not capitalized unless they are the first or
% last word of the caption. Table text will default to \footnotesize as
% the IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
%\begin{table}[!t]
%% increase table row spacing, adjust to taste
%\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
%\caption{An Example of a Table}
%\label{table_example}
%\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
%\begin{tabular}{|c||c|}
%\hline
%One & Two\\
%\hline
%Three & Four\\
%\hline
%\end{tabular}
%\end{table}

% Note that the IEEE does not put floats in the very first column
% - or typically anywhere on the first page for that matter. Also,
% in-text middle ("here") positioning is typically not used, but it
% is allowed and encouraged for Computer Society conferences (but
% not Computer Society journals). Most IEEE journals/conferences use
% top floats exclusively.
% Note that, LaTeX2e, unlike IEEE journals/conferences, places
% footnotes above bottom floats. This can be corrected via the
% \fnbelowfloat command of the stfloats package.

\section{Conclusion}
The conclusion goes here.

% conference papers do not normally have an appendix

% use section* for acknowledgment
\section*{Acknowledgment}

The authors would like to thank...

% trigger a \newpage just before the given reference
% number - used to balance the columns on the last page
% adjust value as needed - may need to be readjusted if
% the document is modified later
%\IEEEtriggeratref{8}
% The "triggered" command can be changed if desired:
%\IEEEtriggercmd{\enlargethispage{-5in}}

% references section

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://mirror.ctan.org/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,bib}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)

\bibliographystyle{IEEEtran}
\bibliography{IEEEabrv,../chapters/bibliography}

% that's all folks
\end{document}

### 2017-05-12 cot/process_scans/fft_get_max_rssi.c
Added plotting code with smoothing, adapted single-channel to it, added option for better smoothing through prev/next day's data

@@ -25,19 +25,22 @@
  */
 
 #ifdef __APPLE__
-  #include <libkern/OSByteOrder.h>
-  #define CONVERT_BE16(val)	val = OSSwapBigToHostInt16(val)
-  #define CONVERT_BE64(val)	val = OSSwapBigToHostInt64(val)
+#include <libkern/OSByteOrder.h>
+#define CONVERT_BE16(val)	val = OSSwapBigToHostInt16(val)
+#define CONVERT_BE64(val)	val = OSSwapBigToHostInt64(val)
 #else
-  #define _BSD_SOURCE
-  #ifdef	__FreeBSD__
-    #include <sys/endian.h>
-  #else
-    #include <endian.h>
-  #endif	/* __FreeBSD__ */
-  #define CONVERT_BE16(val)	val = be16toh(val)
-  #define CONVERT_BE64(val)	val = be64toh(val)
+#define _BSD_SOURCE
+#ifdef    __FreeBSD__
+#include <sys/endian.h>
+#else
+
+#include <endian.h>
+
+#endif	/* __FreeBSD__ */
+#define CONVERT_BE16(val)    val = be16toh(val)
+#define CONVERT_BE64(val)    val = be64toh(val)
 #endif
+
 #include <errno.h>
 #include <stdio.h>
 #include <math.h>
@@ -52,74 +55,74 @@ typedef uint16_t u16;
 typedef uint64_t u64;
 
 enum ath_fft_sample_type {
-        ATH_FFT_SAMPLE_HT20 = 1,
-        ATH_FFT_SAMPLE_HT20_40 = 2,
-	ATH_FFT_SAMPLE_ATH10K = 3,
+    ATH_FFT_SAMPLE_HT20 = 1,
+    ATH_FFT_SAMPLE_HT20_40 = 2,
+    ATH_FFT_SAMPLE_ATH10K = 3,
 };
 
 enum nl80211_channel_type {
-	NL80211_CHAN_NO_HT,
-	NL80211_CHAN_HT20,
-	NL80211_CHAN_HT40MINUS,
-	NL80211_CHAN_HT40PLUS
+    NL80211_CHAN_NO_HT,
+    NL80211_CHAN_HT20,
+    NL80211_CHAN_HT40MINUS,
+    NL80211_CHAN_HT40PLUS
 };
 
 /*
  * ath9k spectral definition
  */
 #define SPECTRAL_HT20_NUM_BINS          56
-#define SPECTRAL_HT20_40_NUM_BINS		128
[…]

### 2017-05-12 cot/process_scans/plot_rssi.py

@@ -1,31 +1,175 @@
 import itertools
 import subprocess
 import sys
-import os.path
+
 from freq_channelnr_map import map_ as channelnr_map
+
 counter = 0
 maxes = []
+import numpy as np
+import matplotlib.pyplot as plt
+import matplotlib.patches as mpatches
 
-if (len(sys.argv) < 5):
-    print "Usage: python plot_rssi.py dump_directory date first_minute filter_num (channel number) \n\
-           Example: To plot for 1 May 2017, first dump at 2 minutes past the hour ignoring the top 100 results for every dump: \n\
-           python plot_rssi.py ../output_dir 2017-05-01 2 100 \n\
-           Example: To plot for channel 36 on 2 May 2017, first dump at 2 minutes past the hour ignoring the top 100 results for every dump: \n\
-           python plot_rssi.py ../output_dir 2017-05-02 2 100 36"
-    exit(0)
 
-for(hour,minute) in itertools.product(range(24), range(int(sys.argv[3]),60,10)):
-    channel_nr = int(sys.argv[5]) if (len(sys.argv) == 6) else 0
-    signals = [] #Fill with every nf+rssi of dump
-    file_name = sys.argv[1] + ('/' if not sys.argv[1].endswith('/') else '') + sys.argv[2] + '_' + str(hour).zfill(2) + '-' + str(minute).zfill(2) + '-01'
-    if (not os.path.isfile(file_name)):
-        #print("File " + file_name + " does not exist in the dump directory")
-        continue
-    signalstr = subprocess.check_output(['./fft_get_max_rssi.out', file_name]) if not channel_nr else subprocess.check_output(['./fft_get_max_rssi.out', file_name, channelnr_map[channel_nr]])
+
+### REMOVE SECONDS FROM FILENAMES WITH
+### ls -d ??????????????????? | xargs rename 's/.{3}$//'
+
+
+
+
+def savitzky_golay(y, window_size, order, deriv=0, rate=1):
+    r"""Smooth (and optionally differentiate) data with a Savitzky-Golay filter.
+    The Savitzky-Golay filter removes high frequency noise from data.
+    It has the advantage of preserving the original shape and
+    features of the signal better than other types of filtering
+    approaches, such as moving averages techniques.
+    Parameters
+    ----------
+    y : array_like, shape (N,)
+        the values of the time history of the signal.
+    window_size : int
+        the length of the window. Must be an odd integer number.
[…]

### 2017-05-15 cot/process_scans/plot_week_smoothings.py
Added script to plot multiple days' smoothings near to eachother (did not replot raw as that would serve no purpose and be hard to interpret). Made version of paper without ieeetrans. See file plot_week_smoothings.py for a command to run it.
Former-commit-id: 8036f473116a691f5672a4a1d845d62988c97d6a [formerly 7029ba1b22d1fb8f80b0ec52e2d34a1144971244 [formerly 7029ba1b22d1fb8f80b0ec52e2d34a1144971244 [formerly 56be9e990703802eec95d4a0c80c5cb47706447c]]]
Former-commit-id: f15a772965304bed59054d56b1aea7b2f2060a2f
Former-commit-id: 608edf4b7c9c4e58227d41aefbe21dd136f2607f
Former-commit-id: 2a12308

@@ -1,0 +1,195 @@
+import itertools
+import subprocess
+import sys
+
+from freq_channelnr_map import map_ as channelnr_map
+
+counter = 0
+maxes = []
+import numpy as np
+import matplotlib.pyplot as plt
+import matplotlib.patches as mpatches
+
+
+
+### REMOVE SECONDS FROM FILENAMES WITH
+### ls -d ??????????????????? | xargs rename 's/.{3}$//'
+
+def savitzky_golay(y, window_size, order, deriv=0, rate=1):
+    r"""Smooth (and optionally differentiate) data with a Savitzky-Golay filter.
+    The Savitzky-Golay filter removes high frequency noise from data.
+    It has the advantage of preserving the original shape and
+    features of the signal better than other types of filtering
+    approaches, such as moving averages techniques.
+    Parameters
+    ----------
+    y : array_like, shape (N,)
+        the values of the time history of the signal.
+    window_size : int
+        the length of the window. Must be an odd integer number.
+    order : int
+        the order of the polynomial used in the filtering.
+        Must be less then `window_size` - 1.
+    deriv: int
+        the order of the derivative to compute (default = 0 means only smoothing)
+    Returns
+    -------
+    ys : ndarray, shape (N)
+        the smoothed signal (or it's n-th derivative).
+    Notes
+    -----
+    The Savitzky-Golay is a type of low-pass filter, particularly
+    suited for smoothing noisy data. The main idea behind this
+    approach is to make for each point a least-square fit with a
+    polynomial of high order over a odd-sized window centered at
+    the point.
+    Examples
+    --------
+    t = np.linspace(-4, 4, 500)
+    y = np.exp( -t**2 ) + np.random.normal(0, 0.05, t.shape)
+    ysg = savitzky_golay(y, window_size=31, order=4)
+    import matplotlib.pyplot as plt
+    plt.plot(t, y, label='Noisy signal')
+    plt.plot(t, np.exp(-t**2), 'k', lw=1.5, label='Original signal')
+    plt.plot(t, ysg, 'r', label='Filtered signal')
+    plt.legend()
+    plt.show()
+    References
+    ----------
+    .. [1] A. Savitzky, M. J. E. Golay, Smoothing and Differentiation of
+       Data by Simplified Least Squares Procedures. Analytical
+       Chemistry, 1964, 36 (8), pp 1627-1639.
[…]

### 2017-05-15 advnetwpaper/paper_no_ieeetrans.tex
Added script to plot multiple days' smoothings near to eachother (did not replot raw as that would serve no purpose and be hard to interpret). Made version of paper without ieeetrans. See file plot_week_smoothings.py for a command to run it.

--- 2017-05-01 advnetwpaper/paper.tex
@@
-%% bare_conf.tex
-%% V1.4b
-%% 2015/08/26
-%% by Michael Shell
-%% See:
-%% http://www.michaelshell.org/
-%% for current contact information.
-%%
-%% This is a skeleton file demonstrating the use of IEEEtran.cls
-%% (requires IEEEtran.cls version 1.8b or later) with an IEEE
-%% conference paper.
-%%
-%% Support sites:
-%% http://www.michaelshell.org/tex/ieeetran/
-%% http://www.ctan.org/pkg/ieeetran
-%% and
-%% http://www.ieee.org/
-
-%%*************************************************************************
-%% Legal Notice:
-%% This code is offered as-is without any warranty either expressed or
-%% implied; without even the implied warranty of MERCHANTABILITY or
-%% FITNESS FOR A PARTICULAR PURPOSE!
-%% User assumes all risk.
-%% In no event shall the IEEE or any contributor to this code be liable for
-%% any damages or losses, including, but not limited to, incidental,
-%% consequential, or any other damages, resulting from the use or misuse
-%% of any information contained here.
-%%
-%% All comments are the opinions of their respective authors and are not
-%% necessarily endorsed by the IEEE.
-%%
-%% This work is distributed under the LaTeX Project Public License (LPPL)
-%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
-%% distributed and modified. A copy of the LPPL, version 1.3, is included
-%% in the base LaTeX documentation of all distributions of LaTeX released
-%% 2003/12/01 or later.
-%% Retain all contribution notices and credits.
-%% ** Modified files should be clearly indicated as such, including  **
-%% ** renaming them and changing author support contact information. **
-%%*************************************************************************
-
-% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
-% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
-% *** with production work. The IEEE's font choices and paper sizes can   ***
-% *** trigger bugs that do not appear when using other class files.       ***                          ***
-% The testflow support page is at:
-% http://www.michaelshell.org/tex/testflow/
-
-\documentclass[conference]{IEEEtran}
+\documentclass[a4paper, 11pt]{article}
 % Some Computer Society conferences also require the compsoc mode option,
@@
 % The documentation is contained in the cite.sty file itself.
-
 \usepackage{siunitx}
@@
 
-% *** GRAPHICS RELATED PACKAGES ***
-%
-\ifCLASSINFOpdf
-  % \usepackage[pdftex]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../pdf/}{../jpeg/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
-\else
-  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
-  % will default to the driver specified in the system graphics.cfg if no
-  % driver is specified.
-  % \usepackage[dvips]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../eps/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.eps}
-\fi
-% graphicx was written by David Carlisle and Sebastian Rahtz. It is
-% required if you want graphics, photos, etc. graphicx.sty is already
-% installed on most LaTeX systems. The latest version and documentation
-% can be obtained at:
-% http://www.ctan.org/pkg/graphicx
-% Another good source of documentation is "Using Imported Graphics in
-% LaTeX2e" by Keith Reckdahl which can be found at:
-% http://www.ctan.org/pkg/epslatex
-%
-% latex, and pdflatex in dvi mode, support graphics in encapsulated
-% postscript (.eps) format. pdflatex in pdf mode supports graphics
-% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
-% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
-% not a bitmapped formats (.jpeg, .png). The IEEE frowns on bitmapped formats
-% which can result in "jaggedy"/blurry rendering of lines and letters as
-% well as large increases in file sizes.
-%
-% You can find documentation about the pdfTeX application at:
-% http://www.tug.org/applications/pdftex
-
 % *** MATH PACKAGES ***
 %
-%\usepackage{amsmath}
+\usepackage{amsmath}
 % A popular package from the American Mathematical Society that provides
@@
 % Do not put math or special symbols in the title.
-\title{Bare Demo of IEEEtran.cls\\ for IEEE Conferences}
+\title{A study on the usage of the wireless signal spectrum in a City of Things network}
 
@@
 % affiliations
-\author{\IEEEauthorblockN{Michael Shell}
-\IEEEauthorblockA{School of Electrical and\\Computer Engineering\\
-Georgia Institute of Technology\\
-Atlanta, Georgia 30332--0250\\
-Email: http://www.michaelshell.org/contact.html}
-\and
-\IEEEauthorblockN{Homer Simpson}
-\IEEEauthorblockA{Twentieth Century Fox\\
-Springfield, USA\\
-Email: homer@thesimpsons.com}
-\and
-\IEEEauthorblockN{James Kirk\\ and Montgomery Scott}
-\IEEEauthorblockA{Starfleet Academy\\
-San Francisco, California 96678--2391\\
-Telephone: (800) 555--1212\\
-Fax: (888) 555--1212}}
+\author{
+	Jakob Struye (s.......)\\ 
+    Thomas Hendriks (s0139858)\\
+}
 
@@
 
-% no keywords
-
-% For peer review papers, you can put extra information on the cover
-% page as needed:
-% \ifCLASSOPTIONpeerreview
-% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
-% \fi
-%
-% For peerreview papers, this IEEEtran command inserts a page break and
-% creates the second title. It will be ignored for other modes.
-\IEEEpeerreviewmaketitle
-
 \section{Introduction}
@@
 This network consists of hundreds of sensors and wireless gateways positioned around the city of Antwerp.
-On TODO (20?) of these gateways, we sampled activity on the 2.4GHz and 5GHz bands every 10 minutes for TODO days.
+On 13 of these gateways, we sampled activity on the 2.4GHz and 5GHz bands every 10 minutes for TODO days.
 The main motivation of this paper was in fact to investigate whether interference around the gateways is significant enough to pose a problem for the network.
@@
 Some citing \cite{adams4242save} % At least one citation is needed, otherwise IEEEtran gives a nice error.
-
-\hfill mds
-
-\hfill August 26, 2015
 
@@
 The per-bin power scales quadratically with this magnitude value: if bin $i$ has twice the magnitude of bin $j$, bin $i$'s share of the channel's power is four times as large as bin $j$'s.
-We can compute a coefficient $c(i)$ for every bin $i \in [1,n]$ as follows: $c(i) = \frac{b(i)^2}{\sum\limits_{j=1}^{n}b(j)^2}$.
-Note that every $c(i) \in [0,1]$ and $\sum\limits_{c=1}^{n}{c(i)} = 1$.
-\\
+We can compute a coefficient $c(i)$ for every bin $i \in [1,n]$ as follows:\\\\\hspace*{1.5em}$c(i) = \dfrac{b(i)^2}{\sum\limits_{j=1}^{n}b(j)^2}$ (Note that every $c(i) \in [0,1]$ and $\sum\limits_{c=1}^{n}{c(i)} = 1$.) \\\\
 As the dBm power values computed from the noise floor and RSSI are not linear but rather logarithmic, we cannot simply multiply by $c(i)$ to get each bin's power value.
@@
 Subsubsection text here.
+
+\section{Measurements}
+At first we measured the spectrum on every channel in both the 2.4GHz and the 5GHz band every 10 minutes.
+Looking at interpretations of the resulting data it was clear that a shorter time between scans would be useful because recurring spikes were missed on some days while they definitely there and also quite interesting.
+\\< images waarop da duidelijk is >\\
+We then reduced the time between scans from 10 minutes to 2 minutes, which was the minimum we could achieve while guaranteeing that scans would be succesful.
+Using an adapted version of Simon Wünderlich's fft_eval code, GNUplot and Savitzky-Golay (source? TODO) filtering we graphically represent our data points per node.
+<tussen maandag 15 tem zondag 21 mei>
+
+\section{Findings}
 

### 2017-05-15 cot/process_scans/plot_week_smoothings.py
Replaced code duplication with an import stmt.
Former-commit-id: d0a92d1d4771baeaea967839437be473e0976469 [formerly 19cab2e8ff5389f7ce89699988455d209ae27cb8 [formerly 19cab2e8ff5389f7ce89699988455d209ae27cb8 [formerly 94f753179a5a64e639c89b7514eb9eef7f61955a]]]
Former-commit-id: 6f5910a716f46e8a7c4755f05d37baa928523b77
Former-commit-id: 35688990b4f6e6ce52d3259304d95069c24f8235
Former-commit-id: 6632583b99e9111363df5a7da1b5fc8a546d099e

@@ -4,88 +4,13 @@ import sys
 
 from freq_channelnr_map import map_ as channelnr_map
 
+from plot_rssi import savitzky_golay as savitzky_golay
+
 counter = 0
 maxes = []
 import numpy as np
 import matplotlib.pyplot as plt
 import matplotlib.patches as mpatches
-
-
-
-### REMOVE SECONDS FROM FILENAMES WITH
-### ls -d ??????????????????? | xargs rename 's/.{3}$//'
-
-def savitzky_golay(y, window_size, order, deriv=0, rate=1):
-    r"""Smooth (and optionally differentiate) data with a Savitzky-Golay filter.
-    The Savitzky-Golay filter removes high frequency noise from data.
-    It has the advantage of preserving the original shape and
-    features of the signal better than other types of filtering
-    approaches, such as moving averages techniques.
-    Parameters
-    ----------
-    y : array_like, shape (N,)
-        the values of the time history of the signal.
-    window_size : int
-        the length of the window. Must be an odd integer number.
-    order : int
-        the order of the polynomial used in the filtering.
-        Must be less then `window_size` - 1.
-    deriv: int
-        the order of the derivative to compute (default = 0 means only smoothing)
-    Returns
-    -------
-    ys : ndarray, shape (N)
-        the smoothed signal (or it's n-th derivative).
-    Notes
-    -----
-    The Savitzky-Golay is a type of low-pass filter, particularly
-    suited for smoothing noisy data. The main idea behind this
-    approach is to make for each point a least-square fit with a
-    polynomial of high order over a odd-sized window centered at
-    the point.
-    Examples
-    --------
-    t = np.linspace(-4, 4, 500)
-    y = np.exp( -t**2 ) + np.random.normal(0, 0.05, t.shape)
-    ysg = savitzky_golay(y, window_size=31, order=4)
-    import matplotlib.pyplot as plt
-    plt.plot(t, y, label='Noisy signal')
-    plt.plot(t, np.exp(-t**2), 'k', lw=1.5, label='Original signal')
-    plt.plot(t, ysg, 'r', label='Filtered signal')
-    plt.legend()
-    plt.show()
-    References
-    ----------
-    .. [1] A. Savitzky, M. J. E. Golay, Smoothing and Differentiation of
-       Data by Simplified Least Squares Procedures. Analytical
-       Chemistry, 1964, 36 (8), pp 1627-1639.
-    .. [2] Numerical Recipes 3rd Edition: The Art of Scientific Computing
[…]

### 2017-05-25 cot/process_scans/plot_week_smoothings.py
Some changes, generating 'clean' graphs.
Former-commit-id: 1a7bc6e8a19fbc705126c6a923c77dfe79a505c4 [formerly 11f514391f8bc341adfd4bb3e392ce7ee29cea38 [formerly 11f514391f8bc341adfd4bb3e392ce7ee29cea38 [formerly 7782d86b59ae16b6435afe35218fa06979b4e7ea]]]
Former-commit-id: c52c1e1111a597ff5b1e15846dc7f35aabc778db
Former-commit-id: 2d549bddc4add11ed890aabea0324df394549664
Former-commit-id: 40e696ebd3d6c1259f82db2c3636103d06c30320

@@ -6,6 +6,8 @@ from freq_channelnr_map import map_ as channelnr_map
 
 from plot_rssi import savitzky_golay as savitzky_golay
 
+#rename 's/...$//' *
+
 counter = 0
 maxes = []
 import numpy as np
@@ -17,6 +19,7 @@ def process_point(day, hour, minute):
     signals = []  # Fill with every nf+rssi of dump
     file_name = directory + ('/' if not directory.endswith('/') else '') + day + '_' + str(hour).zfill(2) + '-' + str(
         minute).zfill(2)
+
     signalstr = subprocess.check_output(
         ['./fft_get_max_rssi.out', file_name]) if not channel_number else subprocess.check_output(
         ['./fft_get_max_rssi.out', file_name, channelnr_map[channel_number]])
@@ -24,15 +27,22 @@ def process_point(day, hour, minute):
         signals.append(int(line))
     signals.sort()
     signals.reverse()
+
+    if (len(signals) == 0):
+        print(file_name)
+        print ""
     return signals[filter_num]
 
 
 if __name__ == '__main__':
-    colors = ["blue","green","red","cyan","magenta","yellow","black","white"]
+    if (len(sys.argv) < 9 and len(sys.argv) >= 8) :
+        print("Please also supply node name at the end of command")
+    print(sys.argv[8])
+    colors = ["blue","green","red","cyan","magenta","yellow","black","0.75", "#788599", "#2c8e36", "#7f0043"]
     if (len(sys.argv) < 6):
         print "Usage: python plot_rssi.py dump_directory comma_separated_list_of_days first_minute filter_num channel_number [prev_day next_day]\n\
                Example: To plot for all channels on 2,3,4 May 2017, first dump at 2 minutes past the hour ignoring the top 100 results for every dump: \n\
-               python plot_week_smoothings.py ../output_dir 2017-05-02,2017-05-03,2017-05-04 0 100 0 2017-05-01 2017-05-05\n\
+               python plot_week_smoothings.py ../output_dir 2017-05-02,2017-05-03,2017-05-04 0 100 0 2017-05-01 2017-05-05 node_name\n\
                Note that the channel number is required: 0 means all channels\n\
                To use previous and next day's last and first hour to avoid potentially ugly results at end points.\n\
                These are only used in calculations and not actually plotted."
[…]

### 2017-05-25 cot/process_scans/make_images.sh
Some changes, generating 'clean' graphs.
Former-commit-id: 7782d86b59ae16b6435afe35218fa06979b4e7ea

@@ -1,0 +1,56 @@
+#!/bin/bash
+#2.4Ghz --> 0
+#5  Ghz --> 1
+type=0
+
+for i in cot-node3-student cot-node4-student cot-node7-student cot-node9-student cot-node10-student cot-node12-student ; do
+  for j in 1 6 11 ; do
+    python plot_week_smoothings.py ~/Desktop/UA/AdvNetwLab/current_backup/$i/outputFLUSH/ 2017-05-15,2017-05-16,2017-05-17,2017-05-18,2017-05-19,2017-05-20,2017-05-21,2017-05-22 $type 5 $j 2017-05-14 2017-05-23 $i
+  done
+done
+
+#For cot-node4-student          : 9999-99-99_99-99
+#For cot-node3-student          : 9999-99-99_99-99
+#For cot-node7-student          : 9999-99-99_99-99
+#For cot-node9-student          : 9999-99-99_99-99
+#For cot-node12-student         : 9999-99-99_99-99
+#For cot-node10-student         : 2017-05-11_11-52 <- enige die 0 is vanwege reboot dus resultaat is nog ok
+
+i='cot-node6-student'
+#For cot-node6-student          : 2017-05-22_21-31 <- 12 -> 20 is doable with 11 and 21 to smoothe
+# python plot_week_smoothings.py ~/Desktop/UA/AdvNetwLab/current_backup/$i/outputFLUSH/ 2017-05-12,2017-05-13,2017-05-14,2017-05-15,2017-05-16,2017-05-17,2017-05-18,2017-05-19,2017-05-20 $type 100 0 2017-05-11 2017-05-21 $i
+i='node1'
+# #For node1                      : 2017-05-19_23-00 <- 13 -> 17 is doable with 12 and 18 to smoothe
+# python plot_week_smoothings.py ~/Desktop/UA/AdvNetwLab/current_backup/$i/outputFLUSH/ 2017-05-13,2017-05-14,2017-05-15,2017-05-16,2017-05-17 $type 100 0 2017-05-12 2017-05-18 $i
+i='cot-node5-student'
+# #For cot-node5-student          : 2017-05-18_21-46 <- 12 -> 16 is doable with 11 and 17 to smoothe things out
[…]

### 2017-05-25 advnetwpaper/paper_no_ieeetrans.tex
Some changes, generating 'clean' graphs.

@@
 
+%- algemene verloop doorheen dag/week
+%- recurring patterns tussen dagen
+%- drukkere/rustigere channels
+%- 2.4 vs 5 (gelijkaardig aan het vorige)
+
 % *** CITATION PACKAGES ***
@@
 \graphicspath{{../images/}}
+
+\usepackage{hyperref}
 
@@
 
-% conference papers do not typically use \thanks and this command
-% is locked out in conference mode. If really needed, such as for
-% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
-% after \documentclass
+How bad is it, in general? 
+Are there patterns, e.g. daily recurrence? 
+Advanced: how can you ensure data quality, i.e., how to validate the results?
 
-% for over three affiliations, or if they all won't fit within the width
-% of the page, use this alternative format:
-%
-%\author{\IEEEauthorblockN{Michael Shell\IEEEauthorrefmark{1},
-%Homer Simpson\IEEEauthorrefmark{2},
-%James Kirk\IEEEauthorrefmark{3},
-%Montgomery Scott\IEEEauthorrefmark{3} and
-%Eldon Tyrell\IEEEauthorrefmark{4}}
-%\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Electrical and Computer Engineering\\
-%Georgia Institute of Technology,
-%Atlanta, Georgia 30332--0250\\ Email: see http://www.michaelshell.org/contact.html}
-%\IEEEauthorblockA{\IEEEauthorrefmark{2}Twentieth Century Fox, Springfield, USA\\
-%Email: homer@thesimpsons.com}
-%\IEEEauthorblockA{\IEEEauthorrefmark{3}Starfleet Academy, San Francisco, California 96678-2391\\
-%Telephone: (800) 555--1212, Fax: (888) 555--1212}
-%\IEEEauthorblockA{\IEEEauthorrefmark{4}Tyrell Inc., 123 Replicant Street, Los Angeles, California 90210--4321}}
-
-% use for special paper notices
-%\IEEEspecialpapernotice{(Invited Paper)}
-
-% make the title area
 \maketitle
@@
 % in the abstract
-\begin{abstract}
-The abstract goes here.
-\end{abstract}
+
+\tableofcontents
+\clearpage
 
@@
 This network consists of hundreds of sensors and wireless gateways positioned around the city of Antwerp.
-On 13 of these gateways, we sampled activity on the 2.4GHz and 5GHz bands every 10 minutes for TODO days.
+On 11 of these gateways, we sampled activity on the 2.4GHz and 5GHz bands every 2 minutes for a little over 10 days, to ensure that we would have at least one full week of usable data for each node.
 The main motivation of this paper was in fact to investigate whether interference around the gateways is significant enough to pose a problem for the network.
 Next to the general magnitude of the interference, we look for patterns in daily and weekly variation of the interference.
-
-Some citing \cite{adams4242save} % At least one citation is needed, otherwise IEEEtran gives a nice error.
-
 \section{Measuring Interference}
@@
 This magnitude incidates  how the power within the channel was divided across the bins.
-We provide an overview of how to interpret these values in TODO.
+We provide an overview of how to interpret these values in \mbox{\hyperref[sec:interpret]{Interpreting spectral scans}}.
 \\
@@
 We combine this with an \textit{iw scan} command, which listens for access point beacons for a provided list of frequencies.
-TODO PASSIVE When providing \textit{iw scan} with a list of all supported frequencies, this closely emulates the \textit{chanscan} behaviour.
-
+When providing \textit{iw scan} with a list of all supported frequencies, this closely emulates the \textit{chanscan} behaviour.
+\label{sec:interpret}
 \subsection{Interpreting spectral scans}
@@
 In contrast dBm expresses absolute power in reference to watt.
-0dBm is defined as 1 milliwatt (mW).
+0 dBm is defined as 1 milliwatt (mW).
 For each increase of 3dBm, the power in mW doubles, and for each increase of 10dBm, the power is multiplied by 10.
-Milliwatt-to-dBm conversion is calculated using the formula $10 log_{10}(\frac{x}{\SI{1}{\milli\watt}})$ with $x$ the power in mW.
-A negative dBm power simply means the power is less than \SI{1}{\milli\watt}.
+Milliwatt-to-dBm conversion is calculated using the formula\\ \\ $dBm = 10 log_{10}(\dfrac{x}{\SI{1}{\milli\watt}})$ with $x$ the power in mW.\\ \\ A negative dBm power simply means the power is less than \SI{1}{\milli\watt}.
 By using dBm, very small or large power values are often avoided.
@@
 These two values are enough to estimate the signal strength in a channel: a noise floor of -96dBm and an RSSI of 40 indicate a received signal of -56dBm.
-\\
+TODO SOURCE ?
+(for example :http://www.n-cg.net/ncgpdf/WiFi\_SignalValues.pdf see conversion for atheros) \\
 The per-bin magnitudes indicate how power is divided within the channel.
@@
 As the dBm power values computed from the noise floor and RSSI are not linear but rather logarithmic, we cannot simply multiply by $c(i)$ to get each bin's power value.
-Instead we first convert $c(i)$ to a logarithmic value: $c_{log}(i) = 10 log_{10}(c(i))$.
-As $log(a*b) = log(a) + log(b)$, we add this coefficient to the total logarithmic power to calculate each share's power: $nf + RSSI + c_{log}(i)$.
-Noting that $log(a/b) = log(a) - log(b)$, writing the formula using only the parameters returned by the spectral scan results in $nf + RSSI  + 10 log_{10}(b(i)^2) - 10 log_{10}(\sum\limits_{j=1}^{n}{b(j)^2})$.
+Instead we first convert $c(i)$ to a logarithmic value: \\ \\ $c_{log}(i) = 10 log_{10}(c(i))$.
+\\ \\ As $log(a*b) = log(a) + log(b)$, we add this coefficient to the total logarithmic power to calculate each share's power:\\ \\ $nf + RSSI + c_{log}(i)$.\\ \\ Noting that $log(a/b) = log(a) - log(b)$, writing the formula using only the parameters returned by the spectral scan results in\\ \\ $nf + RSSI  + 10 log_{10}(b(i)^2) - 10 log_{10}(\sum\limits_{j=1}^{n}{b(j)^2})$ \\ \\
 The earliest reference to this formula being used with the FOSS Atheros drivers was in an email on the ath9k-devel mailing list by Zefir Kurtisi (https://www.spinics.net/lists/linux-wireless/msg101011.html TODO cite), in reply to the Request For Comments (RFC) on spectral scan support by Simon Wunderlich.
@@
 This remains an educated guess by the community.
-\subsection{Subsection Heading Here}
-Subsection text here.
-
-\subsubsection{Subsubsection Heading Here}
-Subsubsection text here.
 
@@
 At first we measured the spectrum on every channel in both the 2.4GHz and the 5GHz band every 10 minutes.
-Looking at interpretations of the resulting data it was clear that a shorter time between scans would be useful because recurring spikes were missed on some days while they definitely there and also quite interesting.
+Looking at graphic interpretations of the resulting data it was clear that a shorter time between scans would be more effective at showing us the state of the spectrum because recurring spikes were missed on some days despite their presence being clear in subsequent smaller tests.
 \\< images waarop da duidelijk is >\\
-We then reduced the time between scans from 10 minutes to 2 minutes, which was the minimum we could achieve while guaranteeing that scans would be succesful.
-Using an adapted version of Simon Wünderlich's fft_eval code, GNUplot and Savitzky-Golay (source? TODO) filtering we graphically represent our data points per node.
-<tussen maandag 15 tem zondag 21 mei>
+Another point that was raised following the first set of graphic interpretations was that of outlier detection and elimination, occasionally there were big spikes in the received signal strength which are most likely caused by interference.
+Playing around with some parameters, we reduced the time between scans from 10 minutes to 2 minutes and discard the 100 largest readings for every sample (when looking at all the channels).
+This limit of 2 minutes was the minimum we could achieve while guaranteeing a scan would succeed and the number of samples to discard was chosen based on some experimentation.
+Using an adapted version of Simon Wünderlich's fft$\_$eval code, Python's Matplotlib library and Savitzky-Golay (source?
+TODO) filtering we graphically represent our data points per node.
+For CoT-node12 (location : UAntwerpen campus Middelheim, indoors), looking at the 2.4 GHz band on every channel from monday May 15th to monday May 22nd, we got the following graph:\\
+\includegraphics[scale=0.7]{images/window_361_2.4/2017-05-22cot-node12-student_image.pdf}\\
+Each line/color represents a day, using this representation we tried to find recurring patterns in the measurements for every node separately.
+When looking at the complete set of channels, patterns were not as clear as we'd hoped.\\
+Given this and the fact that a large amount of static on one channel would make it seem like the entire spectrum is full we switched to drawing the graphs per non-overlapping channel and observing the results in that way.
+This lead to following graph on which we can clearly see that on weekdays there is substantially more traffic during the daytime than on the weekend (which is as expected for a university campus):\\
+\includegraphics[scale=0.7]{images/2017-05-22cot-node12-student_image.pdf}\\
 
 \section{Findings}
+\subsection{Evolution throughout the day}
+\subsection{Evolution throughout the week}
+\subsection{Recurring patterns between days} 
+On the different nodes we had varying degrees of success in finding recurring patterns.
+For this report we've only included the nodes on which we did find such patterns (and we have included the non pattern-y graphs in the zip file that is included // they are available on demand?
+TODO)
+\subsection{Calm and busy channels}
+1 busy, 6, 11 calm
+\subsection{2.4 GHz vs 5 GHz}
+\subsection{Validation}
+scans met / zonder sturende laptop - stukje had gij al
+dagelijkse patronen
+outside / inside t zelfde +-
 
-% An example of a floating figure using the graphicx package.
-% Note that \label must occur AFTER (or within) \caption.
-% For figures, \caption should occur after the \includegraphics.
-% Note that IEEEtran v1.7 and later has special internal code that
-% is designed to preserve the operation of \label within \caption
-% even when the captionsoff option is in effect. However, because
-% of issues like this, it may be the safest practice to put all your
-% \label just after \caption rather than within \caption{}.
-%
-% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
-% option should be used if it is desired that the figures are to be
-% displayed while in draft mode.
-%
-%\begin{figure}[!t]
-%\centering
-%\includegraphics[width=2.5in]{myfigure}
-% where an .eps filename suffix will be assumed under latex,
-% and a .pdf suffix will be assumed for pdflatex; or what has been declared
-% via \DeclareGraphicsExtensions.
-%\caption{Simulation results for the network.}
-%\label{fig_sim}
-%\end{figure}
-
-% Note that the IEEE typically puts floats only at the top, even when this
-% results in a large percentage of a column being occupied by floats.
-
-% An example of a double column floating figure using two subfigures.
-% (The subfig.sty package must be loaded for this to work.)
-% The subfigure \label commands are set within each subfloat command,
-% and the \label for the overall figure must come after \caption.
-% \hfil is used as a separator to get equal spacing.
-% Watch out that the combined width of all the subfigures on a
-% line do not exceed the text width or a line break will occur.
-%
-%\begin{figure*}[!t]
-%\centering
-%\subfloat[Case I]{\includegraphics[width=2.5in]{box}%
-%\label{fig_first_case}}
-%\hfil
-%\subfloat[Case II]{\includegraphics[width=2.5in]{box}%
-%\label{fig_second_case}}
-%\caption{Simulation results for the network.}
-%\label{fig_sim}
-%\end{figure*}
-%
-% Note that often IEEE papers with subfigures do not employ subfigure
-% captions (using the optional argument to \subfloat[]), but instead will
-% reference/describe all of them (a), (b), etc., within the main caption.
-% Be aware that for subfig.sty to generate the (a), (b), etc., subfigure
-% labels, the optional argument to \subfloat must be present. If a
-% subcaption is not desired, just leave its contents blank,
-% e.g., \subfloat[].
-
-% An example of a floating table. Note that, for IEEE style tables, the
-% \caption command should come BEFORE the table and, given that table
-% captions serve much like titles, are usually capitalized except for words
-% such as a, an, and, as, at, but, by, for, in, nor, of, on, or, the, to
-% and up, which are usually not capitalized unless they are the first or
-% last word of the caption. Table text will default to \footnotesize as
-% the IEEE normally uses this smaller font for tables.
-% The \label must come after \caption as always.
-%
-%\begin{table}[!t]
-%% increase table row spacing, adjust to taste
-%\renewcommand{\arraystretch}{1.3}
-% if using array.sty, it might be a good idea to tweak the value of
-% \extrarowheight as needed to properly center the text within the cells
-%\caption{An Example of a Table}
-%\label{table_example}
-%\centering
-%% Some packages, such as MDW tools, offer better commands for making tables
-%% than the plain LaTeX2e tabular which is used here.
-%\begin{tabular}{|c||c|}
-%\hline
-%One & Two\\
-%\hline
-%Three & Four\\
-%\hline
-%\end{tabular}
-%\end{table}
-
-% Note that the IEEE does not put floats in the very first column
-% - or typically anywhere on the first page for that matter. Also,
-% in-text middle ("here") positioning is typically not used, but it
-% is allowed and encouraged for Computer Society conferences (but
-% not Computer Society journals). Most IEEE journals/conferences use
-% top floats exclusively.
-% Note that, LaTeX2e, unlike IEEE journals/conferences, places
-% footnotes above bottom floats. This can be corrected via the
-% \fnbelowfloat command of the stfloats package.
-
-\section{Conclusion}
+\section{Conclusions}
 The conclusion goes here.
-
-% conference papers do not normally have an appendix
-
-% use section* for acknowledgment
-\section*{Acknowledgment}
-
-The authors would like to thank...
-
-% trigger a \newpage just before the given reference
-% number - used to balance the columns on the last page
-% adjust value as needed - may need to be readjusted if
-% the document is modified later
-%\IEEEtriggeratref{8}
-% The "triggered" command can be changed if desired:
-%\IEEEtriggercmd{\enlargethispage{-5in}}
-
-% references section
-
-% can use a bibliography generated by BibTeX as a .bbl file
-% BibTeX documentation can be easily obtained at:
-% http://mirror.ctan.org/biblio/bibtex/contrib/doc/
-% The IEEEtran BibTeX style support page is at:
-% http://www.michaelshell.org/tex/ieeetran/bibtex/
-%\bibliographystyle{IEEEtran}
-% argument is your BibTeX string definitions and bibliography database(s)
-%\bibliography{IEEEabrv,bib}
-%
-% <OR> manually copy in the resultant .bbl file
-% set second argument of \begin to the number of references
-% (used to reserve space for the reference number labels box)
 

### 2017-05-25 cot/process_scans/find_first_zero.py

@@ -1,0 +1,28 @@
+import os, subprocess
+import sys
+from subprocess import Popen, PIPE
+
+def f_comes_first(date1, date2):
+    date1_day = int(date1.split("_")[0].split("-")[2])
+    date2_day = int(date2.split("_")[0].split("-")[2])
+    if (date1_day < date2_day):
+        print(date1, date2)
+        return True
+    else:
+        return False
+
+
+if __name__ == "__main__":
+    filename = sys.argv[1]
+    currMinDate = "9999-99-99_99-99"
+
+    with open(os.devnull, "w") as z:
+        for f in os.listdir(filename):
+            p = Popen(["du", filename +""+ f], stdin=PIPE, stdout=PIPE, stderr=PIPE)
+            output, err = p.communicate(b"input data that is passed to subprocess' stdin")
+            size = int(output[0])
+            if (size == 0 and f_comes_first(f, currMinDate)):
+                print(currMinDate, size)
+                currMinDate = f
+
+    print(currMinDate)

### 2017-05-25 cot/process_scans/make_may1st.sh
Added some findings about 2.4 GHz. Still waiting for 5 GHz results to turn into graphs though for the most part those graphs do not contain anything interesting.
Former-commit-id: b1f19f6291b5d51ba187a394f93d16ed7697fa33

@@ -1,0 +1,21 @@
+#!/bin/bash
+#2.4Ghz --> type =0
+type=0
+start=1
+end=11
+diff=5
+
+#5  Ghz --> type = 1
+# type=1
+# start=36
+# end=64
+# diff=4
+
+for i in cot-node3-student cot-node4-student cot-node5-student cot-node6-student cot-node7-student cot-node8-student cot-node9-student cot-node10-student cot-node11-student cot-node12-student ; do
+  j=$start
+  while [ $j -le $end ]
+  do
+    python plot_week_smoothings_5min.py ~/Desktop/UA/AdvNetwLab/backup_mosaic_data_10_05_2017/$i/output/ 2017-04-29,2017-04-30,2017-05-01,2017-05-02,2017-05-03 $type 5 $j 2017-04-28 2017-05-04 $i
+    j=$[$j+$diff]
+  done
+done

### 2017-05-25 advnetwpaper/paper_no_ieeetrans.tex
Added some findings about 2.4 GHz. Still waiting for 5 GHz results to turn into graphs though for the most part those graphs do not contain anything interesting.

@@
 Looking at graphic interpretations of the resulting data it was clear that a shorter time between scans would be more effective at showing us the state of the spectrum because recurring spikes were missed on some days despite their presence being clear in subsequent smaller tests.
-\\< images waarop da duidelijk is >\\
+\\
 Another point that was raised following the first set of graphic interpretations was that of outlier detection and elimination, occasionally there were big spikes in the received signal strength which are most likely caused by interference.
-Playing around with some parameters, we reduced the time between scans from 10 minutes to 2 minutes and discard the 100 largest readings for every sample (when looking at all the channels).
-This limit of 2 minutes was the minimum we could achieve while guaranteeing a scan would succeed and the number of samples to discard was chosen based on some experimentation.
-Using an adapted version of Simon Wünderlich's fft$\_$eval code, Python's Matplotlib library and Savitzky-Golay (source?
+Playing around with some parameters, we reduced the time between scans from 10 minutes to 2 minutes and added a parameter to discard a number of the strongest signals to deal with these outliers.
+This limit of 2 minutes was the minimum we could achieve while guaranteeing a scan would succeed and the number of samples to discard can be tweaked as needed.
+Using an adapted version of Simon W{\"u}nderlich's fft$\_$eval code, Python's Matplotlib library, the mathematical formula's mentioned above and Savitzky-Golay (source?
 TODO) filtering we graphically represent our data points per node.
-For CoT-node12 (location : UAntwerpen campus Middelheim, indoors), looking at the 2.4 GHz band on every channel from monday May 15th to monday May 22nd, we got the following graph:\\
-\includegraphics[scale=0.7]{images/window_361_2.4/2017-05-22cot-node12-student_image.pdf}\\
-Each line/color represents a day, using this representation we tried to find recurring patterns in the measurements for every node separately.
-When looking at the complete set of channels, patterns were not as clear as we'd hoped.\\
-Given this and the fact that a large amount of static on one channel would make it seem like the entire spectrum is full we switched to drawing the graphs per non-overlapping channel and observing the results in that way.
-This lead to following graph on which we can clearly see that on weekdays there is substantially more traffic during the daytime than on the weekend (which is as expected for a university campus):\\
-\includegraphics[scale=0.7]{images/2017-05-22cot-node12-student_image.pdf}\\
-
+For CoT-node12 (location : UAntwerpen campus Middelheim, indoors), looking at the 2.4 GHz band on channel 1 from monday May 15th to monday May 22nd, yields the following graph:\\
+\begin{figure}[h!]
+    \centering
+    \textbf{CoT-node12-student, channel 1, 15/05 - 22/05}\par\medskip
+	\includegraphics[scale=0.7]{images/2_4_GHz/cot-node12-student_2017-05-22_chan1_image.png}
+\end{figure}
+\clearpage
+Each line/color represents a day, the legend is the following (will be the same for every graph unless indicated otherwise):
+\begin{itemize}
+\setlength\itemsep{-0.4em}
+\item Blue: Monday, May 15th	
+\item Green: Tuesday, May 16th
+\item Red: Wednesday, May 17th
+\item Cyan: Thursday, May 18th
+\item Magenta: Friday, May 19th
+\item Yellow: Saturday, May 20th
+\item Black: Sunday, May 21st
+\item Grey: Monday, May 22nd
+\end{itemize}
+Looking at the spectrum with all the channels as a whole proved not very useful when assessing the state of the spectrum because the if one of the non-overlapping channels is busy it would sway the graph to reflect the entire band to be busy, which would not be the case.
 \section{Findings}
 \subsection{Evolution throughout the day}
-\subsection{Evolution throughout the week}
+
 \subsection{Recurring patterns between days} 
 On the different nodes we had varying degrees of success in finding recurring patterns.
-For this report we've only included the nodes on which we did find such patterns (and we have included the non pattern-y graphs in the zip file that is included // they are available on demand?
-TODO)
+In the 2.4 GHz band, two nodes that clearly showed repeating patterns across several days were cot-node8-student and cot-node12-student.
+Both of these nodes detected patterns, node8 detecting a large increase in spectrum usage in the afternoon and node12 detecting an increase during the daytime on weekdays, while on the weekends there was no increase present.
+\begin{figure}[!htb]
+    \centering
+    \textbf{CoT-node12-student, channel 1, 15/05 - 22/05}\par\medskip
+	\includegraphics[scale=0.5]{images/2_4_GHz/cot-node12-student_2017-05-22_chan1_image.png}
+\end{figure}
+\begin{figure}[htb!]
+    \centering
+    \textbf{CoT-node8-student, channel 1, 12/05 - 16/05}\par\medskip
+	\includegraphics[scale=0.5]{images/2_4_GHz/cot-node8-student_2017-05-16_chan1_image.png}
+\end{figure}
+\\
+For the second graph, belonging to cot-node8-student, we had to change the days because we did not receive all the measurements due to (presumed) hardware failure. For this graph, the legend is: 
+\begin{itemize}
+\setlength\itemsep{-0.4em}
+\item Blue: Friday, May 12th	
+\item Green: Saturday, May 13th
+\item Red: Sunday, May 14th
+\item Cyan: Monday, May 15th
+\item Magenta: Tuesday, May 16th
+\end{itemize}
+Aside from this, for the 2.4GHz band another thing that kept recurring daily was the perceived business of the spectrum on channel 1 on both cot-node3-student and cot-node9-student, as seen in the following graphs:
+\begin{figure}[h!]
+    \centering
+    \textbf{CoT-node3-student, channel 1, 15/05 - 22/05}\par\medskip
+	\includegraphics[scale=0.5]{images/2_4_GHz/cot-node3-student_2017-05-22_chan1_image.png}
+\end{figure}
+\begin{figure}[h!]
+    \centering
+    \textbf{CoT-node9-student, channel 1, 15/05 - 22/05}\par\medskip
+	\includegraphics[scale=0.5]{images/2_4_GHz/cot-node9-student_2017-05-22_chan1_image.png}
+\end{figure}
+and we have included the non pattern-y graphs in the zip file that is included // they are available on demand? TODO) 
+<Nodes that have recurring patterns>
 \subsection{Calm and busy channels}
-1 busy, 6, 11 calm
+As shown in the previous two graphs, channel 1 is usually the most busy channel in the 2.4 GHz band, with channels 6 and 11 trailing far behind.
+The nodes that saw the most traffic on channel 1 were cot-node3-student and cot-node9-student (see graphs previous subsection) but the traffic they saw on channel 6 and channel 11 is not particulary busy compared to the traffic seen on the other nodes, as we can see from these graphs:\\
+\begin{figure}[h!]
+    \centering
+    \textbf{CoT-node3-student, channel 6, 15/05 - 22/05}\par\medskip
+	\includegraphics[scale=0.5]{images/2_4_GHz/cot-node3-student_2017-05-22_chan6_image.png}
+\end{figure}
+\begin{figure}[h!]
+    \centering
+    \textbf{CoT-node9-student, channel 11, 15/05 - 22/05}\par\medskip
+	\includegraphics[scale=0.5]{images/2_4_GHz/cot-node9-student_2017-05-22_chan11_image.png}
+\end{figure}
+\newpage
+From the data gathered on the indoors-nodes (i.e.
+all but node1) we gathered that the 2.4 GHz band's channels 6 and 11 should be usable for these nodes given the amount of traffic we saw on it.
+The busiest spectrum we saw for the indoor nodes went up to about -75dBm (on cot-node12-student and cot-node3-student).
+When we compare this to the outdoors node, node1, we see that it's spectrum is also around that value, though slightly higher.
+For this graph, another legend is used because this node also experienced a (presumed) hardware failure.
+\begin{itemize}
+\setlength\itemsep{-0.4em}
+\item Blue: Saturday, May 13th
+\item Green: Sunday, May 14th
+\item Red: Monday, May 15th
+\item Cyan: Tuesday, May 16th
+\item Magenta: Wednesday, May 17th
+\end{itemize}
+\begin{figure}[h!]
+    \centering
+    \textbf{node1, channel 6, 13/05 - 17/05}\par\medskip
+	\includegraphics[scale=0.5]{images/2_4_GHz/node1_2017-05-17_chan6_image.png}
+\end{figure}
+\begin{figure}[h!]
+    \centering
+    \textbf{node1, channel 11, 13/05 - 17/05}\par\medskip
+	\includegraphics[scale=0.5]{images/2_4_GHz/node1_2017-05-17_chan11_image.png}
+\end{figure}
 \subsection{2.4 GHz vs 5 GHz}
+niks van de 5ghz bevat iets zinnigs (op node3 / chan36 na) - nodes te ver weg ?
+>> still running the graph generator for 5GHz to compare lol
 \subsection{Validation}
-scans met / zonder sturende laptop - stukje had gij al
-dagelijkse patronen
-outside / inside t zelfde +-
+In terms of data validation, we have three parts that we looked at that give us confidence that our data would be valid.
+scans met / zonder sturende laptop - stukje had gij al TODO
 
+A second point is the discovering of daily patterns where we expected them.
+This includes the pattern of daily activity seen in the node at the Middelheim campus as well as seeing a decrease in activity on holidays such as May 1st TODO, we saw this on 12..
+The third and final point that enforces our trust in the validity of the data we collected and thus also our findings is the fact that the outside node's measurements do not vary wildly from the indoor measurements: it's channel 6 and channel 11 activity is lower than it's channel 1 activity, which is the busiest channel.
+There is more jitter on the outdoor node's results which is most likely due to it being outside.
+Our measurements for channel 1 on the outside node can be found below:\\
+\begin{figure}[h!]
+    \centering
+    \textbf{node1, channel 1, 13/05 - 17/05}\par\medskip
+	\includegraphics[scale=0.5]{images/2_4_GHz/node1_2017-05-17_chan1_image.png}
+\end{figure} \newpage
 \section{Conclusions}
-The conclusion goes here.
+So, how bad is it, in general?
+That was the question to which we sought to find an answer in this project.
+From the results, it seems clear that the 5 GHz band near the nodes is mostly unused and the 2.4 GHz band could be used if set to chan6 or chan11 TODO
 

### 2017-05-25 cot/process_scans/make_images.sh

@@ -1,11 +1,22 @@
 #!/bin/bash
-#2.4Ghz --> 0
-#5  Ghz --> 1
-type=0
+#2.4Ghz --> type =0
+# type=0
+# start=1
+# end=11
+# diff=5
+
+#5  Ghz --> type = 1
+type=1
+start=36
+end=64
+diff=4
 
 for i in cot-node3-student cot-node4-student cot-node7-student cot-node9-student cot-node10-student cot-node12-student ; do
-  for j in 1 6 11 ; do
+  j=$start
+  while [ $j -le $end ]
+  do
     python plot_week_smoothings.py ~/Desktop/UA/AdvNetwLab/current_backup/$i/outputFLUSH/ 2017-05-15,2017-05-16,2017-05-17,2017-05-18,2017-05-19,2017-05-20,2017-05-21,2017-05-22 $type 5 $j 2017-05-14 2017-05-23 $i
+    j=$[$j+$diff]
   done
 done
 
@@ -15,22 +26,41 @@ done
 #For cot-node9-student          : 9999-99-99_99-99
 #For cot-node12-student         : 9999-99-99_99-99
 #For cot-node10-student         : 2017-05-11_11-52 <- enige die 0 is vanwege reboot dus resultaat is nog ok
-
 i='cot-node6-student'
-#For cot-node6-student          : 2017-05-22_21-31 <- 12 -> 20 is doable with 11 and 21 to smoothe
-# python plot_week_smoothings.py ~/Desktop/UA/AdvNetwLab/current_backup/$i/outputFLUSH/ 2017-05-12,2017-05-13,2017-05-14,2017-05-15,2017-05-16,2017-05-17,2017-05-18,2017-05-19,2017-05-20 $type 100 0 2017-05-11 2017-05-21 $i
+j=$start
+while [ $j -le $end ]
+do
+  #For cot-node6-student          : 2017-05-22_21-31 <- 12 -> 20 is doable with 11 and 21 to smoothe
+  python plot_week_smoothings.py ~/Desktop/UA/AdvNetwLab/current_backup/$i/outputFLUSH/ 2017-05-12,2017-05-13,2017-05-14,2017-05-15,2017-05-16,2017-05-17,2017-05-18,2017-05-19,2017-05-20 $type 5 $j 2017-05-11 2017-05-21 $i
+  j=$[$j+$diff]
+done
 i='node1'
[…]

### 2017-05-25 cot/process_scans/plot_week_smoothings_5min.py
Added some findings about 2.4 GHz. Still waiting for 5 GHz results to turn into graphs though for the most part those graphs do not contain anything interesting.
Former-commit-id: 114f45c8fe22aa9467b02a13812f473e950d729b [formerly b3aef06c8a014913253c9e3477ec7219164747fd [formerly 29f6a80152820d93548419c504b862c4f143f4f0] [formerly 29f6a80152820d93548419c504b862c4f143f4f0 [formerly b1f19f6291b5d51ba187a394f93d16ed7697fa33]]]
Former-commit-id: 8f6242760071f9f6f657f106fe5f2020f87eeb5a
Former-commit-id: eaa7c452333b83631144d7e822a2bdb53e2065ed
Former-commit-id: 358f64ecd53f3ec5669b681f135034b773d

@@ -1,0 +1,131 @@
+import itertools
+import subprocess
+import sys
+
+from freq_channelnr_map import map_ as channelnr_map
+
+from plot_rssi import savitzky_golay as savitzky_golay
+
+#rename 's/...$//' *
+
+counter = 0
+maxes = []
+import numpy as np
+import matplotlib.pyplot as plt
+import matplotlib.patches as mpatches
+
+
+def process_point(day, hour, minute):
+    signals = []  # Fill with every nf+rssi of dump
+    file_name = directory + ('/' if not directory.endswith('/') else '') + day + '_' + str(hour).zfill(2) + '-' + str(
+        minute).zfill(2)
+
+    signalstr = subprocess.check_output(
+        ['./fft_get_max_rssi.out', file_name]) if not channel_number else subprocess.check_output(
+        ['./fft_get_max_rssi.out', file_name, channelnr_map[channel_number]])
+    for line in signalstr.splitlines():
+        signals.append(int(line))
+    signals.sort()
+    signals.reverse()
+
+    if (len(signals) == 0):
+        print(file_name)
+        print ""
+    return signals[filter_num]
+
+
+if __name__ == '__main__':
+    if (len(sys.argv) < 9 and len(sys.argv) >= 8) :
+        print("Please also supply node name at the end of command")
+    print(sys.argv[8])
+    colors = ["blue","green","red","cyan","magenta","yellow","black","0.75", "#788599", "#2c8e36", "#7f0043"]
+    if (len(sys.argv) < 6):
+        print "Usage: python plot_rssi.py dump_directory comma_separated_list_of_days first_minute filter_num channel_number [prev_day next_day]\n\
+               Example: To plot for all channels on 2,3,4 May 2017, first dump at 2 minutes past the hour ignoring the top 100 results for every dump: \n\
+               python plot_week_smoothings.py ../output_dir 2017-05-02,2017-05-03,2017-05-04 0 100 0 2017-05-01 2017-05-05 node_name\n\
+               Note that the channel number is required: 0 means all channels\n\
+               To use previous and next day's last and first hour to avoid potentially ugly results at end points.\n\
+               These are only used in calculations and not actually plotted."
+        exit(0)
+
+    if len(sys.argv) == 7:
+        print "Either supply both prev_day and next_day or neither"
+        exit(1)
+
[…]

### 2017-05-26 advnetwpaper/5ghz_part.tex
added 5 ghz findings part in seperate file to add to report.

\subsection{5 GHz}
The data gathered during the 5 GHz measurements proved to be less interesting than the data we gathered in the 2.4 GHz measurements. Most of the channels we measured turned out to be empty or contain very little signal which might have been expected given the number of channels available. For the most part, the channels that did show traffic did not show any significant patterns or recurring phenomena though there are some exceptions. An example such an exception was the cot-node3-student, that showed a lower amount of traffic for some time around noon on both May 17th and May 22nd. The reason for this however is not as apparent as the cause of the pattern in the 2.4 GHz case where we saw an absence of a spike in traffic during the daytime on the schoolnode. It might be that the decrease occurs every couple of days, for example due to a router switching channels temporarily or restarting or the decrease might simply be an anomaly, this would require further investigation. (TODO does this make sense)
\begin{figure}[h!]
    \centering
    \textbf{cot-node3-student, channel 36, 15/05 - 22/05}\par\medskip
	\includegraphics[scale=0.5]{images/5_GHz/cot-node3-student_2017-05-22_chan36_image.pdf}
\end{figure}\\
This graph was generated from our cot-node3-student data discarding the highest 5 (out of roughly 700) signals to combat outliers skewing the graph. The real graph could be a few dBm higher but the result would be no different : the channel is moderately used but there is room for more traffic. On other channels, this node sees barely any traffic in the 5 GHz band. \\
For the 5 GHz band it is worth noting that the measurements we have from our indoor nodes show some difference from those measured by the outdoors node when looking at channels other than channel 36. The outdoor node has at least some traffic on every channel whereas most of the indoors nodes show the bare minimum of traffic for most channels. An example of the outdoors node's traffic on channel 56 versus an indoors node on that same channel is shown below:
\begin{figure}[h!]
    \centering
    \textbf{node1, channel 56, 13/05 - 17/05}\par\medskip
	\includegraphics[scale=0.5]{images/5_GHz/node1_2017-05-17_chan56_image.pdf}
\end{figure}\\
\begin{figure}[h!]
    \centering
    \textbf{cot-node3-student, channel 56, 15/05 - 21/05}\par\medskip
	\includegraphics[scale=0.5]{images/5_GHz/cot-node9-student_2017-05-21_chan56_image.pdf}
\end{figure}\\
Which can be attributed to the outdoor node being outdoors and thus having less trouble picking up 5 GHz signals as they penetrate buildings and walls less easily.\\  \\
There were some other interesting phenomena we observed in the 5 GHz channels, for example the following two graphs paired with a decrease of the signal strength on channel 120 for the same node, as shown on the third graph: \newpage
\begin{figure}[h!]
    \centering
    \textbf{cot-node6-student, channel 56, 12/05 - 20/05}\par\medskip
	\includegraphics[scale=0.5]{images/5_GHz/cot-node6-student_2017-05-20_chan56_image.pdf}
\end{figure}\\
\begin{figure}[h!]
    \centering
    \textbf{cot-node6-student, channel 64, 12/05 - 20/05}\par\medskip
	\includegraphics[scale=0.5]{images/5_GHz/cot-node6-student_2017-05-20_chan64_image.pdf}
\end{figure}\\

\begin{figure}[h!]
    \centering
    \textbf{cot-node6-student, channel 120, 12/05 - 20/05}\par\medskip
	\includegraphics[scale=0.5]{images/5_GHz/cot-node6-student_2017-05-20_chan64_image.pdf}
\end{figure}\\
which seems to show some traffic moving from channel 120 to channel 56 and / or 64 (TODO why doesnt it show on chan60?) on friday May 19th, which lasts all the way through saturday May 20th. We could not confirm when or if the signal ends due to the aforementioned (presumed) hardware failure. This signal still only amounts to an estimated signal strength of about -70 dBm which should still leave enough room for other traffic on the channel and in the spectrum.\\
For some nodes (all but cot-node6-student, cot-node8-student and node1), checking the higher up channels (the channels numbered 100 and above) shows no traffic on any of the nodes for those channels. For the three nodes mentioned, the fact that node1 shows some traffic on these channels is consistent with earlier observations of the outdoor node seeing more traffic on the 5 GHz band than the indoor nodes. For nodes cot-node6-student and cot-node8-student we don't have an explanation for why they have more traffic on these channels than the other nodes. Fact remains that the 100+ channels are relatively empty.
Included below are three graphs for three of our nodes; cot-node3-student, cot-node8-student and node1. Each of these was somewhat busy on the 2.4 GHz band, showed some traffic on the lower 5 GHz channels and demonstrates our findings for the higher channels:
\begin{figure}[h!]
    \centering
    \textbf{cot-node3-student, channel 100, 15/05 - 22/05}\par\medskip
	\includegraphics[scale=0.4]{images/5_GHz/cot-node3-student_2017-05-22_chan100_image.pdf}
\end{figure}\\
\begin{figure}[h!]
    \centering
    \textbf{cot-node12-student, channel 120, 12/05 - 16/05}\par\medskip
	\includegraphics[scale=0.4]{images/5_GHz/cot-node8-student_2017-05-16_chan120_image.pdf}
\end{figure}\\
\begin{figure}[h!]
    \centering
    \textbf{node1, channel 140, 13/05 - 17/05}\par\medskip
	\includegraphics[scale=0.4]{images/5_GHz/node1_2017-05-17_chan140_image.pdf}
\end{figure}\\
\subsection{Validation}
In terms of data validation, we have three parts that we looked at that give us confidence that our data would be valid.
scans met / zonder sturende laptop - stukje had gij al TODO

A second point is the discovering of daily patterns where we expected them. This includes the pattern of daily activity seen in the node at the Middelheim campus as well as seeing a decrease in activity on holidays such as May 1st TODO, we saw this on 12..
The third and final point that enforces our trust in the validity of the data we collected and thus also our findings is the fact that the outside node's measurements do not vary wildly from the indoor measurements: it's channel 6 and channel 11 activity is lower than it's channel 1 activity, which is the busiest channel. There is more jitter on the outdoor node's results which is most likely due to it being outside. Our measurements for channel 1 on the outside node can be found below:\\
\begin{figure}[h!]
    \centering
    \textbf{node1, channel 1, 13/05 - 17/05}\par\medskip
	\includegraphics[scale=0.5]{images/2_4_GHz/node1_2017-05-17_chan1_image.png}
\end{figure} \newpage

### 2017-05-26 advnetwpaper/5ghz_part.tex
Added too much. Removed validation from the part file

@@
 \end{figure}\\
-\subsection{Validation}
-In terms of data validation, we have three parts that we looked at that give us confidence that our data would be valid.
-scans met / zonder sturende laptop - stukje had gij al TODO
-
-A second point is the discovering of daily patterns where we expected them.
-This includes the pattern of daily activity seen in the node at the Middelheim campus as well as seeing a decrease in activity on holidays such as May 1st TODO, we saw this on 12..
-The third and final point that enforces our trust in the validity of the data we collected and thus also our findings is the fact that the outside node's measurements do not vary wildly from the indoor measurements: it's channel 6 and channel 11 activity is lower than it's channel 1 activity, which is the busiest channel.
-There is more jitter on the outdoor node's results which is most likely due to it being outside.
-Our measurements for channel 1 on the outside node can be found below:\\
-\begin{figure}[h!]
-    \centering
-    \textbf{node1, channel 1, 13/05 - 17/05}\par\medskip
-	\includegraphics[scale=0.5]{images/2_4_GHz/node1_2017-05-17_chan1_image.png}
-\end{figure} \newpage

### 2017-05-26 cot/process_scans/plot_rssi.py
Many changes to existing parts
Former-commit-id: ab366296c8fa2e9f9ebeb179c4ecb462e6b8437f [formerly fd752131e251aca65b5e2a0cebafe34de6b80de0 [formerly fd752131e251aca65b5e2a0cebafe34de6b80de0 [formerly d536dcedd3f5ce29963bcb0ebb8a0660cd78198f]]]
Former-commit-id: 79894b4785e8a020dbbef5f0249f900819c882ff
Former-commit-id: d6e1e53877df5a9b50a82a9b8c47e2c6c94ec2e5
Former-commit-id: c2fe77d832cf1caccee57882b93a36726bf0102b

@@ -135,21 +135,24 @@ if __name__ == '__main__':
         next_day = sys.argv[7]
 
     if fix_endpoints:
-        for (minute) in range(first_minute, 60, 2):
-            max_point = process_point(prev_day, 23, minute)
-            maxes.append(max_point)
+        for hour in range(20,24,1):
+            for (minute) in range(first_minute, 60, 2):
+                max_point = process_point(prev_day, hour, minute)
+                maxes.append(max_point)
 
     for (hour, minute) in itertools.product(range(24), range(first_minute, 60, 2)):
         max_point = process_point(day, hour, minute)
         maxes.append(max_point)
 
     if fix_endpoints:
-        for (minute) in range(first_minute, 60, 2):
-            max_point = process_point(next_day, 0, minute)
-            maxes.append(max_point)
+        for hour in range(0,4,1):
+
+            for (minute) in range(first_minute, 60, 2):
+                max_point = process_point(next_day, hour, minute)
+                maxes.append(max_point)
 
     # Endpoints fix has 2 hours of additional data
-    x = np.linspace(0, 24, 720) if not fix_endpoints else np.linspace(-1, 25, 780)
+    x = np.linspace(0, 24, 720) if not fix_endpoints else np.linspace(-4, 28, 960)
 
     y = np.asarray(maxes)
 
@@ -162,8 +165,8 @@ if __name__ == '__main__':
     plt.xticks(np.arange(0, 25, 2))
     # Cut off endpoints fix; only show current day
     plt.xlim([0, 24])
-    plt.xlabel('Hours')
-    plt.ylabel('Received signal strength (dBm)')
+    plt.xlabel('Time of day (hours)')
+    plt.ylabel('Smoothed received signal strength (dBm)')
 
     raw_label = mpatches.Patch(color='blue', label='Highest signal strength (top %d removed)' % filter_num)
     smooth1_label = mpatches.Patch(color='red', label='Savitzky-Golay filtered, order 3, window size 101')

### 2017-05-26 cot/process_scans/fft_get_max_rssi.c
Many changes to existing parts

@@ -426,14 +426,16 @@ int main(int argc, char *argv[]) {
     }
 
     for (result = result_list; result; result = result->next) {
-        counter++;
+
         if (argc < 2) {
+            counter++;
             printf("%d\n", result->sample.ath10k.header.noise + result->sample.ath10k.header.rssi);
             if ((result->sample.ath10k.header.noise + result->sample.ath10k.header.rssi) > max_signal) {
                 max_signal = result->sample.ath10k.header.noise + result->sample.ath10k.header.rssi;
             }
         } else {
             if (result->sample.ath10k.header.freq1 == channel_freq) {
+                counter++;
                 printf("%d\n", result->sample.ath10k.header.noise + result->sample.ath10k.header.rssi);
                 if (result->sample.ath10k.header.noise + result->sample.ath10k.header.rssi > max_signal) {
                     max_signal = result->sample.ath10k.header.noise + result->sample.ath10k.header.rssi;
@@ -441,6 +443,7 @@ int main(int argc, char *argv[]) {
             }
         }
     }
+    printf("%d", counter);
     free(fontdir);
     free_scandata();
 

### 2017-05-26 cot/process_scans/plot_week_smoothings.py

@@ -84,6 +84,7 @@ if __name__ == '__main__':
                 max_point = process_point(prev_day, hour, minute)
                 maxes.append(max_point)
 
+
         for (hour, minute) in itertools.product(range(24), range(first_minute, 60, 2)):
             max_point = process_point(day, hour, minute)
             maxes.append(max_point)
@@ -117,13 +118,16 @@ if __name__ == '__main__':
     plt.xticks(np.arange(0, 25, 2))
     # Cut off endpoints fix; only show current day
     plt.xlim([0, 24])
-    plt.xlabel('Time of Day (hours)')
-    plt.ylabel('Smoothed Received Signal Strength (dBm)')
+
+    plt.xlabel('Time of day (hours)')
+    plt.ylabel('Smoothed received signal strength (dBm)')
 
     #Additional 15dBm on top for legend
-    plt.ylim([plt.gca().get_ylim()[0],plt.gca().get_ylim()[1]+15])
+    #plt.ylim([plt.gca().get_ylim()[0],plt.gca().get_ylim()[1]+15])
+
 
     #plt.legend(handles=label_list,prop={'size':5})
+
 
     plt.savefig("./images/channel_specific_2.4/"+sys.argv[8]+"_"+day[0]+"_"+"chan"+str(channel_number)+"_image.png", bbox_inches='tight')
     plt.savefig("./images/channel_specific_2.4/"+sys.argv[8]+"_"+day[0]+"_"+"chan"+str(channel_number)+"_image.pdf", bbox_inches='tight')

### 2017-05-26 advnetwpaper/paper_no_ieeetrans.tex
Many changes to existing parts
Former-commit-id: d536dcedd3f5ce29963bcb0ebb8a0660cd78198f

--- 2017-05-25 advnetwpaper/paper_no_ieeetrans.tex
@@
 % http://www.ctan.org/pkg/algorithmicx
+
+\usepackage[font=small]{caption}
+\makeatletter
+\renewcommand{\fnum@figure}{Fig. \thefigure}
+\makeatother
+\usepackage{wrapfig}
 
@@
 \author{
-	Jakob Struye (s.......)\\ 
-    Thomas Hendriks (s0139858)\\
+	Jakob Struye (20120612)\\ 
+    Thomas Hendriks (20139858)\\
 }
@@
 Most of the radio spectrum requires a license to operate on.
-Devices emitting Radio Frequency (RF) energy without a license are limited to using one of the unlicensed radio bands, known as the Industrial, Scientific and Medical (ISM) radio bands.
+Devices emitting Radio Frequency (RF) energy without a license often use one of the unlicensed Industrial, Scientific and Medical (ISM) radio bands.
 The most well-known of these bands is the 100MHz wide band centered around 2.45GHz, often called the 2.4GHz band.
 Many consumer devices such as wireless mice and keyboards, bluetooth headphones and Radio Controlled (RC) cars communicate in the 2.4GHz band.
-Furthermore most IEEE 802.11 (Wi-Fi) routers operate in the 2.4GHz band, along with the 5GHz band (which actually ranges from 5.725GHz to 5.875GHz).
+Furthermore most IEEE 802.11 (Wi-Fi) routers operate in the 2.4GHz band.
+In addition 802.11ac routers operating in the 5GHz range often use the 5GHz ISM band (which actually ranges from 5.725GHz to 5.875GHz).
+Some other bands in the 5GHz are also allowed for 802.11ac, although more restrictions, such as indoors-use only, apply there~\cite{50ism}.
 These ISM bands were originally intended and still used for devices emitting RF energy for purposes other than communication.
@@
 \\ \\
-To measure this interference we use the City of Things (CoT) network, operated by research institute imec.
+To measure this interference we use the City of Things (CoT) network, operated by research institute imec\footnote{\url{https://www.imec-int.com/en/home}}.
 This network consists of hundreds of sensors and wireless gateways positioned around the city of Antwerp.
-On 11 of these gateways, we sampled activity on the 2.4GHz and 5GHz bands every 2 minutes for a little over 10 days, to ensure that we would have at least one full week of usable data for each node.
-The main motivation of this paper was in fact to investigate whether interference around the gateways is significant enough to pose a problem for the network.
+On 11 of these gateways, we sampled activity on the 2.4GHz and 5GHz bands every 2 minutes for 12 days, including a Monday-to-Sunday week.
+The main motivation of this experiment is to investigate whether interference around the gateways is significant enough to pose a problem for the network.
 Next to the general magnitude of the interference, we look for patterns in daily and weekly variation of the interference.
@@
 This magnitude incidates  how the power within the channel was divided across the bins.
-We provide an overview of how to interpret these values in \mbox{\hyperref[sec:interpret]{Interpreting spectral scans}}.
+We provide an overview of how to interpret these values in section \ref{sec:interpret}.
 \\
@@
 When providing \textit{iw scan} with a list of all supported frequencies, this closely emulates the \textit{chanscan} behaviour.
-\label{sec:interpret}
-\subsection{Interpreting spectral scans}
+\subsection{Interpreting spectral scans} \label{sec:interpret}
 Before going into the interpretation of this specific data, we provide a quick overview of the dBm unit, often used to express power ratios.
@@
 For each increase of 3dBm, the power in mW doubles, and for each increase of 10dBm, the power is multiplied by 10.
-Milliwatt-to-dBm conversion is calculated using the formula\\ \\ $dBm = 10 log_{10}(\dfrac{x}{\SI{1}{\milli\watt}})$ with $x$ the power in mW.\\ \\ A negative dBm power simply means the power is less than \SI{1}{\milli\watt}.
+Milliwatt-to-dBm conversion is calculated using the formula\\ \\ $dBm = 10 log_{10}(\dfrac{x}{\SI{1}{\milli\watt}})$, with $x$ the power in mW.\\ \\ A negative dBm power simply means the power is less than \SI{1}{\milli\watt}.
 By using dBm, very small or large power values are often avoided.
@@
 For each channel scan, the driver reports the noise floor and RSSI value.
-The noise floor value is hard-coded for every channel and ranges between -95dBm and -106dBm.
+The noise floor value is a hard-coded estimation for every channel and ranges between -94dBm and -108dBm.
 This value is the expected power of all noise in a channel combined.
+This includes noise sources such as thermal noise and cosmic noise.
 The chipset can only receive a singal whose power exceeds this noise floor.
@@
 Atheros chips follow a rather simple formula: subtracting the noise floor from the current signal strength in dBm results in the RSSI.
-These two values are enough to estimate the signal strength in a channel: a noise floor of -96dBm and an RSSI of 40 indicate a received signal of -56dBm.
-TODO SOURCE ?
-(for example :http://www.n-cg.net/ncgpdf/WiFi\_SignalValues.pdf see conversion for atheros) \\
+These two values are enough to estimate the signal strength in a channel: a noise floor of -96dBm and an RSSI of 40 indicate a received signal of -56dBm.~\cite{rssi-calc}.
+The RSSI reported by Atheros is often limited to 60.
+As RSSI was originally intended to estimate signal quality within the driver, values above 60 were not interesting; -36dBm (assuming a noise floor of -96dBm) already indicates excellent reception.
+It appears however that in this spectral scan mode, the hardware does return RSSI values of over 60.
+\\
 The per-bin magnitudes indicate how power is divided within the channel.
@@
 \\ \\ As $log(a*b) = log(a) + log(b)$, we add this coefficient to the total logarithmic power to calculate each share's power:\\ \\ $nf + RSSI + c_{log}(i)$.\\ \\ Noting that $log(a/b) = log(a) - log(b)$, writing the formula using only the parameters returned by the spectral scan results in\\ \\ $nf + RSSI  + 10 log_{10}(b(i)^2) - 10 log_{10}(\sum\limits_{j=1}^{n}{b(j)^2})$ \\ \\
-The earliest reference to this formula being used with the FOSS Atheros drivers was in an email on the ath9k-devel mailing list by Zefir Kurtisi (https://www.spinics.net/lists/linux-wireless/msg101011.html TODO cite), in reply to the Request For Comments (RFC) on spectral scan support by Simon Wunderlich.
+The earliest reference to this formula being used with the FOSS Atheros drivers was in an email on the ath9k-devel mailing list by Zefir Kurtisi~\cite{formula}, in reply to the Request For Comments (RFC) on spectral scan support by Simon W{\"u}nderlich.
 As far as we know neither this formula nor the calculation of its parameters were ever confirmed to be correct by Qualcomm Atheros.
@@
 \section{Measurements}
+
 At first we measured the spectrum on every channel in both the 2.4GHz and the 5GHz band every 10 minutes.
-Looking at graphic interpretations of the resulting data it was clear that a shorter time between scans would be more effective at showing us the state of the spectrum because recurring spikes were missed on some days despite their presence being clear in subsequent smaller tests.
-\\< images waarop da duidelijk is >\\
-Another point that was raised following the first set of graphic interpretations was that of outlier detection and elimination, occasionally there were big spikes in the received signal strength which are most likely caused by interference.
-Playing around with some parameters, we reduced the time between scans from 10 minutes to 2 minutes and discard the 100 largest readings for every sample (when looking at all the channels).
-This limit of 2 minutes was the minimum we could achieve while guaranteeing a scan would succeed and the number of samples to discard was chosen based on some experimentation.
-Using an adapted version of Simon Wünderlich's fft$\_$eval code, Python's Matplotlib library and Savitzky-Golay (source?
-TODO) filtering we graphically represent our data points per node.
-For CoT-node12 (location : UAntwerpen campus Middelheim, indoors), looking at the 2.4 GHz band on every channel from monday May 15th to monday May 22nd, we got the following graph:\\
-\includegraphics[scale=0.7]{images/window_361_2.4/2017-05-22cot-node12-student_image.pdf}\\
-Each line/color represents a day, using this representation we tried to find recurring patterns in the measurements for every node separately.
-When looking at the complete set of channels, patterns were not as clear as we'd hoped.\\
-Given this and the fact that a large amount of static on one channel would make it seem like the entire spectrum is full we switched to drawing the graphs per non-overlapping channel and observing the results in that way.
-This lead to following graph on which we can clearly see that on weekdays there is substantially more traffic during the daytime than on the weekend (which is as expected for a university campus):\\
-\includegraphics[scale=0.7]{images/2017-05-22cot-node12-student_image.pdf}\\
+The visualizations of the gathered data using Simon W{\"u}nderlich's FFT\_eval tool\footnote{https://github.com/simonwunderlich/FFT\_eval} showed that a shorter time between scans would be useful.
+There were often large variations between two subsequent samples, showing a use for a shorter interval.
+Additionally we noticed some extreme outliers.
+Occasionally a received signal strength of 129dBm or 8GW was reported.
+As a steady 8GW supply would require over 35 million solar panels~\cite{gigawatt}, so this is clearly an erroneous measurement.
+Although this was a fairly rare phenomenon, it still occurred up to a few times in each node's almost 10 million daily samples.
+To avoid such erroneous samples, we remove the $n$ highest signal strength measurements from every scan.
+The choice of $n$ depends on the sample size.
+For a single-channel 2.4GHz scan, consisting of around 210 samples, we remove the top 5.
+We then consider the maximum signal strength among the remaining samples as the measured strength within the channel.
+\\
+\begin{figure}[ht]
+\begin{minipage}[t]{0.45\textwidth}
+\centering
+\includegraphics[width=\textwidth]{savitzky_0hours}
+\caption{Savitzky-Golay smoothing.}
+\label{savitzky_0}
+\end{minipage}\
+\begin{minipage}[t]{0.45\textwidth}
+\centering
+\includegraphics[width=\textwidth]{savitzky_1hour}
+\caption{Savitzky-Golay smoothing, 1 additional hour of data point for each end.}
+\label{savitzky_1}
+\end{minipage}\
+\end{figure}
+\begin{figure}[htbp]
+\begin{minipage}[t]{0.45\textwidth}
+\centering
+\includegraphics[width=\textwidth]{savitzky_4hours}
+\caption{Savitzky-Golay smoothing, 4 additional hours of data points for each end.}
+\label{savitzky_4}
+\end{minipage}\
 
-\section{Findings}
+\end{figure}
+
+We perform one scan per minute, alternating between 2.4GHz and 5.0GHz.
+As each scan takes at least 25 seconds to complete, we chose not to lower this interval any further.
+This generates around 860MB of raw data per node per day.
+Even when only considering the maximum signal strength per sample, we still need some way to condense this to a more digestible format.
+We experimented with some smoothing algorithms and had the best results with the Savitzky-Golay filter\footnote{\url{https://en.wikipedia.org/wiki/Savitzky\%E2\%80\%93Golay_filter}}.
+A detailed analysis of the filter was considered outside of the scope of this document, but we note that it is a digital filter intended for equally spaced data points that smooths data through convolution.
+We always used polynomials of degree 3.
+The only other parameter is the window size.
+The higher the window size, the smoother the final result.
+Figure \ref{savitzky_0} shows the resulting smoothings for 2 different window sizes.
+The smaller window size reveals some details such as the drop in activity between noon and 1PM (lunch time), while the larger window size is detailed enough to show that daytime is more active than nighttime.
+This filter performs well mainly for the center of the data set.
+The edges of the smoothing are clearly inaccurate, especially the sudden rise at the end of the day.
+This is a known issue with this filter.
+Luckily we usually also have the data of the preceding and following day.
+By starting the smoothing earlier and ending it later, we obtain a better smoothing.
+Figure \ref{savitzky_1} was generated using the data starting at 11PM the day before up to 1AM the day after.
+In figure \ref{savitzky_4} we increased this margin to 4 hours at each end.
+Whiel the 1 extra hour does little to improve the smoothing and even leads to a worse result in one case, the effect disappears completely with the 4 additional hours.
+We use this 4 hour buffer for all smoothings.
+Furthermore we decided to use the larger window size.
+The smoother graphs make it easier on the eye to compare multiple smoothings on one graph.
+Additionally effects such as the drop during lunchtime were often not seen with even tighter window sizes.\\
+\newpage
+\begin{wrapfigure}{r}{0.35\textwidth}
+\centering
+\captionsetup{justification=centering}
+\includegraphics[width=0.25\textwidth]{legend.png}
+\caption{Legend:\\ days of the week}
+\label{legend}
+\end{wrapfigure}
+Using an adapted version of Simon W{\"u}nderlich's FFT$\_$eval code, Python's Matplotlib library, the mathematical formulas mentioned above and a Savitzky-Golay filter with window size 361 we graphically represent our data points per node.
+For one week (Monday 15 May - Sunday 21 May) we plot each day's smoothing, for easy comparison between the days.
+For the 2.4GHz band we perform this separately for the non-overlapping channels 1, 6 and 11.
+Each color in the plot represents a day of the week, using the legend in figure \ref{legend}.
+All other plots use the same legend, although days may be missing.
+
+\subsection{Accuracy of the Measurements} \label{accuracy}
+
+\begin{figure}[ht]
+\begin{minipage}[t]{0.45\textwidth}
+\centering
+\includegraphics[width=\textwidth]{24check.png}
+\caption{2.4GHz spectral scan with active nearby device at 2.437GHz.}
+\label{fig_24}
+\end{minipage}\hfill
+\begin{minipage}[t]{0.45\textwidth}
+\centering
+\includegraphics[width=\textwidth]{24check_nodata.png}
+\caption{2.4GHz baseline spectral scan.}
+\label{fig_24nodata}
+\end{minipage}\
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=0.75\textwidth]{50check.png}
+% where an .eps filename suffix will be assumed under latex,
+% and a .pdf suffix will be assumed for pdflatex; or what has been declared
+% via \DeclareGraphicsExtensions.
+\caption{5.0GHz spectral scan with active nearby device at 5.660GHz}
+\label{fig_50}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=0.75\textwidth]{50check_nodata.png}
+% where an .eps filename suffix will be assumed under latex,
+% and a .pdf suffix will be assumed for pdflatex; or what has been declared
+% via \DeclareGraphicsExtensions.
+\caption{5.0GHz baseline spectral scan.}
+\label{fig_50nodata}
+\end{figure}
+We first confirm that the spectral scan does in fact accurately measure activity per channel.
+To test this, we perform a data transfer with a laptop positioned roughly 50cm from the node.
+We perform this on available 2.4GHz and 5.0GHz networks.
+The 2.4GHz network was using channel 6 (around 2.437GHz) and lead to the spectral scan in figure \ref{fig_24}.
+This scan shows a clear peak centered around channel 6 of around 20MHz wide.
+Figure \ref{fig_24nodata} shows a scan of the same node around the same time, but with our data transfer disabled.
+While the hazy peak around channel 1 is still there, the peak around channel 6 is not visible, indicating it was caused by our data transfer.
+Comparing the two, we see that our data transfer caused an increase in received signal strength of around 40dBm.\\ Our 5.0GHz test was using an access point centered around channel 132 (5.66GHz) and lead to figure \ref{fig_50}.
+Compared to the flat \ref{fig_50nodata} captured with the data transfer disabled, the peak is again very clear.
+Received signal strength is increased by up to 50dBm in this case.
+The scan is detailed enough to indicate that the access point was using a 40MHz wide channel: data is being transferred both in channel 132 (5.66GHz) and channel 136 (5.68Ghz).
+These two tests clearly show that this method is accurate enough for our goals.
+
+\section{Results}
+In this section we present the results of our experiment.
+Due to (presumed) hardware failure, some measurements failed.
+As a result we had to limit and/or shift the timeframe of the presented measurements for some nodes.
+This is indicated below.
 \subsection{Evolution throughout the day}
-\subsection{Evolution throughout the week}
+
 \subsection{Recurring patterns between days} 
 On the different nodes we had varying degrees of success in finding recurring patterns.
-For this report we've only included the nodes on which we did find such patterns (and we have included the non pattern-y graphs in the zip file that is included // they are available on demand?
-TODO)
-\subsection{Calm and busy channels}
-1 busy, 6, 11 calm
+In the 2.4 GHz band, two nodes that clearly showed repeating patterns across several days were node 8 at the Museum Plantin-Moretus and node 12 at the University of Antwerp, Middelheim campus.
+Both of these nodes detected patterns, with node 8 detecting a large increase in spectrum usage in the afternoon and node 12 detecting an increase during the daytime on weekdays.
+There were no peaks during the weekend.
+Node 8 was one of the nodes with failure, so we had to limit its timeframe.
+
+\begin{figure}[!h]
+\begin{minipage}{0.47\textwidth}
+    \centering
+	\includegraphics[width=\textwidth]{images/2_4_GHz/cot-node12-student_2017-05-22_chan1_image}
+    \caption{Node 12, ch. 1, 15/05 - 22/05} \label{node12-1}
+\end{minipage}\hfill
+\begin{minipage}{0.47\textwidth}
+    \centering
+	\includegraphics[width=\textwidth]{images/2_4_GHz/cot-node8-student_2017-05-16_chan1_image}
+    \caption{Node 8, ch. 1, 12/05 - 16/05} \label{node8-1}
+\end{minipage}\hfill
+\end{figure}
+\begin{figure}[!h]
+\begin{minipage}{0.47\textwidth}
+    \centering
+	\includegraphics[width=\textwidth]{images/2_4_GHz/cot-node12-student_2017-05-22_chan6_image}
+    \caption{Node 12, ch. 6, 15/05 - 22/05} \label{node12-6}
+\end{minipage}\hfill
+\begin{minipage}{0.47\textwidth}
+    \centering
+	\includegraphics[width=\textwidth]{images/2_4_GHz/cot-node8-student_2017-05-16_chan6_image}
+    \caption{Node 8, ch. 6, 12/05 - 16/05} \label{node8-6}
+\end{minipage}\hfill
+\end{figure}
+\begin{figure}[!h]
+\begin{minipage}{0.47\textwidth}
+    \centering
+	\includegraphics[width=\textwidth]{images/2_4_GHz/cot-node12-student_2017-05-22_chan11_image}
+    \caption{Node 12, ch. 11, 15/05 - 22/05} \label{node12-11}
+\end{minipage}\hfill
+\begin{minipage}{0.47\textwidth}
+    \centering
+	\includegraphics[width=\textwidth]{images/2_4_GHz/cot-node8-student_2017-05-16_chan11_image}
+    \caption{Node 8, ch. 11, 12/05 - 16/05} \label{node8-11}
+\end{minipage}\hfill
+\end{figure}
+
+The most noticeable peak for node 12 is on channel 1, with the received signal strength during the daytime being up to 20dBm higher than during the night time.
+During the weekend (black and yellow) there is no peak whatsoever.
+This indicates the traffic is most likely caused by Wi-Fi usage at the university, which is closed during the weekend.
+For the other channels there a noticeable peak during the daytime for some but not all weekdays.
+The weekend is still the quietest time of the week.
+Node 8 also shows a peak during the afternoon for channel 1, except on Sunday.
+Surprisingly the museum is open on Sunday, but closed on Monday.
+Of the days with a peak, Monday's peak is the lowest.
+This seems to indicate that while visitors at the museum have some effect on the overall signal strength, the peak has another source, possibly an office nearby.
+\\ \\
+We found no such peaks for any other nodes: spectrum usage was stable throughout the day. In the following section we investigate how active the different channels are.
+
+\subsection{Channel Activity}
+
+\begin{figure}[!h]
+\begin{minipage}{0.47\textwidth}
+    \centering
+	\includegraphics[width=\textwidth]{images/2_4_GHz/cot-node3-student_2017-05-21_chan1_image}
+    \caption{Node 3, ch. 1, 15/05 - 22/05} \label{node3-1}
+\end{minipage}\hfill
+\begin{minipage}{0.47\textwidth}
+    \centering
+	\includegraphics[width=\textwidth]{images/2_4_GHz/cot-node9-student_2017-05-21_chan1_image}
+    \caption{Node 9, ch. 1, 15/05 - 22/05} \label{node9-1}
+
+\end{minipage}\hfill
+\end{figure}
+\begin{figure}[!h]
+\begin{minipage}{0.47\textwidth}
+    \centering
+	\includegraphics[width=\textwidth]{images/2_4_GHz/cot-node3-student_2017-05-21_chan6_image}
+    \caption{Node 3, ch. 6, 15/05 - 22/05} \label{node3-6}
+\end{minipage}\hfill
+\begin{minipage}{0.47\textwidth}
+    \centering
+	\includegraphics[width=\textwidth]{images/2_4_GHz/cot-node9-student_2017-05-21_chan6_image}
+    \caption{Node 9, ch. 6, 15/05 - 22/05} \label{node9-6}
+\end{minipage}\hfill
+\end{figure}
+
+\begin{figure}[!h]
+\begin{minipage}{0.47\textwidth}
+    \centering
+	\includegraphics[width=\textwidth]{images/2_4_GHz/cot-node3-student_2017-05-21_chan11_image}
+    \caption{Node 3, ch. 11, 15/05 - 22/05} \label{node3-11}
+\end{minipage}\hfill
+\begin{minipage}{0.47\textwidth}
+    \centering
+	\includegraphics[width=\textwidth]{images/2_4_GHz/cot-node9-student_2017-05-21_chan11_image}
+    \caption{Node 9, ch. 11, 15/05 - 22/05} \label{node9-11}
+\end{minipage}\hfill
+\end{figure}
+
+As shown in figures \ref{node3-1} to \ref{node9-11}, channel 1 is usually the busist channel in the 2.4 GHz band, with channels 6 and 11 trailing far behind.
+Node 3 and 9, shown in those figures, measure a very busy medium around channel 1.
+However channels 6 and 11 are not particulary busy compared to the traffic seen on the other nodes, as we can see from these graphs:\\
+
+\newpage
+All nodes covered so far were indoors node.
+We had access to one outdoors node, node 1, on which we also performed the experiment.
+On the indoors nodes we noticed that channel 6 and 11 were never any busier than channel 1.
+This is also true for node1: while activity is fairly constant throughout all tested days, channel 1 is again the busiest.
+Note that we again do not have a full week's data for this note.
+Note the remarkable difference between weekdays and the weekend here.
+For channel 11, the weekend is more quiet, however for channel 6 the weekend is the busiest period.
+\begin{figure}[!h]
+\begin{minipage}{0.47\textwidth}
+    \centering
+	\includegraphics[width=\textwidth]{images/2_4_GHz/node1_2017-05-17_chan1_image}
+    \caption{Node 1, ch. 1, 13/05 - 17/05} \label{node1-1}
+\end{minipage}\hfill
+\begin{minipage}{0.47\textwidth}
+    \centering
+	\includegraphics[width=\textwidth]{images/2_4_GHz/node1_2017-05-17_chan6_image}
+    \caption{Node 1, ch. 6, 13/05 - 17/05} \label{node1-6}
+\end{minipage}\hfill
+\end{figure}
+\begin{figure}[h!]
+\begin{minipage}{0.47\textwidth}
+    \centering
+	\includegraphics[width=\textwidth]{images/2_4_GHz/node1_2017-05-17_chan11_image}
+    \caption{Node 1, ch. 11, 13/05 - 17/05} \label{node1-11}
+\end{minipage}\hfill
+\end{figure}
+%\begin{figure}[h!]
+%    \centering
+%    \textbf{node1, channel 11, 13/05 - 17/05}\par\medskip
+%	\includegraphics[scale=0.5]{images/2_4_GHz/node1_2017-05-17_chan11_image}
+%\end{figure}
 \subsection{2.4 GHz vs 5 GHz}
+niks van de 5ghz bevat iets zinnigs (op node3 / chan36 na) - nodes te ver weg ?
+>> still running the graph generator for 5GHz to compare lol
 \subsection{Validation}
-scans met / zonder sturende laptop - stukje had gij al
-dagelijkse patronen
-outside / inside t zelfde +-
+While there is no official confirmation from the chip manufacturer that our interpretation of the spectral scan data is correct, we are confident that our interpretation gives at least a solid estimation of spectrum activity.
+First of all the experiment in section \ref{accuracy} shows that when we add a source of RF activity near the node, this is clearly seen in the spectral scan at exactly the right frequency.
+\\
+Furthermore we discovered daily patterns where we expected them.
+The Middelheim offices are busy during the daytime on weekdays but fairly silent during the weekends and nights.
+While not shown in the plots above, holidays were just as quiet as weekends.
 
 \section{Conclusions}
-The conclusion goes here.
-
+In this report we presented our findings on activity in the RF spectrum in the 2.4GHz and 5.0GHz bands in the city of Antwerp.
+One clear theme in our results is that the 5.0GHz is overall quieter than the 2.4GHz band.
+This is not surprising.
+The 5.0GHz band is a lot wider than the 2.4GHz band, and its signals do not carry as far.
+Furthermore, assuming that a lot of interference originated from Wi-Fi routers, one factor could be that many active routers and other devices do not yet support 5.0GHz.
+In addition, technologies such as Bluetooth, which could also be creating signals, only work on 2.4GHz.\\ \\
+In some environments we notice a daily pattern where the signal strength is about 20dBm higher during the day compared to the night.
+As this was mainly observed on nodes in office spaces, this effect disappeared during the weekend.
+Other nodes show absolutely no difference between daytime and nighttime.
+Often this is when the spectrum is just very quiet.
+One interesting observation is that channels 6 and 11 were never busier than channel 1.
 \bibliographystyle{IEEEtran}
-\bibliography{IEEEabrv,../chapters/bibliography}
+\bibliography{IEEEabrv,./bibliography}
 

### 2017-05-26 advnetwpaper/paper_no_ieeetrans.tex
Added 5GHz part.

@@
 In figure \ref{savitzky_4} we increased this margin to 4 hours at each end.
-Whiel the 1 extra hour does little to improve the smoothing and even leads to a worse result in one case, the effect disappears completely with the 4 additional hours.
+While the 1 extra hour does little to improve the smoothing and even leads to a worse result in one case, the effect disappears completely with the 4 additional hours.
 We use this 4 hour buffer for all smoothings.
@@
 This is indicated below.
-\subsection{Evolution throughout the day}
-
-\subsection{Recurring patterns between days} 
+\subsection{2.4 GHz band}
+\subsubsection{Recurring patterns between days} 
 On the different nodes we had varying degrees of success in finding recurring patterns.
@@
 
-\subsection{Channel Activity}
+\subsubsection{Channel Activity}
 
@@
 
-As shown in figures \ref{node3-1} to \ref{node9-11}, channel 1 is usually the busist channel in the 2.4 GHz band, with channels 6 and 11 trailing far behind.
+As shown in figures \ref{node3-1} to \ref{node9-11}, channel 1 is usually the busiest channel in the 2.4 GHz band, with channels 6 and 11 trailing far behind.
 Node 3 and 9, shown in those figures, measure a very busy medium around channel 1.
@@
 \newpage
-All nodes covered so far were indoors node.
+All nodes covered so far were indoors nodes.
 We had access to one outdoors node, node 1, on which we also performed the experiment.
@@
 This is also true for node1: while activity is fairly constant throughout all tested days, channel 1 is again the busiest.
-Note that we again do not have a full week's data for this note.
+Note that we again do not have a full week's data for this node.
 Note the remarkable difference between weekdays and the weekend here.
@@
 %\end{figure}
-\subsection{2.4 GHz vs 5 GHz}
-niks van de 5ghz bevat iets zinnigs (op node3 / chan36 na) - nodes te ver weg ?
->> still running the graph generator for 5GHz to compare lol
+\newpage
+\subsection{5 GHz band}
+\subsubsection{Recurring patterns between days}
+The data gathered during the 5 GHz measurements proved to be less interesting than the data we gathered in the 2.4 GHz measurements.
+Most of the channels we measured turned out to be empty or contain very little signal which might have been expected given the number of channels available.
+For the most part, the channels that did show traffic did not show any significant patterns or recurring phenomena though there are some exceptions.
+One such an exception was the cot-node3-student, that showed a lower amount of traffic for some time around noon on both May 17th and May 22nd.
+The reason for this however is not as apparent as the cause of the pattern in the 2.4 GHz case where we saw an absence of a spike in traffic during the daytime on the schoolnode.
+It might be that the decrease occurs every couple of days, for example due to a router switching channels temporarily or restarting or the decrease might simply be an anomaly, this would require further investigation.
+\begin{figure}[ht]
+\centering
+\includegraphics[width=0.47\textwidth]{images/5_GHz/cot-node3-student_2017-05-22_chan36_image.pdf}
+% where an .eps filename suffix will be assumed under latex,
+% and a .pdf suffix will be assumed for pdflatex; or what has been declared
+% via \DeclareGraphicsExtensions.
+\caption{Node 3, ch. 56, 15/05 - 22/05}
+\label{cot3_5ghz_pattern_chan36}
+\end{figure}
+This graph was generated from our cot-node3-student data discarding the highest 5 (out of roughly 700) signals to combat outliers skewing the graph.
+The real graph could be a few dBm higher but the result would be no different : the channel is moderately used but there is room for more traffic.
+On other channels, this node sees barely any traffic in the 5 GHz band.
+\\
+For the 5 GHz band it is worth noting that the measurements we have from our indoor nodes show some difference from those measured by the outdoors node when looking at channels other than channel 36.
+The outdoor node has at least some traffic on every channel whereas most of the indoors nodes show the bare minimum of traffic for most channels.
+An example of the outdoors node's traffic on channel 56 versus an indoors node on that same channel is shown below:
+\begin{figure}[!h]
+\begin{minipage}{0.47\textwidth}
+	\centering
+	\includegraphics[width=\textwidth]{images/5_GHz/node1_2017-05-17_chan56_image.pdf}
+	\caption{Node 1, ch. 56, 13/05 - 17/05}
+	\label{node1_5ghz_traffic_chan56}
+\end{minipage}\hfill
+\begin{minipage}{0.47\textwidth}
+    \centering
+	\includegraphics[width=\textwidth]{images/5_GHz/cot-node9-student_2017-05-21_chan56_image.pdf}
+    \caption{Node 9, ch. 56, 15/05 - 21/05} \label{node9_5ghz_traffic_chan56}
+\end{minipage}\hfill
+\end{figure}
+Which can be attributed to the outdoor node being outdoors and thus having less trouble picking up 5 GHz signals as they penetrate buildings and walls less easily.\\  \\
+There were some other interesting phenomena we observed in the 5 GHz channels, for example the following two graphs paired with a decrease of the signal strength on channel 120 for the same node, as shown on the third graph: \newpage
+\begin{figure}[!h]
+\begin{minipage}{0.47\textwidth}
+	\centering
+	\includegraphics[width=\textwidth]{images/5_GHz/cot-node6-student_2017-05-20_chan56_image.pdf}
+	\caption{Node 6, ch. 56, 14/05 - 20/05}
+	\label{node6_5ghz_traffic_chan56}
+\end{minipage}\hfill
+\begin{minipage}{0.47\textwidth}
+	\centering
+	\includegraphics[width=\textwidth]{images/5_GHz/cot-node6-student_2017-05-20_chan64_image.pdf}
+	\caption{Node 6, ch. 64, 14/05 - 20/05}
+	\label{node6_5ghz_traffic_chan64}
+\end{minipage}\hfill
+\end{figure}
+\begin{figure}[ht]
+\centering
+\includegraphics[width=0.47\textwidth]{images/5_GHz/cot-node6-student_2017-05-20_chan120_image.pdf}
+% where an .eps filename suffix will be assumed under latex,
+% and a .pdf suffix will be assumed for pdflatex; or what has been declared
+% via \DeclareGraphicsExtensions.
+\caption{Node 6, ch. 120, 14/05 - 20/05}
+\label{node6_5ghz_traffic_chan120}
+\end{figure}
+which seems to show some traffic moving from channel 120 to channel 56 and / or 64 (TODO why doesnt it show on chan60?) on friday May 19th, which lasts all the way through saturday May 20th.
+We could not confirm when or if the signal ends due to the aforementioned (presumed) hardware failure.
+This signal still only amounts to an estimated signal strength of about -70 dBm which should still leave enough room for other traffic on the channel and in the spectrum.\\
+\subsubsection{Channel Activity}
+For some nodes (all but cot-node6-student, cot-node8-student and node1), checking the higher up channels (the channels numbered 100 and above) shows no traffic on any of the nodes for those channels.
+For the three nodes mentioned, the fact that node1 shows some traffic on these channels is consistent with earlier observations of the outdoor node seeing more traffic on the 5 GHz band than the indoor nodes.
+For nodes cot-node6-student and cot-node8-student we don't have an explanation for why they have more traffic on these channels than the other nodes.
+Fact remains that the 100+ channels are relatively empty.
+Included below are three graphs for three of our nodes; cot-node3-student, cot-node8-student and node1.
+Each of these was somewhat busy on the 2.4 GHz band, showed some traffic on the lower 5 GHz channels and demonstrates our findings for the higher channels:
+\begin{figure}[!h]
+\begin{minipage}{0.47\textwidth}
+	\centering
+	\includegraphics[width=\textwidth]{images/5_GHz/cot-node3-student_2017-05-22_chan100_image.pdf}
+	\caption{Node 8, ch. 120, 15/05 - 22/05}
+	\label{node3_5ghz_traffic_chan100}
+\end{minipage}\hfill
+\begin{minipage}{0.47\textwidth}
+	\centering
+	\includegraphics[width=\textwidth]{images/5_GHz/cot-node8-student_2017-05-16_chan120_image.pdf}
+	\caption{Node 8, ch. 120, 12/05 - 16/05}
+	\label{node8_5ghz_traffic_chan120}
+\end{minipage}\hfill
+\end{figure}
+\begin{figure}[ht]
+\centering
+\includegraphics[width=0.47\textwidth]{images/5_GHz/node1_2017-05-17_chan140_image.pdf}
+% where an .eps filename suffix will be assumed under latex,
+% and a .pdf suffix will be assumed for pdflatex; or what has been declared
+% via \DeclareGraphicsExtensions.
+\caption{Node 1, ch. 140, 13/05 - 17/05}
+\label{node1_5ghz_traffic_chan140}
+\end{figure}
+For the lower channels, i.e.
+channels between 36 and 64, more traffic is observed : mainly on channel 36 though in some cases channel 56 is also used though often not enough to say that the spectrum is fully occupied.
+We found no clear indication that channels other than 36 and 56 were being heavily used in this range (and 64 in the case of node6 TODO incorporate this?).
+Examples of this usage can be seen in the following graphs.
+Note that the intensity of the usage does vary between nodes, for channel 40 we have included an illustrative graph, all nodes display similar results for this channel.
+n3 36 n1 36 n12 56
+\begin{figure}[!h]
+\begin{minipage}{0.47\textwidth}
+	\centering
+	\includegraphics[width=\textwidth]{images/5_GHz/cot-node3-student_2017-05-22_chan36_image.pdf}
+	\caption{Node 3, ch. 36, 15/05 - 22/05}
+	\label{node3_ch36_usage}
+\end{minipage}\hfill
+\begin{minipage}{0.47\textwidth}
+	\centering
+	\includegraphics[width=\textwidth]{images/5_GHz/node1_2017-05-17_chan36_image.pdf}
+	\caption{Node 1, ch. 36, 13/05 - 17/05}
+	\label{node1_ch36_usage}
+\end{minipage}\hfill
+\end{figure}
+\begin{figure}[!h]
+\begin{minipage}{0.47\textwidth}
+	\centering
+	\includegraphics[width=\textwidth]{images/5_GHz/cot-node12-student_2017-05-21_chan56_image.pdf}
+	\caption{Node 12, ch. 56, 15/05 - 22/05}
+	\label{node12_ch56_usage}
+\end{minipage}\hfill
+\begin{minipage}{0.47\textwidth}
+	\centering
+	\includegraphics[width=\textwidth]{images/5_GHz/cot-node10-student_2017-05-22_chan40_image.pdf}
+	\caption{Node 10, ch. 40, 15/05 - 22/05}
+	\label{node10_ch40_usage}
+\end{minipage}\hfill
+\end{figure}
 \subsection{Validation}
@@
 While not shown in the plots above, holidays were just as quiet as weekends.
-
 \section{Conclusions}
@@
 Furthermore, assuming that a lot of interference originated from Wi-Fi routers, one factor could be that many active routers and other devices do not yet support 5.0GHz.
-In addition, technologies such as Bluetooth, which could also be creating signals, only work on 2.4GHz.\\ \\
+In addition, technologies such as Bluetooth, which could also be creating signals, only affects 2.4GHz.\\ \\
 In some environments we notice a daily pattern where the signal strength is about 20dBm higher during the day compared to the night.
@@
 One interesting observation is that channels 6 and 11 were never busier than channel 1.
+\\
+Furthermore, we found nothing indicating that the wireless spectrum would be unusable for a City of Things network if operating in the 5 GHz band, especially if taking care during channel selection to select a mostly unused channel.
+When operating on the 2.4 GHz band we observed more traffic that would, in some places and on some channels, prevent communication between nodes so in this case it would be especially important to choose a channel that is not already completely occupied but our findings show that it should be possible.
 \bibliographystyle{IEEEtran}

### 2017-05-26 advnetwpaper/paper_no_ieeetrans.tex
Changes to 5.0GHz part
Former-commit-id: cb3496b7bc55ad2184fa9706e9cb6fa91e40aff4 [formerly 8e0b2182f821195e61767d8b6bae7e59598de252 [formerly 8e0b2182f821195e61767d8b6bae7e59598de252 [formerly 9d85fe00ea5488caad88d735d27e541eff76949e]]]
Former-commit-id: be7a14ef9462925f0991737b6d6c14c8144a4698
Former-commit-id: 593dd31e0b4c51544bdff05dc6feb394b3405c2d
Former-commit-id: d835cbede24c08497359dae696160b89464d8630

--- 2017-05-26 advnetwpaper/paper_no_ieeetrans.tex
@@
 \end{minipage}\
-
 \end{figure}
@@
 Additionally effects such as the drop during lunchtime were often not seen with even tighter window sizes.\\
-\newpage
+\newpage %Required to fix wrapfigure formatting
 \begin{wrapfigure}{r}{0.35\textwidth}
@@
 
-\begin{figure}[!h]
+\begin{figure}
 \begin{minipage}{0.47\textwidth}
@@
 \end{minipage}\hfill
-\end{figure}
-\begin{figure}[!h]
+
 \begin{minipage}{0.47\textwidth}
@@
 \end{minipage}\hfill
-\end{figure}
-\begin{figure}[!h]
 \begin{minipage}{0.47\textwidth}
@@
 
-The most noticeable peak for node 12 is on channel 1, with the received signal strength during the daytime being up to 20dBm higher than during the night time.
+The most noticeable peak for node 12 is on channel 1 in figure \ref{node12-1}, with the received signal strength during the daytime being up to 20dBm higher than during the night time.
 During the weekend (black and yellow) there is no peak whatsoever.
@@
 The weekend is still the quietest time of the week.
-Node 8 also shows a peak during the afternoon for channel 1, except on Sunday.
+Node 8 also shows a peak during the afternoon for channel 1 in \ref{node8-1}, except on Sunday.
 Surprisingly the museum is open on Sunday, but closed on Monday.
@@
 
-\subsection{Channel Activity}
+\subsubsection{Channel Activity}
 
-\begin{figure}[!h]
+\begin{figure}
 \begin{minipage}{0.47\textwidth}
@@
 \end{minipage}\hfill
-\end{figure}
-
-\begin{figure}[!h]
 \begin{minipage}{0.47\textwidth}
@@
 
-\newpage
-All nodes covered so far were indoors node.
+All nodes covered so far were indoors nodes.
 We had access to one outdoors node, node 1, on which we also performed the experiment.
@@
 This is also true for node1: while activity is fairly constant throughout all tested days, channel 1 is again the busiest.
-Note that we again do not have a full week's data for this note.
+Note that we again do not have a full week's data for this node.
 Note the remarkable difference between weekdays and the weekend here.
 For channel 11, the weekend is more quiet, however for channel 6 the weekend is the busiest period.
-\begin{figure}[!h]
+
+\subsection{5 GHz band}
+\subsubsection{Recurring patterns between days}
+\begin{figure}
 \begin{minipage}{0.47\textwidth}
@@
 \end{minipage}\hfill
-\end{figure}
-\begin{figure}[h!]
 \begin{minipage}{0.47\textwidth}
@@
 \end{minipage}\hfill
+\begin{minipage}{0.47\textwidth}
+\centering
+\includegraphics[width=\textwidth]{images/5_GHz/cot-node3-student_2017-05-22_chan36_image.pdf}
+\caption{Node 3, ch. 56, 15/05 - 22/05}
+\label{cot3_5ghz_pattern_chan36}
+\end{minipage}\hfill
 \end{figure}
-%\begin{figure}[h!]
-%    \centering
-%    \textbf{node1, channel 11, 13/05 - 17/05}\par\medskip
-%	\includegraphics[scale=0.5]{images/2_4_GHz/node1_2017-05-17_chan11_image}
-%\end{figure}
-\subsection{2.4 GHz vs 5 GHz}
-niks van de 5ghz bevat iets zinnigs (op node3 / chan36 na) - nodes te ver weg ?
->> still running the graph generator for 5GHz to compare lol
+The data gathered during the 5 GHz measurements proved to be less interesting than the data we gathered in the 2.4 GHz measurements.
+Most of the channels we measured turned out to be silent or contain very little signal.
+This is partly to be expected.
+Many common communication devices only use the 2.4GHz, and many routers in use do not yet support the 5.0GHz band.
+Furthermore, any communication that does occur in this band can be more spread out as the band is a lot wider than the 2.4GHz band.
+For the most part, the channels that did show traffic did not show any significant patterns or recurring phenomena.
+In most cases the received signal strength was very stable on all channels.
+We did notice some remarkable behaviour though.
+For example, node 3 measured a lower signal strength for some time around noon on both May 17\textsuperscript{th} and May 22\textsuperscript{nd} on channel 56, in figure \ref{cot3_5ghz_pattern_chan36}.
+The reason for this however is not as clear as with the daily peaks experienced in the 2.4GHz It might be that the decrease occurs every couple of days, for example due to a router switching channels temporarily.
+It could also simply be an anomaly.
+
+For this node, channel 56 was the only somewhat active channel. All other channels saw barely any traffic in the 5.0 GHz band. \\\\
+It is worth noting that, apart from channel 36, there is some difference in the stability of the received signal between the outdoor node and the indoor nodes.
+The outdoor node has at least some activity on every channel whereas most of the indoors nodes show the bare minimum of activity for most channels.
+An example of the outdoors node's activity on channel 56 versus an indoors node on that same channel is shown in figures \ref{node1_5ghz_traffic_chan56} and \ref{node9_5ghz_traffic_chan56}.
+There could be some outside source of slight intermittent interference.
+As we only had access to one outdoors node we cannot determine whether this generally occurs with outside nodes, or is specific to this one.
+
+\begin{figure}[ht]
+\begin{minipage}{0.47\textwidth}
+	\centering
+	\includegraphics[width=\textwidth]{images/5_GHz/node1_2017-05-17_chan56_image.pdf}
+	\caption{Node 1, ch. 56, 13/05 - 17/05}
+	\label{node1_5ghz_traffic_chan56}
+\end{minipage}\hfill
+\begin{minipage}{0.47\textwidth}
+    \centering
+	\includegraphics[width=\textwidth]{images/5_GHz/cot-node9-student_2017-05-21_chan56_image.pdf}
+    \caption{Node 9, ch. 56, 15/05 - 21/05} \label{node9_5ghz_traffic_chan56}
+\end{minipage}\hfill
+\end{figure}
+
+There were some other interesting phenomena we observed in the 5 GHz channels.
+For example, node 6 shows a sudden drop in activity on both channels 56 and 64, coinciding with an increase in traffic on channels 112 and 120, as seen in figures \ref{node6_5ghz_traffic_chan56} up to \ref{node6_5ghz_traffic_chan120}.
+Looking at a random spectral scan from before and after the shift, we indeed see the same behaviour.
+There was clear activity in channels 56 and 64 before the shift, seen in figure \ref{cot6_50_early}.
+\begin{figure}[ht]
+\begin{minipage}{0.47\textwidth}
+	\centering
+	\includegraphics[width=\textwidth]{images/5_GHz/cot-node6-student_2017-05-20_chan56_image.pdf}
+	\caption{Node 6, ch. 56, 14/05 - 20/05}
+	\label{node6_5ghz_traffic_chan56}
+\end{minipage}\hfill
+\begin{minipage}{0.47\textwidth}
+	\centering
+	\includegraphics[width=\textwidth]{images/5_GHz/cot-node6-student_2017-05-20_chan64_image.pdf}
+	\caption{Node 6, ch. 64, 14/05 - 20/05}
+	\label{node6_5ghz_traffic_chan64}
+\end{minipage}\hfill
+\begin{minipage}{0.47\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{images/5_GHz/cot-node6-student_2017-05-20_chan112_image.pdf}
+    \caption{Node 6, ch. 112, 14/05 - 20/05}
+    \label{node6_5ghz_traffic_chan112}
+\end{minipage}\hfill
+\begin{minipage}{0.47\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{images/5_GHz/cot-node6-student_2017-05-20_chan120_image.pdf}
+    \caption{Node 6, ch. 120, 14/05 - 20/05}
+    \label{node6_5ghz_traffic_chan120}
+\end{minipage}\hfill
+\end{figure}
+ While there is some very slight activity in channel 60, so few samples contained any trace of it that it was filtered out.
+Later on, figure \ref{cot6_50_late} shows that this traffic signature has moved to channels 112 and 120, now with the channel in between showing no traffic at all.
+We are not sure what device exactly causes this, but it is clear that it was at some point switched to another channel.
+This shows that the activity in a channel can change at a moment's notice, and indicates that systematic or even constant monitoring of the spectrum is not without merit.
+We do note that the received high signal is still only -70dBm, meaning there is still room for communication.
+Furthermore the spectral scan samples show that the those channels are still silent for a large fraction of the time.
+
+\begin{figure}
+\begin{minipage}[t]{0.47\textwidth}
+\centering
+\includegraphics[width=\textwidth]{cot6_50_early.png}
+\caption{Node 6 channels 56-64, before shift}
+\label{cot6_50_early}
+\end{minipage}\hfill
+\begin{minipage}[t]{0.47\textwidth}
+\centering
+\includegraphics[width=\textwidth]{cot6_50_late.png}
+\caption{Node 6 channels 112-120, after shift}
+\label{cot6_50_late}
+\end{minipage}\hfill
+\end{figure}
+
+\subsubsection{Channel Activity}
+\begin{figure}[ht]
+\begin{minipage}{0.47\textwidth}
+	\centering
+	\includegraphics[width=\textwidth]{images/5_GHz/cot-node3-student_2017-05-22_chan100_image.pdf}
+	\caption{Node 3, ch. 120, 15/05 - 22/05}
+	\label{node3_5ghz_traffic_chan100}
+\end{minipage}\hfill
+\begin{minipage}{0.47\textwidth}
+	\centering
+	\includegraphics[width=\textwidth]{images/5_GHz/cot-node8-student_2017-05-16_chan120_image.pdf}
+	\caption{Node 8, ch. 120, 12/05 - 16/05}
+	\label{node8_5ghz_traffic_chan120}
+\end{minipage}\hfill
+\end{figure}
+\begin{figure}[ht]
+\begin{minipage}{0.47\textwidth}
+\centering
+\includegraphics[width=\textwidth]{images/5_GHz/node1_2017-05-17_chan140_image.pdf}
+\caption{Node 1, ch. 140, 13/05 - 17/05}
+\label{node1_5ghz_traffic_chan140}
+\end{minipage}\hfill
+\end{figure}
+For all nodes except 1 (outdoor), 6 and 8, channels 100 and above seem to contain no activity.
+For the outside node this could merely be some outside interference, although more outside nodes are required to confirm this.
+For the other two nodes this might simply be a nearby 5.0GHz router.
+Even for those nodes channels 100 and up are fairly silent.
+Figures \ref{node3_5ghz_traffic_chan100} to \ref{node1_5ghz_traffic_chan140} show traffic on higher channels for nodes 3, 8 and 1.
+Each of these was somewhat busy on the 2.4 GHz band, and showed some activity on the lower 5 GHz channels.
+In the higher channels they barely exceed -80dBm.
+We do again notice shifts in signal strength with node 8, most likely caused by devices hopping to another frequency.
+
+For the lower channels, i.e.
+channels between 36 and 64, some more activity is observed: mainly channel 36 and occasionally channel 56 are somewhat used.
+Channel 36 is constantly active on nodes 3 and 1 (outside), as seen in figures \ref{node3_ch36_usage} and \ref{node1_ch36_usage}.
+Furthermore channel 56 shows clear daytime peaks except during the weekend in \ref{node12_ch56_usage}, likely indicating Wi-Fi usage.
+Channel 40 is in all cases silent, one example being shown in figure \ref{node10_ch40_usage}.
+Here we do note a peculiarity of the spectral scan at channel 40.
+Each node, except for the outside node, consistently shows the pattern in figure \ref{ch40_peak}.
+It seems that every inside node always measures this slight peak, but it is completely absent for the outside node.
+As all inside nodes are identical hardware-wise, but the outside node is not, we suspect some component of the inside nodes is continuously creating a signal on channel 40.
+\begin{figure}[!ht]
+\begin{minipage}{0.47\textwidth}
+	\centering
+	\includegraphics[width=\textwidth]{images/5_GHz/cot-node3-student_2017-05-22_chan36_image.pdf}
+	\caption{Node 3, ch. 36, 15/05 - 22/05}
+	\label{node3_ch36_usage}
+\end{minipage}\hfill
+\begin{minipage}{0.47\textwidth}
+	\centering
+	\includegraphics[width=\textwidth]{images/5_GHz/node1_2017-05-17_chan36_image.pdf}
+	\caption{Node 1, ch. 36, 13/05 - 17/05}
+	\label{node1_ch36_usage}
+\end{minipage}\hfill
+\begin{minipage}{0.47\textwidth}
+	\centering
+	\includegraphics[width=\textwidth]{images/5_GHz/cot-node12-student_2017-05-21_chan56_image.pdf}
+	\caption{Node 12, ch. 56, 15/05 - 22/05}
+	\label{node12_ch56_usage}
+\end{minipage}\hfill
+\begin{minipage}{0.47\textwidth}
+	\centering
+	\includegraphics[width=\textwidth]{images/5_GHz/cot-node10-student_2017-05-22_chan40_image.pdf}
+	\caption{Node 10, ch. 40, 15/05 - 22/05}
+	\label{node10_ch40_usage}
+\end{minipage}\hfill
+
+	\centering
+	\includegraphics[width=\textwidth]{ch40_peak.png}
+	\caption{Spectral scan: peak for channel 40}
+	\label{ch40_peak}
+
+\end{figure}
+\subsection{Implications of high signal strength}
+A high or unstable measured signal strength does not necessarily make a channel unusable for a City of Things node.
+First of all, we note that all received signal strengths previously reported was due to activity caused by external sources (i.e.
+not the node itself).
+When considering using the node for communication, all those signals are considered as noise.
+As long as this noise is lower than the signal strength used for the node's communication, there is no issue.
+While a constant noise of -20dBm will make communication very difficult, noise of -70dBm, -60dBm or even -50dBm should be no problem for close-range communication, as the node and its peer could probably generate a stronger signal.
+Furthermore it is important to note that in many cases the measured high signal strength was not constant.
+As it was probably caused by a communication device (likely Wi-Fi), the signal source remains silent for a significant fraction of the time.
+Furthermore such sources of potential interference probably implement some sort of listen-before-talk, further facilitating communication with the node in the presence of such sources.
+It is however difficult to predict exactly how well communication will work for a specific situation without actually trying it out.
 \subsection{Validation}
@@
 One interesting observation is that channels 6 and 11 were never busier than channel 1.
+\\
+Furthermore, we found nothing indicating that City of Things network would experience considerable interference in the 5 GHz band, especially for the higher channel numbers.
+When operating on the 2.4 GHz band we observed more activity that could, in some places and on some channels, hinder communication with faraway hosts.
+However there was always at least one channel fairly silent.
 \bibliographystyle{IEEEtran}

### 2017-05-26 advnetwpaper/paper_no_ieeetrans.tex
Minor consistency fixes, hid ugly link boxes
Former-commit-id: a4a3c1868c79a3f1346f58a0aada652d9fac1749 [formerly 2078174ed1f41efecafda47883090f48355117b9 [formerly 2078174ed1f41efecafda47883090f48355117b9 [formerly 0c70af2af63832736dd85b9edcace265a132c918]]]
Former-commit-id: 8e61cb95cda72d4fa4d41cab8638d8afb9cdf15b
Former-commit-id: a19393d6cbd5385e1c4faefdd295c8a3659c65a5
Former-commit-id: cca104387f7f17c3681533f4e09507b5102638e6

@@
 
-\usepackage{hyperref}
+\usepackage[hidelinks]{hyperref}
 
@@
 % Do not put math or special symbols in the title.
-\title{A study on the usage of the wireless signal spectrum in a City of Things network}
+\title{An Analysis of the Usage of the 2.4 GHz and 5 GHz Radio Bands in Antwerp}
 
@@
 Devices emitting Radio Frequency (RF) energy without a license often use one of the unlicensed Industrial, Scientific and Medical (ISM) radio bands.
-The most well-known of these bands is the 100MHz wide band centered around 2.45GHz, often called the 2.4GHz band.
-Many consumer devices such as wireless mice and keyboards, bluetooth headphones and Radio Controlled (RC) cars communicate in the 2.4GHz band.
-Furthermore most IEEE 802.11 (Wi-Fi) routers operate in the 2.4GHz band.
+The most well-known of these bands is the 100MHz wide band centered around 2.45GHz, often called the 2.4 GHz band.
+Many consumer devices such as wireless mice and keyboards, bluetooth headphones and Radio Controlled (RC) cars communicate in the 2.4 GHz band.
+Furthermore most IEEE 802.11 (Wi-Fi) routers operate in the 2.4 GHz band.
 In addition 802.11ac routers operating in the 5GHz range often use the 5GHz ISM band (which actually ranges from 5.725GHz to 5.875GHz).
@@
 All this leads to high interference in these bands.
-In this paper, we investigate how bad this 2.4GHz and 5GHz interference actually is in the city of Antwerp.
+In this paper, we investigate how bad this 2.4 GHz and 5GHz interference actually is in the city of Antwerp.
 \\ \\
@@
 This network consists of hundreds of sensors and wireless gateways positioned around the city of Antwerp.
-On 11 of these gateways, we sampled activity on the 2.4GHz and 5GHz bands every 2 minutes for 12 days, including a Monday-to-Sunday week.
+On 11 of these gateways, we sampled activity on the 2.4 GHz and 5GHz bands every 2 minutes for 12 days, including a Monday-to-Sunday week.
 The main motivation of this experiment is to investigate whether interference around the gateways is significant enough to pose a problem for the network.
@@
 
-At first we measured the spectrum on every channel in both the 2.4GHz and the 5GHz band every 10 minutes.
+At first we measured the spectrum on every channel in both the 2.4 GHz and the 5GHz band every 10 minutes.
 The visualizations of the gathered data using Simon W{\"u}nderlich's FFT\_eval tool\footnote{https://github.com/simonwunderlich/FFT\_eval} showed that a shorter time between scans would be useful.
@@
 The choice of $n$ depends on the sample size.
-For a single-channel 2.4GHz scan, consisting of around 210 samples, we remove the top 5.
+For a single-channel 2.4 GHz scan, consisting of around 210 samples, we remove the top 5.
 We then consider the maximum signal strength among the remaining samples as the measured strength within the channel.
@@
 
-We perform one scan per minute, alternating between 2.4GHz and 5.0GHz.
+We perform one scan per minute, alternating between 2.4 GHz and 5 GHz.
 As each scan takes at least 25 seconds to complete, we chose not to lower this interval any further.
@@
 For one week (Monday 15 May - Sunday 21 May) we plot each day's smoothing, for easy comparison between the days.
-For the 2.4GHz band we perform this separately for the non-overlapping channels 1, 6 and 11.
+For the 2.4 GHz band we perform this separately for the non-overlapping channels 1, 6 and 11.
 Each color in the plot represents a day of the week, using the legend in figure \ref{legend}.
@@
 
-\subsection{Accuracy of the Measurements} \label{accuracy}
+\subsection{Accuracy of the measurements} \label{accuracy}
 
@@
 \includegraphics[width=\textwidth]{24check.png}
-\caption{2.4GHz spectral scan with active nearby device at 2.437GHz.}
+\caption{2.4 GHz spectral scan with active nearby device at 2.437GHz.}
 \label{fig_24}
@@
 \includegraphics[width=\textwidth]{24check_nodata.png}
-\caption{2.4GHz baseline spectral scan.}
+\caption{2.4 GHz baseline spectral scan.}
 \label{fig_24nodata}
@@
 % via \DeclareGraphicsExtensions.
-\caption{5.0GHz spectral scan with active nearby device at 5.660GHz}
+\caption{5 GHz spectral scan with active nearby device at 5.660GHz}
 \label{fig_50}
@@
 % via \DeclareGraphicsExtensions.
-\caption{5.0GHz baseline spectral scan.}
+\caption{5 GHz baseline spectral scan.}
 \label{fig_50nodata}
@@
 To test this, we perform a data transfer with a laptop positioned roughly 50cm from the node.
-We perform this on available 2.4GHz and 5.0GHz networks.
-The 2.4GHz network was using channel 6 (around 2.437GHz) and lead to the spectral scan in figure \ref{fig_24}.
+We perform this on available 2.4 GHz and 5 GHz networks.
+The 2.4 GHz network was using channel 6 (around 2.437GHz) and lead to the spectral scan in figure \ref{fig_24}.
 This scan shows a clear peak centered around channel 6 of around 20MHz wide.
@@
 While the hazy peak around channel 1 is still there, the peak around channel 6 is not visible, indicating it was caused by our data transfer.
-Comparing the two, we see that our data transfer caused an increase in received signal strength of around 40dBm.\\ Our 5.0GHz test was using an access point centered around channel 132 (5.66GHz) and lead to figure \ref{fig_50}.
+Comparing the two, we see that our data transfer caused an increase in received signal strength of around 40dBm.\\ Our 5 GHz test was using an access point centered around channel 132 (5.66GHz) and lead to figure \ref{fig_50}.
 Compared to the flat \ref{fig_50nodata} captured with the data transfer disabled, the peak is again very clear.
@@
 
-\subsubsection{Channel Activity}
+\subsubsection{Channel activity}
 
@@
 This is partly to be expected.
-Many common communication devices only use the 2.4GHz, and many routers in use do not yet support the 5.0GHz band.
-Furthermore, any communication that does occur in this band can be more spread out as the band is a lot wider than the 2.4GHz band.
+Many common communication devices only use the 2.4 GHz, and many routers in use do not yet support the 5 GHz band.
+Furthermore, any communication that does occur in this band can be more spread out as the band is a lot wider than the 2.4 GHz band.
 For the most part, the channels that did show traffic did not show any significant patterns or recurring phenomena.
@@
 For example, node 3 measured a lower signal strength for some time around noon on both May 17\textsuperscript{th} and May 22\textsuperscript{nd} on channel 56, in figure \ref{cot3_5ghz_pattern_chan36}.
-The reason for this however is not as clear as with the daily peaks experienced in the 2.4GHz It might be that the decrease occurs every couple of days, for example due to a router switching channels temporarily.
+The reason for this however is not as clear as with the daily peaks experienced in the 2.4 GHz It might be that the decrease occurs every couple of days, for example due to a router switching channels temporarily.
 It could also simply be an anomaly.
 
-For this node, channel 56 was the only somewhat active channel. All other channels saw barely any traffic in the 5.0 GHz band. \\\\
+For this node, channel 56 was the only somewhat active channel. All other channels saw barely any traffic in the 5 GHz band. \\\\
 It is worth noting that, apart from channel 36, there is some difference in the stability of the received signal between the outdoor node and the indoor nodes.
@@
 
-\subsubsection{Channel Activity}
+\subsubsection{Channel activity}
 \begin{figure}[ht]
@@
 For the outside node this could merely be some outside interference, although more outside nodes are required to confirm this.
-For the other two nodes this might simply be a nearby 5.0GHz router.
+For the other two nodes this might simply be a nearby 5 GHz router.
 Even for those nodes channels 100 and up are fairly silent.
@@
 \section{Conclusions}
-In this report we presented our findings on activity in the RF spectrum in the 2.4GHz and 5.0GHz bands in the city of Antwerp.
-One clear theme in our results is that the 5.0GHz is overall quieter than the 2.4GHz band.
+In this report we presented our findings on activity in the RF spectrum in the 2.4 GHz and 5 GHz bands in the city of Antwerp.
+One clear theme in our results is that the 5 GHz is overall quieter than the 2.4 GHz band.
 This is not surprising.
-The 5.0GHz band is a lot wider than the 2.4GHz band, and its signals do not carry as far.
-Furthermore, assuming that a lot of interference originated from Wi-Fi routers, one factor could be that many active routers and other devices do not yet support 5.0GHz.
-In addition, technologies such as Bluetooth, which could also be creating signals, only affects 2.4GHz.\\ \\
+The 5 GHz band is a lot wider than the 2.4 GHz band, and its signals do not carry as far.
+Furthermore, assuming that a lot of interference originated from Wi-Fi routers, one factor could be that many active routers and other devices do not yet support 5 GHz.
+In addition, technologies such as Bluetooth, which could also be creating signals, only affects 2.4 GHz.\\ \\
 In some environments we notice a daily pattern where the signal strength is about 20dBm higher during the day compared to the night.

### 2017-05-26 advnetwpaper/paper_no_ieeetrans.tex
A few minor changes to spelling and wording.

@@
 
-How bad is it, in general? 
-Are there patterns, e.g. daily recurrence? 
-Advanced: how can you ensure data quality, i.e., how to validate the results?
-
 \maketitle
@@
 We use this 4 hour buffer for all smoothings.
-Furthermore we decided to use the larger window size.
-The smoother graphs make it easier on the eye to compare multiple smoothings on one graph.
-Additionally effects such as the drop during lunchtime were often not seen with even tighter window sizes.\\
+Furthermore we decided to use the larger window size, the smoother graphs make it easier on the eyes to compare multiple smoothings on one graph.
+Additionally effects such as the drop during lunchtime were often not seen with tighter window sizes.\\
 \newpage %Required to fix wrapfigure formatting
@@
 \includegraphics[width=\textwidth]{images/5_GHz/cot-node3-student_2017-05-22_chan36_image.pdf}
-\caption{Node 3, ch. 56, 15/05 - 22/05}
+\caption{Node 3, ch. 36, 15/05 - 22/05}
 \label{cot3_5ghz_pattern_chan36}
@@
 This is partly to be expected.
-Many common communication devices only use the 2.4 GHz, and many routers in use do not yet support the 5 GHz band.
+Many common communication devices only use the 2.4 GHz frequency band and many routers in use do not yet support the 5 GHz band.
 Furthermore, any communication that does occur in this band can be more spread out as the band is a lot wider than the 2.4 GHz band.
@@
 It could also simply be an anomaly.
-
-For this node, channel 56 was the only somewhat active channel. All other channels saw barely any traffic in the 5 GHz band. \\\\
+For this node, channel 36 was the only somewhat active channel.
+All other channels saw barely any traffic in the 5 GHz band.
+\\\\
 It is worth noting that, apart from channel 36, there is some difference in the stability of the received signal between the outdoor node and the indoor nodes.
@@
 We do note that the received high signal is still only -70dBm, meaning there is still room for communication.
-Furthermore the spectral scan samples show that the those channels are still silent for a large fraction of the time.
+Furthermore the spectral scan samples show that those channels (i.e.
+112 and 120) are still silent for a large fraction of the time.
 
@@
 
-\end{figure}
+\end{figure}\\
 \subsection{Implications of high signal strength}
@@
 Furthermore, assuming that a lot of interference originated from Wi-Fi routers, one factor could be that many active routers and other devices do not yet support 5 GHz.
-In addition, technologies such as Bluetooth, which could also be creating signals, only affects 2.4 GHz.\\ \\
+In addition, technologies such as Bluetooth, which could also be creating signals, only affect 2.4 GHz.\\ \\
 In some environments we notice a daily pattern where the signal strength is about 20dBm higher during the day compared to the night.
@@
 When operating on the 2.4 GHz band we observed more activity that could, in some places and on some channels, hinder communication with faraway hosts.
-However there was always at least one channel fairly silent.
+However there was always at least one channel fairly silent and thus usable.
 \bibliographystyle{IEEEtran}

### 2017-05-26 advnetwpaper/paper_no_ieeetrans.tex
Edited typo.
Former-commit-id: 20a1d32934a72a40191a8f3be762f9db72241544 [formerly 58f91511c9b8f4307339aee0a51c9d924dfc5f00 [formerly 58f91511c9b8f4307339aee0a51c9d924dfc5f00 [formerly 4cabe397ca4ab466574fc41579a84e8b2741720a]]]
Former-commit-id: ee92fb8ba2e3765fed4448a5218341dcedc51181
Former-commit-id: 5001c58f3ae6ec9f8b2ed49e42509bba371c4b62
Former-commit-id: e5d13ad5a8cc732bf9b97f698d3fd1788a8080bc

@@
 This includes noise sources such as thermal noise and cosmic noise.
-The chipset can only receive a singal whose power exceeds this noise floor.
+The chipset can only receive a signal whose power exceeds this noise floor.
 The RSSI value is a (unitless) integer indicating, as the name implies, the received power in the channel.
@@
 Figure \ref{savitzky_0} shows the resulting smoothings for 2 different window sizes.
-The smaller window size reveals some details such as the drop in activity between noon and 1PM (lunch time), while the larger window size is detailed enough to show that daytime is more active than nighttime.
+The smaller window size reveals some details such as the drop in activity between noon and 1PM (lunchtime), while the larger window size is detailed enough to show that daytime is more active than nighttime.
 This filter performs well mainly for the center of the data set.

### 2017-05-27 advnetwpaper/paper_no_ieeetrans.tex
Some final changes to the report

@@
 Devices emitting Radio Frequency (RF) energy without a license often use one of the unlicensed Industrial, Scientific and Medical (ISM) radio bands.
-The most well-known of these bands is the 100MHz wide band centered around 2.45GHz, often called the 2.4 GHz band.
+The most well-known of these bands is the 100MHz wide band centered around 2.45 GHz, often called the 2.4 GHz band.
 Many consumer devices such as wireless mice and keyboards, bluetooth headphones and Radio Controlled (RC) cars communicate in the 2.4 GHz band.
 Furthermore most IEEE 802.11 (Wi-Fi) routers operate in the 2.4 GHz band.
-In addition 802.11ac routers operating in the 5GHz range often use the 5GHz ISM band (which actually ranges from 5.725GHz to 5.875GHz).
-Some other bands in the 5GHz are also allowed for 802.11ac, although more restrictions, such as indoors-use only, apply there~\cite{50ism}.
+In addition 802.11ac routers operating in the 5 GHz range often use the 5 GHz ISM band (which actually ranges from 5.725 GHz to 5.875 GHz).
+Some other bands in the 5 GHz are also allowed for 802.11ac, although more restrictions, such as indoors-use only, apply there~\cite{50ism}.
 These ISM bands were originally intended and still used for devices emitting RF energy for purposes other than communication.
@@
 All this leads to high interference in these bands.
-In this paper, we investigate how bad this 2.4 GHz and 5GHz interference actually is in the city of Antwerp.
+In this paper, we investigate how bad this 2.4 GHz and 5 GHz interference actually is in the city of Antwerp.
 \\ \\
@@
 This network consists of hundreds of sensors and wireless gateways positioned around the city of Antwerp.
-On 11 of these gateways, we sampled activity on the 2.4 GHz and 5GHz bands every 2 minutes for 12 days, including a Monday-to-Sunday week.
+On 11 of these gateways, we sampled activity on the 2.4 GHz and 5 GHz bands every 2 minutes for 12 days, including a Monday-to-Sunday week.
 The main motivation of this experiment is to investigate whether interference around the gateways is significant enough to pose a problem for the network.
@@
 The only missing option relevant to our case is the \textit{chanscan} mode of spectral scan.
-When this mode is triggered the driver gathers a configurable number of samples for every channel it supports.
+When this mode is triggered, the driver gathers a configurable number of samples for every channel it supports.
 We emulate this behaviour using the \textit{background} mode, which continuously scans the chipset's currently configured channel as long as it is not sending or receiving.
@@
 The regular decibel (dB) unit is a dimensionless unit.
-In contrast dBm expresses absolute power in reference to watt.
-0 dBm is defined as 1 milliwatt (mW).
+The dBm unit instead expresses absolute power in reference to watt.
+0dBm is defined as 1 milliwatt (mW).
 For each increase of 3dBm, the power in mW doubles, and for each increase of 10dBm, the power is multiplied by 10.
@@
 Atheros chips follow a rather simple formula: subtracting the noise floor from the current signal strength in dBm results in the RSSI.
-These two values are enough to estimate the signal strength in a channel: a noise floor of -96dBm and an RSSI of 40 indicate a received signal of -56dBm.~\cite{rssi-calc}.
+As a result, the noise floor and RSSI are enough to estimate the signal strength in a channel: a noise floor of -96dBm and an RSSI of 40 indicate a received signal of -56dBm~\cite{rssi-calc}.
 The RSSI reported by Atheros is often limited to 60.
@@
 This remains an educated guess by the community.
+\\ \\
+For the remainder of this document, we will only take the noise floor and RSSI values into account. We saw no need for a granularity of the measurements lower than 20MHz.
 
@@
 
-At first we measured the spectrum on every channel in both the 2.4 GHz and the 5GHz band every 10 minutes.
+At first we measured the spectrum on every channel in both the 2.4 GHz and the 5 GHz band every 10 minutes.
 The visualizations of the gathered data using Simon W{\"u}nderlich's FFT\_eval tool\footnote{https://github.com/simonwunderlich/FFT\_eval} showed that a shorter time between scans would be useful.
-There were often large variations between two subsequent samples, showing a use for a shorter interval.
+There were often large variations between two subsequent samples.
 Additionally we noticed some extreme outliers.
 Occasionally a received signal strength of 129dBm or 8GW was reported.
-As a steady 8GW supply would require over 35 million solar panels~\cite{gigawatt}, so this is clearly an erroneous measurement.
+As a steady 8GW supply would require over 35 million solar panels~\cite{gigawatt}, this is clearly an erroneous measurement.
 Although this was a fairly rare phenomenon, it still occurred up to a few times in each node's almost 10 million daily samples.
@@
 We use this 4 hour buffer for all smoothings.
-Furthermore we decided to use the larger window size, the smoother graphs make it easier on the eyes to compare multiple smoothings on one graph.
-Additionally effects such as the drop during lunchtime were often not seen with tighter window sizes.\\
-\newpage %Required to fix wrapfigure formatting
+Furthermore we decided to use the larger window size, as we noticed plots with multiple days combined were more readable with this size.
+Additionally effects such as the drop during lunchtime were often not seen even with tighter window sizes.\\
+%\newpage %Required to fix wrapfigure formatting
 \begin{wrapfigure}{r}{0.35\textwidth}
@@
 For one week (Monday 15 May - Sunday 21 May) we plot each day's smoothing, for easy comparison between the days.
-For the 2.4 GHz band we perform this separately for the non-overlapping channels 1, 6 and 11.
+We repeat this for multiple channels in both bands.
 Each color in the plot represents a day of the week, using the legend in figure \ref{legend}.
@@
 \label{fig_50}
-\end{figure}
 
-\begin{figure}[ht]
 \centering
@@
 We first confirm that the spectral scan does in fact accurately measure activity per channel.
-To test this, we perform a data transfer with a laptop positioned roughly 50cm from the node.
-We perform this on available 2.4 GHz and 5 GHz networks.
+To test this, we performed a data transfer with a laptop positioned roughly 50cm from the node, using available 2.4 GHz and 5 GHz networks.
 The 2.4 GHz network was using channel 6 (around 2.437GHz) and lead to the spectral scan in figure \ref{fig_24}.
@@
 Comparing the two, we see that our data transfer caused an increase in received signal strength of around 40dBm.\\ Our 5 GHz test was using an access point centered around channel 132 (5.66GHz) and lead to figure \ref{fig_50}.
-Compared to the flat \ref{fig_50nodata} captured with the data transfer disabled, the peak is again very clear.
+Compared to the flat figure \ref{fig_50nodata} captured with the data transfer disabled, the peak is again very clear.
 Received signal strength is increased by up to 50dBm in this case.
@@
 There were no peaks during the weekend.
-Node 8 was one of the nodes with failure, so we had to limit its timeframe.
+Node 8 was one of the failing nodes, so we had to limit its timeframe.
 
@@
 During the weekend (black and yellow) there is no peak whatsoever.
-This indicates the traffic is most likely caused by Wi-Fi usage at the university, which is closed during the weekend.
+This indicates the activity is most likely caused by Wi-Fi usage at the university, which is closed during the weekend.
 For the other channels there a noticeable peak during the daytime for some but not all weekdays.
@@
 Node 3 and 9, shown in those figures, measure a very busy medium around channel 1.
-However channels 6 and 11 are not particulary busy compared to the traffic seen on the other nodes, as we can see from these graphs:\\
+However channels 6 and 11 are not particulary busy compared to other nodes.
+The two nodes show channel 1 activity of over -25dBm.
+This is likely caused by something else than a Wi-Fi router, although we are not sure what.
+This would likely make communication with the node over channel 1 very difficult.\\
 
 All nodes covered so far were indoors nodes.
-We had access to one outdoors node, node 1, on which we also performed the experiment.
+We had access to one outdoors node, node 1, on which we also performed the experiment, as seen in figures \ref{node1-1} to \ref{node1-11}.
 On the indoors nodes we noticed that channel 6 and 11 were never any busier than channel 1.
@@
 Note that we again do not have a full week's data for this node.
-Note the remarkable difference between weekdays and the weekend here.
+There is a remarkable difference between weekdays and the weekend here.
 For channel 11, the weekend is more quiet, however for channel 6 the weekend is the busiest period.
@@
 Furthermore, any communication that does occur in this band can be more spread out as the band is a lot wider than the 2.4 GHz band.
-For the most part, the channels that did show traffic did not show any significant patterns or recurring phenomena.
+For the most part, the channels that did show activity did not show any significant patterns or recurring phenomena.
 In most cases the received signal strength was very stable on all channels.
@@
 For this node, channel 36 was the only somewhat active channel.
-All other channels saw barely any traffic in the 5 GHz band.
+All other channels saw barely any activity in the 5 GHz band.
 \\\\
@@
 There were some other interesting phenomena we observed in the 5 GHz channels.
-For example, node 6 shows a sudden drop in activity on both channels 56 and 64, coinciding with an increase in traffic on channels 112 and 120, as seen in figures \ref{node6_5ghz_traffic_chan56} up to \ref{node6_5ghz_traffic_chan120}.
+For example, node 6 shows a sudden drop in activity on both channels 56 and 64, coinciding with an increase in activity on channels 112 and 120, as seen in figures \ref{node6_5ghz_traffic_chan56} to \ref{node6_5ghz_traffic_chan120}.
 Looking at a random spectral scan from before and after the shift, we indeed see the same behaviour.
@@
  While there is some very slight activity in channel 60, so few samples contained any trace of it that it was filtered out.
-Later on, figure \ref{cot6_50_late} shows that this traffic signature has moved to channels 112 and 120, now with the channel in between showing no traffic at all.
+Later on, figure \ref{cot6_50_late} shows that this activity signature has moved to channels 112 and 120, now with the channel in between showing no activity at all.
 We are not sure what device exactly causes this, but it is clear that it was at some point switched to another channel.
-This shows that the activity in a channel can change at a moment's notice, and indicates that systematic or even constant monitoring of the spectrum is not without merit.
+This shows that the activity in a channel can change at a moment's notice, and indicates that scheduled or even constant monitoring of the spectrum is not without merit.
 We do note that the received high signal is still only -70dBm, meaning there is still room for communication.
-Furthermore the spectral scan samples show that those channels (i.e.
-112 and 120) are still silent for a large fraction of the time.
+Furthermore the spectral scan samples show that the affected channels are still silent for a large fraction of the time.
 
@@
 Even for those nodes channels 100 and up are fairly silent.
-Figures \ref{node3_5ghz_traffic_chan100} to \ref{node1_5ghz_traffic_chan140} show traffic on higher channels for nodes 3, 8 and 1.
+Figures \ref{node3_5ghz_traffic_chan100} to \ref{node1_5ghz_traffic_chan140} show activity on higher channels for nodes 3, 8 and 1.
 Each of these was somewhat busy on the 2.4 GHz band, and showed some activity on the lower 5 GHz channels.
@@
 Channel 36 is constantly active on nodes 3 and 1 (outside), as seen in figures \ref{node3_ch36_usage} and \ref{node1_ch36_usage}.
-Furthermore channel 56 shows clear daytime peaks except during the weekend in \ref{node12_ch56_usage}, likely indicating Wi-Fi usage.
+Furthermore channel 56 shows clear daytime peaks except during the weekend in figure \ref{node12_ch56_usage}, likely indicating Wi-Fi usage.
 Channel 40 is in all cases silent, one example being shown in figure \ref{node10_ch40_usage}.
@@
 A high or unstable measured signal strength does not necessarily make a channel unusable for a City of Things node.
-First of all, we note that all received signal strengths previously reported was due to activity caused by external sources (i.e.
-not the node itself).
+First of all, we note that all received signal strengths previously reported were due to activity caused by external sources (as opposed to by the node itself).
 When considering using the node for communication, all those signals are considered as noise.
